\section{Implementation in the BX3 Framework}\label{sec:implementation}

The preceding sections specify the Bailout Protocol as a six-state deterministic machine
$\mathcal{B} = (Q, q_0, \Sigma, \rightarrow, Q_F)$ and as a formal escalation algorithm.
This section describes how that abstract specification is realized concretely within the
\bxthree{} three-layer architecture. For each layer we identify the specific architectural
mechanisms that participate in protocol execution, the precise enforcement constraints
the layer must satisfy, and the structural dependency that makes the layer indispensable
to the protocol's correctness. We then establish the precise relationship between the
Bailout Protocol and the upstream accountability guarantee that defines \bxthree's
central commitment, and we show that the state machine is embedded in the \fact{} layer
not as an auxiliary component but as an extension of the \fact{} layer's own enforcement
primitives.

% --------------------------------------------------------------
\subsection{Purpose Layer: Human Accountability Anchor}\label{sec:purpose-layer}
% --------------------------------------------------------------

The \purpose{} layer carries intent, judgment, and final authority. Within the Bailout
Protocol its role is architecturally irreplaceable: it is the \emph{exclusive terminus}
for all exception propagation and the only layer capable of issuing a binding directive
that closes a protocol run. The operationalization of this role proceeds through four
structured interactions with the protocol.

\medskip\noindent\textbf{Provisioning.} At node initialization the \purpose{} layer
registers the node's human accountability anchor $h(n)$ with the \fact{} layer's Pathway
Registry and records the anchor assignment as a \purpose-layer-authorized entry in the
authorization ledger. This entry constitutes the human principal's explicit, attributable
acceptance of accountability for node $n$'s provisioned operating scope. Critically, for
all nodes $n_i$ in any spawn chain $C$ originating from $n$, $h(n_i) = h(n)$ by structural
constraint: the anchor is non-negotiable and non-migratory across the delegation tree.
The \purpose{} layer enforces this constraint by issuing the anchor assignment as a
single atomic provisioning action; once recorded, it cannot be modified by any machine
actor without explicit human re-authorization.

\medskip\noindent\textbf{Exception receipt.} When an exception state reaches the
\purpose{} layer via the Escalate algorithm, the layer receives the full diagnostic
context accumulated along the propagation chain: the originating node, the trigger
condition $\text{TRIGGER}(n, x)$, the operating range $R(n)$, the time elapsed since
first detection, and the resolution history of any prior occurrences of this exception
class. This context is sufficient for the human principal to evaluate the exception
without additional inference or system reconstruction.

\medskip\noindent\textbf{Directive issuance.} The \purpose{} layer evaluates the
exception and issues one of three binding directives:
\begin{itemize}\itemsep=0pt
\item \texttt{ACK} -- acknowledge and resume normal operation;
\item \texttt{RESOLVE} -- execute a specified binding resolution; or
\item \texttt{REJECT} -- disallow the proposed action and enter a quiescent error state.
\end{itemize}
No machine actor may issue, simulate, or substitute any of these directives. Upon
receipt of the directive the state machine transitions to \texttt{HUMAN\_ACKNOWLEDGED}
and then to \texttt{RESOLVED}; the directive is committed to the authorization ledger
as a \purpose-layer-authorized entry bearing the human principal's attributed signature.
The ledger entry is the only acceptable evidence that a binding resolution occurred;
any attempt to close a protocol run without such an entry fails the \fact{} layer's
pre-commit hook.

\medskip\noindent\textbf{Consequential authority.} The \purpose{} layer is the only
layer in the \bxthree{} architecture that carries the authority to commit consequential
decisions to the authorization ledger. This exclusivity is what makes it the sole
permissible terminus for the accountability chain. The functional separation between
\purpose{}, \bounds{}, and \fact{} layers establishes formally that no other layer
carries both intent and final accountability: the \bounds{} layer detects exceptions
but carries no judgment authority; the \fact{} layer enforces transition relations and
records events but carries no intent authority. When an exception exceeds the
\bounds{} layer's provisioned scope, no machine actor exists that has the authority
to adjudicate it. The decision must return to the \purpose{} layer. The Bailout Protocol
is the mechanism by which this structural necessity is enforced at runtime---not as a
preference but as a compulsion derived from the functional separation itself.

% --------------------------------------------------------------
\subsection{Bounds Engine: Triggering Bail-Out}\label{sec:bounds-trigger}
% --------------------------------------------------------------

The \bounds{} layer carries bounded reasoning and constrained execution within the
provisioned operating range $R(n)$. Its participation in the Bailout Protocol is
confined to a single non-discretionary function: detecting conditions that exceed
$R(n)$ and initiating the state machine's compulsory transition to \texttt{BOUNDED}.
No machine actor may interpose, delay, or redirect this function once activated.

\medskip\noindent\textbf{Trigger condition evaluation.} The Bounds Engine evaluates
$\text{TRIGGER}(n, x)$ against every input element $x$ in the node's observable input
space. This evaluation is a read-only query against the operating range definition
$R(n)$ stored at the node's \bounds{} layer worksheet. When all three sub-conditions
of TC hold simultaneously -- $x \notin R(n)$, the condition is not previously resolved,
and $\text{confidence}(n, x) < \tau$ -- the trigger fires and the state machine
transitions from \texttt{IDLE} to \texttt{BOUNDED} via T1. The evaluation is driven
directly by the Bailout Monitor's pre-commit hook, which intercepts the query and
commits the result to the state machine before any agent-level code can act on $x$.
There is no intervening decision step at which a machine actor could suppress or
reclassify the exception.

The \bounds{} layer exercises no discretion over the transition:
\begin{enumerate}\itemsep=0pt
\item The evaluation of $\text{TRIGGER}(n, x)$ is compulsory: once the three
sub-conditions hold, no machine actor may prevent T1 from firing.
\item The \bounds{} layer may not delay, reclassify, or absorb an exception
after $\text{TRIGGER}(n, x)$ evaluates to true. The state machine's deterministic
transition relation is the only permissible response.
\item The confidence threshold $\tau$ is provisioned at node initialization and
is not adjustable at runtime by any machine actor. This prevents a machine actor
from inflating its own confidence to suppress a warranted escalation.
\end{enumerate}

\medskip\noindent\textbf{Secondary trigger.} The explicit \texttt{bailout\_signal}
handles the case where the Bounds Engine detects a logical contradiction or
operational impossibility it recognizes as unresolvable even though both
contributing inputs fall individually within $R(n)$. The \texttt{bailout\_signal}
bypasses TC entirely and immediately initiates T3, ensuring that the trigger is
not limited to conditions expressible as simple boundary violations.

\medskip\noindent\textbf{Time-bounded resolution window.} Upon entering
\texttt{BOUNDED}, the Bounds Engine is granted $T_{\text{bounds}}$ -- a provisioned
constant set at node initialization -- to attempt local resolution. During this
window all resolution paths within current authority remain open. If the Bounds
Engine produces a resolution $r$ that the \fact{} layer approves before
$T_{\text{bounds}}$ expires, the state machine returns to \texttt{IDLE} via T2 and
normal execution resumes. If all resolution paths are exhausted or $T_{\text{bounds}}$
expires, the state machine advances to \texttt{BAIL\_PENDING} via T3 and immediately,
atomically, to \texttt{ESCALATING} via T4 -- with no dwell time and no machine-actor
intervention permitted in the \texttt{BAIL\_PENDING} $\rightarrow$ \texttt{ESCALATING}
transition. This atomic transition is the architectural guarantee that the bailout
trigger is compulsory. The Bounds Engine cannot suppress escalation at the moment
of maximum temptation -- when $T_{\text{bounds}}$ expires and the exception must
propagate. The \fact{} layer's pre-commit hook enforces the transition relation
directly, bypassing any agent-level code that might otherwise interpose.

% --------------------------------------------------------------
\subsection{Fact Layer: Enforcement Substrate and Forensic Ledger}\label{sec:fact-layer}
% --------------------------------------------------------------

The \fact{} layer carries deterministic enforcement and forensic auditability. Within
the Bailout Protocol it serves two distinct but cooperating functions: enforcing the
state machine's transition relation as a pre-commit invariant, and maintaining the
append-only Forensic Ledger as an immutable record of every protocol event.

The \fact{} layer is the only mechanism in the \bxthree{} architecture capable of
authorizing a state transition or committing an entry to the authorization ledger.
This exclusivity is what makes its enforcement structurally reliable. No machine
actor -- including the Bounds Engine, any spawned subordinate agent, or any external
system operating outside the \fact{} layer's jurisdiction -- can authorize a
transition that violates the state machine's transition relation or commit a ledger
entry that misrepresents the protocol's actual execution.

The \fact{} layer enforces the Bailout Protocol through three cooperating mechanisms:

\begin{itemize}\itemsep=0pt
\item[(\textit{i})] \textbf{Pre-commit hook.} Before any state transition is committed,
the \fact{} layer's Bailout Monitor evaluates whether the proposed transition satisfies
the current state's enabled transition set. If no transition is defined for the
$(q, \sigma)$ pair, the transition is rejected and a protocol violation is recorded in
the Forensic Ledger. This hook runs synchronously and atomically; no intermediate or
partially-executing state is observable externally.

\item[(\textit{ii})] \textbf{Authorization ledger gate.} The \fact{} layer is the sole
write authority for the authorization ledger. Any resolution that purports to close a
protocol run without a corresponding \purpose-layer-authorized directive fails the
gate and is rejected. This prevents the Bounds Engine from issuing a retroactive
resolution after timeout to suppress a warranted escalation -- a specific failure
mode the gate is designed to interdict.

\item[(\textit{iii})] \textbf{Pathway Registry.} The \fact{} layer maintains the
Pathway Health Registry, tracking all communication pathways to the human anchor
$h(n)$. When a pathway is degraded or unavailable, the registry enables the Escalate
algorithm to identify and route through alternative functional pathways. Historical
health data is append-only; current-pathway state is a mutable field updated by the
Bailout Monitor.
\end{itemize}

\medskip\noindent\textbf{Forensic Ledger recording.} The Forensic Ledger is the
immutable evidentiary record produced by the \fact{} layer's enforcement of the
Bailout Protocol. Every protocol event -- trigger evaluations, state transitions,
pathway selections, acknowledgments, directives, and timeout conditions -- is
committed to the ledger as an append-only entry. Each entry is timestamped, attributed
to the originating node, and linked to its predecessor by a SHA-256 hash chain that
makes retrospective tampering computationally detectable.

The Forensic Ledger records each protocol event across three discrete accountability
planes of the 9-Plane DAP architecture:

\begin{enumerate}\itemsep=0pt
\item \textbf{Purpose Plane (P1):} Records the authorization context at the time of
bailout -- the originating node, its human anchor $h(n)$, the active mandate, and
the constraint boundary or trigger condition that was exceeded. This plane establishes
the provenance of the exception in the delegation chain and identifies who is
accountable.

\item \textbf{Bounds Plane (P2):} Records the diagnostic context accumulated during
propagation -- the trigger condition, the proposed resolution attempt, the Bounds
Engine's confidence level at time of firing, and the complete propagation chain
showing every node the exception traversed before reaching the human anchor. This
plane captures what the machine reasoning concluded and why.

\item \textbf{Fact Plane (P3):} Records the physical or digital system state at the
time of bailout -- actuator positions, sensor readings, system telemetry, and any
physical consequences of actions taken or blocked prior to bailout resolution. This
plane provides the evidentiary baseline against which post-resolution consequences
can be measured.
\end{enumerate}

The Forensic Ledger serves three distinct accountability functions:

\begin{enumerate}\itemsep=0pt
\item \textbf{Attribution.} Every escalation is attributable to a specific node,
trigger condition, and human anchor. The chain of diagnostic context accumulated
during propagation is preserved in full, enabling post-hoc reconstruction of the
escalation's origin and propagation path.

\item \textbf{Non-repudiation.} The human principal's directive -- \texttt{ACK},
\texttt{RESOLVE}, or \texttt{REJECT} -- is immutably recorded as a
\purpose-layer-authorized ledger entry. The principal cannot credibly deny having
issued the directive, and the system cannot fabricate a directive that was not
issued.

\item \textbf{Verifiability.} The SHA-256 hash chain enables any authorized auditor
to verify that the ledger has not been altered since the protocol run. Each entry's
hash covers the entry's content and the hash of the preceding entry, making
insertion, deletion, or modification of historical entries computationally
impractical.
\end{enumerate}

The Forensic Ledger is not a conventional audit log. It is a structurally
append-only data structure whose immutability is enforced by the \fact{} layer's
exclusive write authority and by the hash-chain integrity mechanism. A machine
actor cannot delete, modify, or suppress a ledger entry without detection.

% --------------------------------------------------------------
\subsection{Runtime Expression of the Upstream Accountability Guarantee}\label{sec:runtime-expression}
% --------------------------------------------------------------

The upstream accountability guarantee of the \bxthree{} Framework is a single,
architecturally precise commitment: \bxthree{} systems fail upward into human
consciousness, never downward into autonomous chaos. The Bailout Protocol is the
runtime mechanism that makes this guarantee structurally operative -- not for the
exception conditions anticipated and resolved by the four preceding pillars, but for
the complementary set of all unanticipated exceptions that exceed machine-actor
provisioned scope.

The relationship between the protocol and the guarantee is one of mutual structural
definition. The upstream accountability guarantee requires the Bailout Protocol to be
operative: without it, the guarantee would be unenforceable for the exception
conditions that the four preceding pillars do not cover. The guarantee is a claim
about the behavior of the system; the protocol is the mechanism that makes the claim
true. Conversely, the Bailout Protocol requires the structural properties that only
the \bxthree{} three-layer architecture provides: the \purpose{} layer must be
non-delegable as an exception terminus, the $h(n)$ anchor must be immutable at runtime,
and the \fact{} layer must have exclusive write authority over the Forensic Ledger.
These are not optional features that improve the protocol's performance -- they are
structural requirements without which the protocol's guarantees cannot be realized.

The structural completeness of this three-layer implementation is what makes the
upstream accountability guarantee unconditional. The guarantee does not say that human
oversight is \emph{available} if machine actors choose to escalate; it says that
human oversight is \emph{compulsory} for every unresolvable exception, regardless of
machine-actor preferences, operational urgency, or the availability of alternative
resolution paths. The Bailout Protocol, implemented across the three \bxthree{} layers,
is the mechanism that makes this compulsion structurally operative at runtime.

The five pillars of \bxthree{} collectively establish the conditions under which
the upstream accountability guarantee applies. Loop Isolation, Recursive Spawning,
Spatial Firewall, and Sandbox Gate address exception conditions anticipated at design
time and resolvable within the machine-only execution graph. The Bailout Protocol
addresses the complementary remainder -- conditions not anticipated, that exceed the
operating range of any machine actor, and that cannot be resolved within the machine-only
execution graph. The completeness of this coverage is what makes the upstream
accountability guarantee unconditional: it is not that \emph{most} exceptions reach
$h(n)$, but that \emph{every} exception reaches $h(n)$, because the trigger condition
fires on any condition outside $R(n)$ and the escalation path has no machine-actor
terminus by architectural constraint.

% --------------------------------------------------------------
\subsection{State Machine as Fact Layer Extension}\label{sec:state-machine-fact-layer}
% --------------------------------------------------------------

The Bailout Protocol state machine $\mathcal{B}$ is embedded in the \fact{} layer of
every \bxthree{} node as a Fact Layer extension -- a deterministic enforcement automaton
whose transition relation is enforced by the \fact{} layer's pre-commit hook, whose
state is stored in the \fact{} layer's worksheet, and whose state transitions generate
entries in the Forensic Ledger.

This embedding is not incidental. The state machine's enforcement requirements are
precisely matched to the \fact{} layer's architectural properties:

\begin{itemize}\itemsep=0pt
\item[(\textit{i})] \textbf{Deterministic transition enforcement.} For every
$(q, \sigma) \in Q \times \Sigma$, at most one transition is defined. This determinism
is delivered by the Bailout Monitor's priority function $\pi$, which totally orders
simultaneous trigger conditions before committing any state change. The \fact{} layer
provides the scheduling guarantee and atomic pre-commit hook required to make this
ordering consistent and crash-safe across all nodes, including under conditions of
network partition or partial node failure.

\item[(\textit{ii})] \textbf{Atomic state transitions.} No intermediate or
partially-executing state is observable externally. The \fact{} layer's pre-commit hook
evaluates and commits the transition in a single synchronized operation, preventing
any interleaving machine actor from observing or modifying the state mid-transition.
This atomicity is the enforcement mechanism for INV-2, which states that any node in
\texttt{BAIL\_PENDING} has an active bailout and the only transition defined for this
state with an enabled event is to \texttt{ESCALATING}.

\item[(\textit{iii})] \textbf{State observability.} The state machine's current state
must be observable by the parent node and by the Bailout Monitor without granting
write access. The \fact{} layer exposes a state observation interface to the parent
node, enabling parent-level monitoring of child escalation status independently of the
child node's internal representation. This enables parent-initiated timeouts and
cross-node invariant monitoring.

\item[(\textit{iv})] \textbf{Ledger co-commitment.} Every state transition generates a
mandatory Forensic Ledger entry as a co-commitment of the transition itself. No state
change in $\mathcal{B}$ is committed without a corresponding ledger entry recording the
transition event, the triggering condition, and the diagnostic context. This
co-commitment is the mechanism by which the Forensic Ledger achieves attribution:
every state change is attributable because every state change is recorded.

\item[(\textit{v})] \textbf{Immutable parent-pointer enforcement.} The parent pointer
$\alpha(n)$ and the human-root identifier $h(n)$ are stored in \fact-layer-managed
memory and are read-only at the machine-actor level. A machine actor cannot redirect
the parent chain, cannot insert itself as an intermediate resolver, and cannot modify
the $humanRootId$ field that terminates the chain. Any attempted modification requires
\purpose{} layer authorization with a corresponding ledger entry -- which cannot be
completed without human approval if the modification would affect the bailout terminus.
This immutability is what guarantees that the escalation path is fixed at provisioning
and cannot be altered by any machine actor at runtime.
\end{itemize}

The state machine is therefore not merely \emph{located} in the \fact{} layer -- it
is defined by the \fact{} layer's enforcement capabilities. The six-state machine,
the seven-transition relation (T1--T7), the timeout parameters $T_{\text{bounds}}$ and
$T_{\text{escalate}}$, and the Escalate algorithm are all realized as \fact{} layer
primitives. The \fact{} layer does not merely host the state machine; it \emph{is}
the state machine's enforcement substrate. The mutual definition of protocol and
framework is thus complete at the level of implementation: the \fact{} layer provides
the enforcement primitives that make the state machine compulsive, and the state
machine is the application of those primitives to the accountability domain that the
upstream guarantee requires.

The three state invariants are maintained by the \fact{} layer's enforcement
architecture:

\begin{align}
\text{INV}_1 &: \forall n \in \text{nodes},\; \text{state}(n) \in Q \setminus
\{\text{ESCALATING}\} \iff \neg\text{active\_bailout}(n). \tag{INV-1}
\end{align}

INV-1 is enforced by the \fact{} layer's pre-commit hook: a node may enter a
non-escalating state only when the Bailout Monitor confirms that no active propagation
is in flight. The state and the active-bailout flag are updated atomically, preserving
the equivalence. This invariant guarantees that the state machine cannot represent a
node as idle while an active escalation is in flight.

\begin{align}
\text{INV}_2 &: \text{state}(n) = \text{BAIL\_PENDING} \implies
\text{active\_bailout}(n) \land
\text{transition}(\text{BAIL\_PENDING}, e_{\text{timeout}}) = \text{ESCALATING}.
\tag{INV-2}
\end{align}

INV-2 follows from the \fact{} layer's enforcement of the transition relation. Any
node in \texttt{BAIL\_PENDING} has an active bailout, and the only transition defined
for this state with an enabled event is to \texttt{ESCALATING}. The Bailout Monitor
enforces this transition compulsorily upon receipt of $e_{\text{timeout}}$; no machine
actor may interpose. This invariant is the architectural expression of the upstream
accountability guarantee at the state-machine level: a node in \texttt{BAIL\_PENDING}
must escalate, and the escalation must reach \texttt{ESCALATING}.

Together, the state invariants and the \fact{} layer's enforcement architecture ensure
that the Bailout Protocol state machine operates as a deterministic, atomic, auditable
extension of the \fact{} layer -- and that every transition it executes is immutably
recorded in the Forensic Ledger as evidence that the upstream accountability guarantee
has been upheld for the current protocol run.