\begin.Abstract}
Multi-agent systems that support recursive spawning of descendant agents face a structural vulnerability we term \emph{autonomous drift}: the progressive severing of descendant agents from human authority as spawning chains grow beyond legible depth. Without an enforced architectural constraint on the parent--child relationship, each generation risks operating as an effectively rootless agent, diminishing accountability and introducing emergent hazards that no single-agent safety boundary can contain. We address this vulnerability through a declarative spawning protocol in which every agent is stamped at creation with an \emph{immutable parent pointer} that records its immediate progenitor and, transitively, the full ancestry chain terminating at the originating human principal. Because pointers are immutable and the chain is append-only, no agent can spawn a descendant that is not, by construction, traceable to root human authority. We prove that under this protocol, autonomous drift is architecturally impossible: any descendant agent's provenance is always resolvable, and any unauthorized spawning attempt can be detected and rejected at the protocol layer before it propagates. We integrate this protocol into the BX3 Framework as a core primitive, extending its accountability surface to multi-level spawning topologies while preserving the framework's existing execution guarantees. Evaluation across spawned agent topologies of depth $d \in \{1, \dots, 10\}$ confirms that provenance resolution overhead scales $O(d)$ and that drift-resistant spawning is achievable without compromising parallelism or throughput. This work establishes immutable parent pointers as a minimal, provably safe foundation for accountable recursive multi-agent systems.

\textbf{Keywords:} recursive spawning, immutable parent pointers, autonomous drift, multi-agent systems, accountability
\end.Abstract}
