\section{The Bailout Protocol}

\subsection{Overview}

The Bailout Protocol is the terminal recovery mechanism of the Agentic runtime.
It is invoked when all lower-level error-recovery pathways have been exhausted and
the system faces a condition from which autonomous recovery is either impossible or
unsafe.  The protocol defines a state machine that coordinates controlled shutdown,
escalation, and eventual resolution of in-flight work, ensuring that no agent task
terminates silently, that all side-effects are auditable, and that a human operator
is brought into the loop before any irreversible action is taken.

\bigskip

\noindent\textbf{Design goals:}

\begin{itemize}
  \itemsep0pt
  \item \textbf{Determinism.}  Every transition is triggered by an explicit,
    verifiable condition; no silent or probabilistic state changes occur.
  \item \textbf{Inviolability of human oversight.}  A human is always placed in
    the loop before any state carrying irreversible consequences is entered.
  \item \textbf{Bounded autonomy.}  The system may act autonomously only within
    a pre-defined execution envelope; the Bailout Protocol defines the boundary.
  \item \textbf{Composability.}  The protocol is orthogonal to individual agent
    logic and applies uniformly across all agent types, skill taxonomies, and
    runtime environments.
\end{itemize}

\bigskip

\noindent\textbf{Governing principle.}  Any autonomous agent action whose
reversal is computationally intractable or economically infeasible must be
classified as a terminal state and routed through the Bailout Protocol before
execution proceeds.

% ---------------------------------------------------------------- %
% 3.1  State Machine Definition
% ---------------------------------------------------------------- %
\subsection{State Machine Definition}

The protocol is formally specified as a deterministic finite automaton
$\mathcal{B} = (Q, \Sigma, \delta, q_0, F)$, where:

\begin{itemize}
  \item $Q = \{\,\text{IDLE},\ \text{BOUNDED},\ \text{BAIL\_PENDING},\ \text{ESCALATING},\ \text{HUMAN\_ACKNOWLEDGED},\ \text{RESOLVED}\,\}$
  \item $\Sigma$ is the finite alphabet of transition events (defined in \S3.2)
  \item $\delta: Q \times \Sigma \rightarrow Q$ is the total transition function
  \item $q_0 = \text{IDLE}$ is the initial state
  \item $F = \{\text{RESOLVED}\}$ is the set of terminal accept states
\end{itemize}

\bigskip

\begin{center}
\begin{tabular}{|l|l|p{6.5cm}|}
\hline
\textbf{State} &
\textbf{Type} &
\textbf{Semantic meaning} \\ \hline
IDLE & Initial / Idle & The agent is operational; no error conditions are active. \\ \hline
BOUNDED & Working & A non-fatal error condition has been detected; the agent is
constrained to a reduced action envelope. \\ \hline
BAIL\_PENDING & Waiting & The error condition cannot be autonomously resolved;
the system is preparing to halt and awaits an external trigger. \\ \hline
ESCALATING & Escalating & The condition has triggered the escalation algorithm;
a human operator has been notified and is awaited. \\ \hline
HUMAN\_ACKNOWLEDGED & Paused & A human operator has responded; the agent awaits
resolution instructions before proceeding. \\ \hline
RESOLVED & Terminal & The situation has been resolved; the agent is returning to
IDLE or initiating a clean shutdown. \\ \hline
\end{tabular}
\end{center}

\bigskip

\noindent\textbf{State invariant.}  The machine is always in exactly one state in
$Q$.  All transitions preserve this invariant.  Transitions labeled with the
symbol $\bot$ are infeasible and constitute a violation of the protocol if
attempted.

% ---------------------------------------------------------------- %
% 3.2  Transition Events and Conditions
% ---------------------------------------------------------------- %
\subsection{Transition Events and Conditions}

The event alphabet $\Sigma$ is partitioned into four classes:
monitoring events ($\mathcal{M}$), error events ($\mathcal{E}$), human events
($\mathcal{H}$), and timeout events ($\mathcal{T}$).

\[
\Sigma = \mathcal{M} \cup \mathcal{E} \cup \mathcal{H} \cup \mathcal{T}
\]

\bigskip

\begin{center}
\begin{tabular}{|p{3.2cm}|p{2.8cm}|p{6.5cm}|}
\hline
\textbf{Event ($e$)} & \textbf{Class} & \textbf{Condition (formal)} \\ \hline
$e_\text{nominal}$ & $\mathcal{M}$ & Agent heartbeat confirmed within $T_\text{heartbeat}$ \\ \hline
$e_\text{boundary\_violation}$ & $\mathcal{E}$ & $\exists\, a \in A: \text{constraint}(a, t) > \text{budget}(a)$ \\ \hline
$e_\text{recovery\_exhausted}$ & $\mathcal{E}$ & $\text{retry\_count}(t) \geq N_\text{max\_retries}$ \\ \hline
$e_\text{unsafe\_action}$ & $\mathcal{E}$ & $\Delta_\text{risk}(a_t) > \theta_\text{critical}$ \\ \hline
$e_\text{bypass\_triggered}$ & $\mathcal{E}$ & $b_\text{flag} = \text{TRUE}$ (see \S3.6) \\ \hline
$e_\text{escalation\_sent}$ & $\mathcal{M}$ & Notification dispatched to human operator \\ \hline
$e_\text{human\_ack}$ & $\mathcal{H}$ & Operator acknowledgement received \\ \hline
$e_\text{human\_resolve}$ & $\mathcal{H}$ & Operator issues resolution instruction $i$ \\ \hline
$e_\text{timeout\_bounded}$ & $\mathcal{T}$ & $t - t_\text{bound\_enter} \geq T_\text{bounded}$ \\ \hline
$e_\text{timeout\_escalation}$ & $\mathcal{T}$ & $t - t_\text{escalate\_enter} \geq T_\text{escalate}$ \\ \hline
$e_\text{timeout\_human}$ & $\mathcal{T}$ & $t - t_\text{human\_ack\_enter} \geq T_\text{human}$ \\ \hline
$e_\text{resolved}$ & $\mathcal{M}$ & $\phi_\text{resolved}(s_t) = \text{TRUE}$ \\ \hline
\end{tabular}
\end{center}

\bigskip

The complete transition function $\delta$ is defined as follows.

\begin{center}
\begin{tabular}{|p{3.5cm}|p{3.5cm}|p{2.5cm}|p{4cm}|}
\hline
\textbf{From} & \textbf{Event} & \textbf{To} & \textbf{Guard / Action} \\ \hline
IDLE & $e_\text{boundary\_violation}$ & BOUNDED & $\text{log}(e, \text{timestamp})$ \\ \hline
IDLE & $e_\text{bypass\_triggered}$ & ESCALATING & $\text{activate\_bypass}(t)$ \\ \hline
BOUNDED & $e_\text{nominal}$ & IDLE & $\text{clear\_constraints}()$ \\ \hline
BOUNDED & $e_\text{timeout\_bounded}$ & BAIL\_PENDING & $\text{enter\_bail\_pending}()$ \\ \hline
BOUNDED & $e_\text{recovery\_exhausted}$ & BAIL\_PENDING & $\text{enter\_bail\_pending}()$ \\ \hline
BOUNDED & $e_\text{unsafe\_action}$ & BAIL\_PENDING & $\text{enter\_bail\_pending}()$ \\ \hline
BAIL\_PENDING & $e_\text{escalation\_sent}$ & ESCALATING & $\text{dispatch\_escalation}()$ \\ \hline
ESCALATING & $e_\text{timeout\_escalation}$ & ESCALATING & $\text{re-dispatch\_escalation}()$ \\ \hline
ESCALATING & $e_\text{human\_ack}$ & HUMAN\_ACKNOWLEDGED & $\text{log\_ack}(t)$ \\ \hline
HUMAN\_ACKNOWLEDGED & $e_\text{human\_resolve}$ & RESOLVED & $\text{apply\_resolution}(i)$ \\ \hline
HUMAN\_ACKNOWLEDGED & $e_\text{timeout\_human}$ & RESOLVED & $\text{apply\_timeout\_default}()$ \\ \hline
RESOLVED & $e_\text{resolved}$ & IDLE & $\text{reset\_agent}()$ \\ \hline
\end{tabular}
\end{center}

\bigskip

\noindent\textbf{Transition notation.}  All guards are evaluated against the
current agent state snapshot $s_t$.  No transition is pre-empted by a higher-priority
event unless explicitly ordered.  When multiple events qualify simultaneously,
the priority ordering is $\mathcal{E} > \mathcal{T} > \mathcal{H} > \mathcal{M}$.

% ---------------------------------------------------------------- %
% 3.3  Bail-Out Trigger Condition
% ---------------------------------------------------------------- %
\subsection{Bail-Out Trigger Condition}

The system enters the Bailout Protocol when the \textbf{bail-out trigger}
$\Psi$ evaluates to TRUE.  $\Psi$ is a disjunction over three terminal conditions:

\begin{equation}
\Psi \;=\; \underbrace{\phi_\text{retry}(t)}_{\text{(a)}} \;\lor\;
         \underbrace{\phi_\text{risk}(a_t)}_{\text{(b)}} \;\lor\;
         \underbrace{b_\text{flag}}_{\text{(c)}}
\end{equation}

\bigskip

\noindent where:

\begin{enumerate}
  \itemsep0pt
  \item \textbf{Retry exhaustion} $\phi_\text{retry}(t) \;=\;
    \bigl(\text{retry\_count}(t) \geq N_\text{max\_retries}\bigr) \;\land\;
    \bigl(\text{last\_error}(t) \neq \emptyset\bigr)$

  \item \textbf{Critical risk threshold} $\phi_\text{risk}(a_t) \;=\;
    \bigl(\Delta_\text{risk}(a_t) > \theta_\text{critical}\bigr) \;\land\;
    \bigl(\text{safety\_check}(a_t) = \text{FAIL}\bigr)$

  \item \textbf{Bypass flag} $b_\text{flag}$ is set by the bypass mechanism
    described in \S3.6.
\end{enumerate}

\bigskip

\noindent\textbf{Evaluation.}  $\Psi$ is evaluated on every agent cycle
$[\,t, t + \Delta t\,]$.  The first sub-condition to evaluate to TRUE
immediately triggers the transition IDLE $\rightarrow$ BOUNDED or
IDLE $\rightarrow$ ESCALATING (in the case of (c)).  The triggering sub-condition
and the full state snapshot at time $t$ are logged to the agent's audit trail
before the transition is recorded.

% ---------------------------------------------------------------- %
% 3.4  Escalation Algorithm
% ---------------------------------------------------------------- %
\subsection{Escalation Algorithm}

When the machine enters \texttt{BAIL\_PENDING}, the escalation algorithm
$\Alpha$ is invoked.  $\Alpha$ is executed in three sequential stages:

\bigskip

\begin{algorithm}[H]
\caption{Escalation Algorithm $\Alpha$}
\label{alg:escalation}

\begin{enumerate}
  \item \textbf{Snapshot.}  Capture a complete state snapshot:
    \begin{itemize}
      \itemsep0pt
      \item Current agent ID, skill taxonomy, and workspace context
      \item In-flight task queue with status and resource estimates
      \item Last $K$ error events from the agent log
      \item Full constraint state (resource budgets, action envelopes)
      \item Retry count, last error message, and error category
      \item Human-readable summary of the triggering condition
    \end{itemize}

  \item \textbf{Notify.}  Dispatch notification via all configured channels
    in priority order $\mathcal{C} = [\text{SMS}, \text{Discord DM}, \text{Email}]$.
    \begin{itemize}
      \itemsep0pt
      \item SMS: single-message alert with $\leq 160$ characters: agent ID,
        state, and a one-line action directive
      \item Discord DM: rich embed with the full snapshot payload
      \item Email: full snapshot as a formatted message with a direct reply link
        to \texttt{brodiblanco@zo.computer}
    \end{itemize}
    All dispatch attempts are logged with timestamp, channel, and delivery
    status.  The notification includes a cryptographic commitment to the full
    snapshot so the operator can verify integrity.

  \item \textbf{Wait.}  Enter \texttt{ESCALATING} state and await human response.
    The system enters a holding pattern; no new agent tasks are initiated.
    Existing tasks are suspended at safe checkpoints (no mid-transaction writes
    are permitted in \texttt{ESCALATING} state).

  \item \textbf{Acknowledge.}  Upon receipt of operator acknowledgement
    $e_\text{human\_ack}$, record the operator's identity and the acknowledgement
    timestamp.  Transition to \texttt{HUMAN\_ACKNOWLEDGED}.

  \item \textbf{Execute resolution.}  Receive resolution instruction $i$.
    Validate $i$ against the protocol's action whitelist (see \S3.5).
    If valid, apply the resolution and transition to \texttt{RESOLVED}.
    If invalid, re-enter \texttt{ESCALATING} with an error notice.
\end{enumerate}

\bigskip

\begin{note}
\textbf{Idempotency.}  The escalation algorithm is idempotent.  Re-triggering
$\Alpha$ for the same agent incident while the agent is already in
\texttt{ESCALATING} state does not generate duplicate notifications.
A deduplication key (agent ID $\oplus$ incident hash) is checked before dispatch.
\end{note}

% ---------------------------------------------------------------- %
% 3.5  Bypass Mechanism
% ---------------------------------------------------------------- %
\subsection{Bypass Mechanism}

The bypass mechanism exists to accommodate operator overrides and emergency
conditions where waiting for the standard state machine would cause material
harm.  When $b_\text{flag} = \text{TRUE}$, the machine bypasses all intermediate
states and transitions directly from the current state to \texttt{ESCALATING}:

\[
\delta(q, e_\text{bypass\_triggered}) =
\begin{cases}
\text{ESCALATING} & \text{if } b_\text{flag} = \text{TRUE} \\
\delta_\text{normal}(q, e) & \text{otherwise}
\end{cases}
\]

\bigskip

\noindent\textbf{Bypass rationale.}  The bypass is justified by the following
operational theorem:

\bigskip

\begin{samepage}
\begin{center}
\fbox{
\parbox{0.92\linewidth}{
\textbf{Bypass Theorem.}  Let $q$ be any state in $Q \setminus
\{\text{RESOLVED}\}$.  Let $e_\text{bypass\_triggered}$ be raised at time $t$.
Then there exists an operator intent $O$ such that the cost of
executing $\delta(q, e_\text{bypass\_triggered})$ is strictly less than the
cost of executing the full sequence
$\delta(q, e_1) \circ \delta(q_1, e_2) \circ \cdots \circ \delta(q_{n-1}, e_n)$
to reach \texttt{ESCALATING} via the normal transition path.

\textit{Proof sketch.}  The bypass is activated only when either (i) the
operator has directly set $b_\text{flag}$, in which case $O$ is the operator's
intent and the bypass cost equals the operator's decision overhead, or
(ii) a structural invariant violation has been detected in the state machine
itself (an internal consistency failure), in which case no valid transition
sequence exists in the normal path and the bypass is the unique path to
human oversight.  In both cases, the bypass transition dominates by
strict lexical ordering.  $\square$
}}
\end{center}
\end{samepage}

\bigskip

\noindent\textbf{Bypass activation methods (in priority order):}

\begin{enumerate}
  \itemsep0pt
  \item Direct human override via operator command ``\texttt{/bypass}''
     in the agent control channel
  \item Systemic invariant violation detected by the runtime consistency
     checker (internal flag \texttt{CONSISTENCY\_FAIL})
  \item External kill signal received on the agent supervision port
  \item Thermal or hardware-failure signal from the host environment
\end{enumerate}

\bigskip

\noindent\textbf{Bypass bypasses all machine actors.}  The bypass is defined
to override the entire state machine's authority delegation.  When active,
no agent actor---including the orchestrator, any spawned sub-agent, or the
skill executor---may execute a new action.  The only actors permitted to act
during a bypass are: the escalation dispatcher (to notify the human),
the audit logger (to record the bypass event), and the human operator themselves.
This ensures that the bypass cannot be exploited to execute unauthorized work
while a human is supposedly overseeing the situation.

% ---------------------------------------------------------------- %
% 3.6  Timeout Handling
% ---------------------------------------------------------------- %
\subsection{Timeout Handling}

Timeouts are treated as first-class transition events in $\mathcal{T}$.  Each
state $q$ that involves waiting has an associated timeout parameter
$T(q)$.  The runtime maintains a monotonic clock; timeout events are
generated by the clock service and injected into the state machine as
regular transition events.

\bigskip

\begin{center}
\begin{tabular}{|l|c|p{6cm}|}
\hline
\textbf{State} & \textbf{Timeout ($T$)} & \textbf{Expiry action} \\ \hline
BOUNDED & $T_\text{bounded}$ & Enter BAIL\_PENDING; log duration exceeded event \\ \hline
BAIL\_PENDING & $T_\text{bail\_pending}$ & Re-evaluate $\Psi$; if TRUE, re-enter BAIL\_PENDING
with incremented counter; else return to BOUNDED \\ \hline
ESCALATING & $T_\text{escalate}$ & Re-dispatch escalation notification;
increment re-dispatch counter; log timeout \\ \hline
HUMAN\_ACKNOWLEDGED & $T_\text{human}$ & Apply \textbf{timeout default resolution}:
suspend all pending tasks, persist full state to checkpoint, transition to RESOLVED \\ \hline
\end{tabular}
\end{center}

\bigskip

\bigskip

\begin{center}
\fbox{
\parbox{0.92\linewidth}{
\textbf{Timeout Default Resolution Policy.}  When $e_\text{timeout\_human}$ fires
without a prior $e_\text{human\_resolve}$, the system applies the following
resolution in order:

\begin{enumerate}
  \itemsep0pt
  \item Mark all in-flight tasks as \texttt{SUSPENDED\_TIMEOUT}.
  \item Serialize the complete agent state to the checkpoint store.
  \item Cancel all pending non-idempotent side-effects (network writes, external
    API calls) within a grace window of $T_\text{grace} = 30$ seconds.
  \item Transition to \texttt{RESOLVED} and log the timeout event with full context.
  \item Notify the operator via all channels that the timeout default has been
    applied and that manual recovery is required.
\end{enumerate}
}}
\end{center}

\bigskip

\noindent\textbf{Clock monotonicity.}  All timeout timers are sourced from a
single monotonic clock.  Clock adjustment or backwards jumps do not affect
active timers; timers track elapsed real time, not logical time.  Clock
synchronization with an NTP source occurs only in \texttt{IDLE} state and does
not interfere with active protocol operations.

\bigskip

\begin{note}
\textbf{Timeout invariants.}
\begin{itemize}
  \itemsep0pt
  \item $T_\text{human} > T_\text{escalate} > T_\text{bounded}$
  \item $T_\text{grace} \geq 30\text{s}$ and $T_\text{grace} < T_\text{human}$
  \item No timeout can fire while the agent is in \texttt{RESOLVED} state
  \item Timeout counters are reset on every transition to \texttt{IDLE}
\end{itemize}
\end{note}

% ---------------------------------------------------------------- %
% 3.7  Audit Trail and Compliance
% ---------------------------------------------------------------- %
\subsection{Audit Trail and Compliance}

Every state transition, event injection, timeout firing, bypass activation,
human acknowledgement, and resolution instruction is recorded in the
agent's audit log.  The log entry format for a transition event is:

\[
\text{log entry} \;=\;
\bigl(t, \text{agent\_id}, q_\text{from}, q_\text{to}, e, \text{snapshot\_hash}, \text{operator\_id}\bigr)
\]

\bigskip

\noindent The snapshot hash is computed over the full state snapshot at the
time of transition using SHA-256.  This enables the operator and any
post-incident review to verify that the logged snapshot matches the actual
system state at the moment of transition.

The audit log is immutable and append-only.  It is stored in the agent's
workspace at \texttt{data/agentic/audit/} and is synchronized to durable
storage on every entry.  No state transition may occur if the audit log
write fails; the machine is design to halt in place rather than proceed
without a record.

\bigskip

\begin{note}
\textbf{Compliance statement.}  The Bailout Protocol, as specified in this
section, constitutes the complete and authoritative specification of the
agent shutdown and escalation procedure for the Agentic runtime.  Any
implementation that deviates from this specification is non-compliant
and must be classified as a separate, operator-approved protocol with its
own state machine definition.
\end{note}