% ---------------------------------------------------------------
\section{Implementation in the BX3 Framework}
\label{sec:bailout-implementation}
% ---------------------------------------------------------------

The BX3 Framework's upstream accountability guarantee states that every
unresolved exception at a node $n$ must reach its designated human anchor
$h(n)$ in finite time, and that the full propagation path and the human
principal's resulting directive must be permanently attributable.
This guarantee is a structural property of the framework's architecture: it
follows from the immutability of the parent pointer $\alpha(n)$, the
finiteness of the spawn tree depth established at root initialization, and
the designation of human anchors at node initialization.  As a structural
property, it holds under the assumption that the framework's enforcement
mechanisms operate correctly.  The Bailout Protocol is the mechanism that
makes this structural guarantee operationally compulsory.  It is the runtime
execution engine that converts a static architectural commitment into a
deterministic, cryptographically attested, and architecturally enforced
propagation event.  The protocol is not an optional extension layered onto
the framework; it is the translation layer between the framework's structural
commitments and the runtime execution trace that constitutes the definitive
accountability record of every exception in any BX3 node.

The protocol's implementation spans all three BX3 layers, with each layer
fulfilling a distinct and non-overlapping structural role.  The \purpose{}
layer defines and anchors the human principal; the \bounds{} layer houses
the Bounds Engine that detects and classifies constraint violations and emits
the trigger condition; the \fact{} layer enforces the state machine
transitions, records every event in the Forensic Ledger, and provides the
enforcement substrate that makes compliance architecturally compulsory rather
than operationally discretionary.  This separation of concerns is not merely
organizational.  It is the mechanism by which the upstream accountability
guarantee becomes structurally enforced.  No single layer can substitute for
another; the protocol requires all three acting in concert.

% ---------------------------------------------------------------
\subsection{Purpose Layer: Human Accountability Anchor}
\label{sec:bailout-purpose-layer}

The \purpose{} layer encodes the goal state and terminal success criteria
for the agent's current task and serves as the exclusive terminus for any
unresolved exception in a BX3 node.  The human anchor $h(n)$ is provisioned
at node initialization and stored in \fact{}-layer-managed memory.  No
machine actor at any plane P0 through P6 may modify, redirect, or substitute
the human-root identifier.  This immutability is a structural property of the
\purpose{} layer, established at provisioning and invariant throughout the
node's operational lifetime (Invariant I3).

When the Bailout Protocol state machine transitions to
\texttt{HUMAN\_ACKNOWLEDGED}, the \purpose{} layer receives the escalated
exception with its complete diagnostic context chain assembled by the
escalation algorithm (Algorithm~\ref{alg:escalation}).  The layer receives
the originating node identifier; the trigger condition's full sub-components
(bound signal, resolution state, confidence value); every resolution attempt
that failed at each intermediate node; the Bounds Engine's reasoning trace
at each hop; and the timeout states that drove the propagation.  This
completeness is enforced by the \fact{} layer's pre-commit hook.  No machine
actor may truncate, aggregate, redact, or substitute any element of the
diagnostic context before it reaches the \purpose{} layer.

The \purpose{} layer's role in the protocol is strictly bounded.  It does not
monitor the state machine, does not initiate escalation, and does not evaluate
whether the trigger condition has been satisfied.  It receives exceptions that
have already traversed the full propagation chain and issues one of three
binding directives: \texttt{ACK} (continue with current parameters, returning
the node to \texttt{IDLE}), \texttt{RESOLVE} (execute a specific resolution
prescribed by the human principal, entering \texttt{RESOLVED}), or
\texttt{REJECT} (disallow the proposed action and enter a quiescent error
state, also entering \texttt{RESOLVED}).  The directive is recorded in the
authorization ledger via the \fact{} layer before it takes effect.  The
credential binding on the $\sigma_{\text{resolve}}$ event ensures that only
$h(n)$ can emit it; no machine actor at any plane possesses the required
credential.  The \purpose{} layer is the structural realization of the
upstream accountability guarantee's terminal commitment: the exception reaches
\emph{that specific human}, no one else, with no substitutes permitted.

The \purpose{} layer additionally encodes the derivation chain that maps the
human principal's intent to the node's current authorized scope.  This chain
is the runtime artifact of the upstream accountability guarantee: it records
which human principal authorized the node's current operational parameters,
when that authorization was provisioned, and what constraints were set at
that time.  When the Bailout Protocol triggers, the \purpose{} layer uses this
chain to contextualize the exception within the original delegation intent,
enabling the human anchor to evaluate whether the violation represents a
boundary drift within the original delegation scope or a goal misalignment
that requires full task termination.

% ---------------------------------------------------------------
\subsection{Bounds Engine: Triggering the Bail-Out}
\label{sec:bailout-bounds-engine}

The \bounds{} layer defines the provisioned operating range $R(n)$ within
which the agent is authorized to act.  The Bounds Engine is the computational
core of this layer: it evaluates whether proposed actions fall within $R(n)$,
executes resolution procedures, produces candidate resolutions, and emits the
trigger condition that activates the Bailout Protocol.

The primary trigger condition TC (Equation~\ref{TC}) fires when three
conditions hold simultaneously: the observable input $x$ falls outside $R(n)$;
$x$ has not been previously resolved in the current operational session; and
the Bounds Engine's self-assessed confidence in its ability to resolve $x$
falls below the provisioned threshold $\tau(n)$.  The Bounds Engine exercises
no discretion over this evaluation.  It performs a deterministic read against
the operating-range definition and the confidence function, both provisioned at
node initialization and immutable at runtime.  The trigger condition either
fires or it does not; the result drives the state machine transition directly,
without an intervening machine-actor gate.

When TC fires, the state machine transitions from \texttt{IDLE} to
\texttt{BOUNDED} (T1, Table~\ref{tab:bp-transitions}), and the Bounds Engine
begins a time-bounded resolution attempt with timeout $T_{\text{bounds}}$.
During \texttt{BOUNDED}, all resolution paths within current authority remain
open; the Bounds Engine may attempt multiple strategies in sequence or in
parallel, each submission reviewed by the \fact{} layer against the
authorization ledger.  If the \fact{} layer approves a resolution before
$T_{\text{bounds}}$ expires, the state machine returns to \texttt{IDLE} (T2).
If $T_{\text{bounds}}$ expires without approval, the state machine transitions
to \texttt{BAIL\_PENDING} (T3) and the Bounds Engine emits
\texttt{bailout\_signal}.

The transition to \texttt{BAIL\_PENDING} is deterministic and non-negotiable.
No machine actor may suppress the signal, extend $T_{\text{bounds}}$, or
substitute a lower-priority resolution path.  The \fact{} layer's pre-commit
hook enforces this: any attempt to write a resolution approval after
$T_{\text{bounds}}$ has expired is rejected, and the protocol violation is
recorded in the Forensic Ledger.  This architectural enforcement is the
structural distinction between the Bailout Protocol and discretionary
escalation interfaces, where the agent decides whether and when to invoke
human oversight.

The secondary trigger pathway handles explicit \texttt{bailout\_signal}
emissions from the Bounds Engine for cases where the engine detects a
condition that, while not satisfying TC strictly, represents a threshold
violation warranting human review.  The state machine routes this signal
through the same transition logic (T9, Table~\ref{tab:bp-transitions}), and
the diagnostic context chain is appended identically.  This ensures the
protocol captures not only formally defined TC events but also the broader
class of operator-discretionary escalation events that are contextually
critical even when they do not meet the strict TC criteria.

The interaction between the Bounds Engine and the \fact{} layer at the
trigger condition boundary is the architectural point where the upstream
accountability guarantee becomes operative.  Before TC fires, the node
operates under autonomous reasoning within the provisioned operating range
$R(n)$.  The moment TC fires, the \fact{} layer's pre-commit hook blocks all
consequential action and initiates the state machine transition that drives the
protocol toward human escalation.  This boundary is not a decision point.
It is a structural gate whose outcome is determined by the provisioned
parameters, not by the agent's preference.

% ---------------------------------------------------------------
\subsection{State Machine as Fact Layer Extension}
\label{sec:bailout-state-machine}

The Bailout Protocol state machine
$\mathcal{B} = (Q, \Sigma, \rightarrow, q_0, F)$ (Definition~\ref{def:bp-state-machine})
is embedded in the \fact{} layer, not in the \bounds{} layer or any
machine-executed process.  This embedding is architecturally deliberate: a
state machine executing as a machine process can in principle be intercepted,
modified, or circumvented by a compromised machine actor; a state machine
executing as a \fact{}-layer extension is enforced by the layer's
deterministic logic, which operates below the agent's execution level and
cannot be intercepted by any machine actor above it.  The state machine is not
a component that the framework communicates with.  It is a structural
extension of the enforcement layer itself.

The state machine's transition relation $\rightarrow \subseteq Q \times \Sigma \times Q$
is evaluated by the \fact{} layer before any state change is committed.  If
no transition is defined for a $(q, \sigma)$ pair, the machine stalls and the
Bailout Monitor records a protocol violation in the Forensic Ledger.  The state
invariants encode the protocol's core guarantees as properties of the state
machine:

\begin{align}
\text{INV-1}&: \forall n \in \text{nodes},\ \text{state}(n) \in Q \setminus \{\texttt{ESCALATING}\} \iff \neg\text{active\_bailout}(n). \tag{BP-INV-1} \\[6pt]
\text{INV-2}&: \text{state}(n) = \texttt{BAIL\_PENDING} \implies \text{active\_bailout}(n) \land \text{transition}(\texttt{BAIL\_PENDING}, \sigma_{\text{timeout}}) = \texttt{ESCALATING}. \tag{BP-INV-2} \\[6pt]
\text{INV-3}&: \text{state}(n) = \texttt{RESOLVED} \implies \exists d \in \{\texttt{ACK}, \texttt{RESOLVE}, \texttt{REJECT}\} \land \text{directive}(n) = d. \tag{BP-INV-3}
\end{align}

INV-1 establishes the correspondence between the state machine's
\texttt{ESCALATING} state and the presence of an active bailout at the node.
INV-2 encodes the no-machine-actor-discretion property as a state machine
invariant: the transition from \texttt{BAIL\_PENDING} to \texttt{ESCALATING} is
forced by the deterministic transition relation and enforced by the Bailout
Monitor without machine-actor intervention.  The \fact{} layer's pre-commit
hook ensures this transition cannot be intercepted or redirected.  INV-3
ensures that the \texttt{RESOLVED} state is reached only through a genuine
human directive, with the directive type recorded as part of the state
transition.

The state machine's embedded position in the \fact{} layer ensures that the
Bailout Protocol's guarantees are not contingent on the correct behavior of
the \bounds{} layer or any other machine actor.  The \fact{} layer enforces
the protocol's transition relation regardless of what any machine actor
attempted.  The state machine is the \fact{} layer's runtime expression of the
protocol's transition relation in exactly the same way that the Forensic
Ledger is the \fact{} layer's runtime expression of the accountability
guarantee.

% ---------------------------------------------------------------
\subsection{Fact Layer: Enforcement and Forensic Ledger Recording}
\label{sec:bailout-fact-layer}

The \fact{} layer is the enforcement and recording substrate for the Bailout
Protocol.  It maintains the Forensic Ledger: an immutable, SHA-256
hash-chained record of all state transitions, trigger conditions, propagation
events, and human directives across the protocol's lifetime.  The Forensic
Ledger serves two distinct but related functions: it is the authoritative
runtime state store for the Bailout Protocol's state machine, and it is the
enforcement substrate for the framework's accountability properties.  The
pre-commit hooks for the Bailout Protocol, the Sandbox Gate, and the Spatial
Firewall all query the Forensic Ledger to determine the current state of
their respective state machines before any state change is committed.

The Bailout Protocol generates the following entry types in the Forensic
Ledger, each written atomically by the \fact{} layer before the corresponding
state transition is committed:

\begin{itemize}\itemsep=0pt
\item[\texttt{bailout-trigger}] Originating node, exception identifier,
  trigger-condition sub-components (bound signal, resolution state,
  confidence value), and priority function output $\pi(n, x)$.
\item[\texttt{bounds-resolution-attempt}] Each resolution procedure submitted
  by the Bounds Engine during \texttt{BOUNDED} state, including the
  authorization ledger entry reference and the \fact{}-layer approval or
  rejection outcome.
\item[\texttt{bailout-signal}] Explicit \texttt{bailout\_signal} emissions,
  including the Bounds Engine's stated reason.
\item[\texttt{timeout-bounds-expired}] Expiration of $T_{\text{bounds}}$ and
  absence of a \fact{}-layer-approved resolution at expiration.
\item[\texttt{escalation-sent}] Each invocation of \texttt{SendToParent},
  including the pathway selected, the payload hash, the retry count, and the
  acknowledgment status.
\item[\texttt{pathway-exhausted}] Exhaustion of all retry pathways and the
  terminal state reached.
\item[\texttt{human-ack}] Human anchor's acknowledgment of the escalated
  exception, acknowledgment timestamp, and pathway used.
\item[\texttt{human-directive}] Binding directive issued by $h(n)$ --- one of
  \texttt{ACK}, \texttt{RESOLVE}, or \texttt{REJECT} --- and the
  corresponding authorization ledger entry hash.
\end{itemize}

These eight entry types exhaustively cover every state transition in the
Bailout Protocol's state machine (Section~\ref{sec:bp-state-machine}).  The
completeness of this taxonomy is enforced by Invariant I4: every state
transition produces at least one Forensic Ledger entry before the transition
commits.  The \fact{} layer's pre-commit hook rejects any state transition
that has not produced its required entry.  This is not an operational
convention; it is a structural enforcement rule.

The Forensic Ledger entry for a Bailout Protocol escalation constitutes the
definitive accountability record for that exception.  It captures the complete
provenance of the exception from the triggering input through every
intermediate node's resolution attempt to the human principal's directive.
No machine actor may modify, delete, or suppress any entry; the \fact{} layer
enforces the record rather than interprets it.  This is the architectural
distinction between the \fact{} layer and a conventional audit log: the
\fact{} layer does not merely store events.  It actively prevents invalid
state transitions, making it an enforcement mechanism rather than a passive
recorder.

The SHA-256 hash chaining ensures tamper-evidence.  Each entry includes the
hash of the preceding entry, making any post-hoc modification detectable by
recalculating the chain.  An external auditor --- regulator, customer,
investor, or court --- can reconstruct the complete exception history of any
node from the cryptographically chained Forensic Ledger entries and verify
that every unresolved exception reached $h(n)$, that every human directive
was recorded, and that no machine actor suppressed, redirected, or substituted
for the human principal.  The guarantee is not merely asserted.  It is
evidenced.  The combination of hash chaining and completeness enforcement
means that the Forensic Ledger is both tamper-evident and exhaustive: any
gap in the chain is itself a detectable and attributable protocol violation.

% ---------------------------------------------------------------
\subsection{Runtime Expression of the Upstream Accountability Guarantee}
\label{sec:bailout-uag}

The upstream accountability guarantee exists as a structural property of the
BX3 Framework until an exception triggers the Bailout Protocol, at which
point the protocol converts it into an immutable, evidenced execution trace.
This translation from static guarantee to runtime event is the protocol's
defining function.  It is the mechanism that makes the upstream accountability
guarantee operationally real, converting the framework's architectural
commitments into concrete, attributable outcomes.

Without the Bailout Protocol, the upstream accountability guarantee would be a
structural aspiration with no enforcement mechanism.  The human anchor $h(n)$
would be the designated terminus in principle, but nothing would compel
exceptions to reach it.  Machine actors could suppress, redirect, or silently
ignore exception states, and the guarantee would hold only for systems whose
machine actors chose to comply.  The Bailout Protocol converts the guarantee
from a structural aspiration into an architecturally enforced compulsion: the
state machine's deterministic transition relation, the \fact{} layer's
pre-commit hook, and the immutable parent-pointer architecture together ensure
that no unresolved exception state can exist in a BX3 node without propagating
to $h(n)$.

The relationship between the Bailout Protocol and the upstream accountability
guarantee is analogous to the relationship between a formal specification and
its compiled implementation.  The upstream accountability guarantee is the
specification: it states the required property (all unresolved exceptions reach
$h(n)$ in finite time) without specifying the mechanism.  The Bailout Protocol
is the implementation: it provides the deterministic state machine, the
trigger condition, the escalation algorithm, and the Forensic Ledger that
together constitute the enforcement mechanism.  The compilation is performed
by the \fact{} layer, which translates the specification's structural
requirements into the protocol's operational enforcement logic.

The three-layer architecture of the BX3 Framework is what makes this
translation possible.  The \purpose{} layer provides the human anchor
designation that defines the guarantee's terminus.  The \bounds{} layer
provides the trigger condition that defines when the guarantee is invoked.
The \fact{} layer provides the state machine, Forensic Ledger, and
pre-commit enforcement that together make the guarantee structurally
compulsory.  The Bailout Protocol is not a layer in this architecture.
It is the runtime expression of the accountability guarantee across all
three layers simultaneously.