\section{The Bailout Protocol}

\subsection{Overview}

The Bailout Protocol is a deterministic exception-resolution mechanism that halts normal agent operations, bounds resource exposure, and surfaces actionable control to a human operator when automated recovery has failed or the cost threshold has been exceeded. It is not a recovery strategy—it is a controlled surrender protocol. Every agent operating under the BX3 AgentOS framework is subject to the Bailout Protocol; there is no opt-out.

The protocol operates as a finite state machine over a typed state variable $\psi \in \Psi$, where $\Psi = \{\text{IDLE}, \text{BOUNDED}, \text{BAIL\_PENDING}, \text{ESCALATING}, \text{HUMAN\_ACKNOWLEDGED}, \text{RESOLVED}\}$. Transitions are governed exclusively by boolean conditions evaluated against observable runtime signals; no agent, tool, or external actor may transition $\psi$ arbitrarily.

\bigskip

\subsection{Formal State Definitions}

\begin{table}[htbp]
\centering
\caption{State Ontology for the Bailout Protocol}
\begin{tabular}{lp{0.62\linewidth}}
\toprule
\textbf{State} & \textbf{Formal Definition} \\
\midrule
IDLE & $\psi = \text{IDLE} \iff \neg \texttt{bailout\_armed} \land \neg \texttt{cost\_breach} \land \neg \texttt{escalation\_active}$ \\
BOUNDED & $\psi = \text{BOUNDED} \iff \texttt{bailout\_armed} \land \neg \texttt{cost\_breach} \land \neg \texttt{timeout\_fired}$ \\
BAIL\_PENDING & $\psi = \text{BAIL\_PENDING} \iff \left(\texttt{cost\_breach} \lor \texttt{timeout\_fired}\right) \land \neg \texttt{escalation\_active} \land \neg \texttt{human\_seen}$ \\
ESCALATING & $\psi = \text{ESCALATING} \iff \texttt{escalation\_active} \land \neg \texttt{human\_seen}$ \\
HUMAN\_ACKNOWLEDGED & $\psi = \text{HUMAN\_ACKNOWLEDGED} \iff \texttt{human\_seen} \land \neg \texttt{resolved}$ \\
RESOLVED & $\psi = \text{RESOLVED} \iff \texttt{resolved} \lor \left(\neg \texttt{escalation\_active} \land \neg \texttt{cost\_breach} \land \neg \texttt{timeout\_fired} \land \texttt{human\_seen}\right)$ \\
\bottomrule
\end{tabular}
\end{table}

Each predicate ($\texttt{bailout\_armed}$, $\texttt{cost\_breach}$, $\texttt{timeout\_fired}$, $\texttt{escalation\_active}$, $\texttt{human\_seen}$, $\texttt{resolved}$) is a read-only runtime flag observable by the protocol engine. No agent may write these flags directly.

\bigskip

\subsection{State Machine Transitions}

The state transition function is total and deterministic:

\[
\delta : \Psi \times 2^\mathcal{P} \to \Psi
\]

where $\mathcal{P} = \{\texttt{bailout\_armed}, \texttt{cost\_breach}, \texttt{timeout\_fired}, \texttt{escalation\_active}, \texttt{human\_seen}, \texttt{resolved}\}$ is the set of observable predicates.

\begin{table}[htbp]
\centering
\caption{Transition Table}
\begin{tabular}{cp{2.8cm}c}
\toprule
\textbf{From} & \textbf{Condition(s)} & \textbf{To} \\
\midrule
IDLE & $\texttt{bailout\_armed}$ & BOUNDED \\
BOUNDED & $\texttt{cost\_breach} \lor \texttt{timeout\_fired}$ & BAIL\_PENDING \\
BOUNDED & $\neg \texttt{bailout\_armed}$ & IDLE \\
BAIL\_PENDING & $\texttt{escalation\_active}$ & ESCALATING \\
BAIL\_PENDING & $\texttt{resolved}$ & RESOLVED \\
ESCALATING & $\texttt{human\_seen}$ & HUMAN\_ACKNOWLEDGED \\
ESCALATING & $\texttt{resolved}$ & RESOLVED \\
HUMAN\_ACKNOWLEDGED & $\texttt{resolved}$ & RESOLVED \\
RESOLVED & $\neg \texttt{bailout\_armed} \land \neg \texttt{cost\_breach}$ & IDLE \\
\bottomrule
\end{tabular}
\end{table}

\begin{algorithm}[htbp]
\caption{Bailout State Machine Transition Engine}
\begin{algorithmic}[1]
\State $\psi \leftarrow \text{IDLE}$
\Loop
  \State $\mathcal{O} \leftarrow \text{observe\_predicates}()$
  \State $\psi' \leftarrow \delta(\psi, \mathcal{O})$
  \If{$\psi' \neq \psi$}
    \State $\text{log\_transition}(\psi, \psi', \mathcal{O})$
    \State $\psi \leftarrow \psi'$
    \State $\text{on\_enter}(\psi)$
  \EndIf
  \State $\text{sleep}(t_\text{tick})$
\EndLoop
\end{algorithmic}
\end{algorithm}

\bigskip

\subsection{Bail-Out Trigger Condition}

The protocol initiates when any of the following trigger conditions are satisfied:

\begin{definition}{Bail-Out Trigger (BOT)}
The predicate $\texttt{bailout\_triggered}$ is defined as the disjunction:
\[
\texttt{bailout\_triggered} \;\equiv\; \texttt{cost\_breach} \;\lor\; \texttt{timeout\_fired} \;\lor\; \texttt{escalation\_active}
\]
where the constituent predicates are defined as follows:
\end{definition}

\medskip

\noindent\textbf{Cost Breach.} $\texttt{cost\_breach} \;\equiv\; C_\text{accrued} > C_\text{limit}$, where $C_\text{accrued}$ is the cumulative resource cost (compute, API calls, token expenditure) accumulated by the agent's current execution context since the last resolved state, and $C_\text{limit}$ is the per-execution cost ceiling established at agent initialization or by the parent scope. $C_\text{limit}$ is immutable for a given execution context once set.

\medskip

\noindent\textbf{Timeout Fired.} $\texttt{timeout\_fired} \;\equiv\; T_\text{elapsed} > T_\text{deadline}$, where $T_\text{elapsed}$ is the elapsed wall-clock time since the agent began its current task, and $T_\text{deadline}$ is the maximum allowable wall-clock duration for that task. $T_\text{deadline}$ is set at task dispatch and is non-negotiable within the execution context.

\medskip

\noindent\textbf{Escalation Active.} $\texttt{escalation\_active} \;\equiv\; \text{the agent has attempted } N_\text{max\_retries} \text{ automated recoveries without resolution}$. Each recovery attempt increments an atomic counter $N_\text{attempts}$. When $N_\text{attempts} \geq N_\text{max\_retries}$, $\texttt{escalation\_active}$ is set to \texttt{true} and the Bailout Protocol transitions to the ESCALATING state.

\begin{note}
BOT is evaluated continuously in the IDLE and BOUNDED states. Once the protocol has entered BAIL\_PENDING or beyond, the trigger conditions are locked and do not cause re-execution of the trigger logic until RESOLVED.
\end{note}

\bigskip

\subsection{Escalation Algorithm}

When the state machine enters ESCALATING, the following deterministic sequence executes:

\begin{algorithm}[htbp]
\caption{Escalation Sequence}
\begin{algorithmic}[1]
\Function{escalate}{\text{context}: \text{ExecutionContext}}
  \State $\text{ctx} \leftarrow \text{context}$
  \State $\text{ctx.agent\_status} \leftarrow \texttt{PAUSED}$
  \State $\text{ctx.tool\_permissions} \leftarrow \emptyset$ \Comment{Revoke all tool access}
  \State $\text{ctx.resource\_ceiling} \leftarrow 0$ \Comment{Block new resource allocation}
  \State $\text{notification\_payload} \leftarrow \text{assemble\_payload}(\text{ctx})$
  \State $\text{deliver}(\text{notification\_payload})$ \Comment{SMS + email + Discord}
  \State $\text{await\_human\_seen}()$
\EndFunction

\Function{assemble\_payload}{\text{ctx}: \text{ExecutionContext}}
  \State $\text{summary} \leftarrow \text{ctx.task\_description}$
  \State $\text{metrics} \leftarrow \langle C_\text{accrued}, T_\text{elapsed}, N_\text{attempts}, \text{last\_error} \rangle$
  \State $\text{diagnostic\_bundle} \leftarrow \text{ctx.execution\_trace}[\text{last } 50 \text{ steps}]$
  \State \Return $\{\text{summary}, \text{metrics}, \text{diagnostic\_bundle}\}$
\EndFunction

\Function{await\_human\_seen}{}
  \Repeat
    \State $\text{sleep}(t_\text{poll})$
    \State $\text{seen} \leftarrow \texttt{human\_seen} \gets \text{check\_acknowledgment\_channel}()$
  \Until{$\text{seen} \oplus (T_\text{elapsed} > T_\text{max\_wait})$}
  \If{$\neg \text{seen} \land T_\text{elapsed} > T_\text{max\_wait}$}
    \State $\text{retry\_deliver}()$ \Comment{Secondary notification path}
  \EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}

\bigskip

\subsection{Bypass Mechanism}

The bypass exists to address a fundamental structural problem: when all machine actors in an agent's execution path have failed—whether due to a bug, a misconfigured tool, or a cascade of exhausted retries—there is no machine actor remaining capable of detecting or resolving the failure. The bypass is the protocol's answer to this regressio n into incoherence.

\begin{invariant}{Bypass Necessity}
Let $\mathcal{A} = \{a_1, a_2, \ldots, a_n\}$ be the set of all machine actors in an agent's execution path. If $\forall a \in \mathcal{A}, \; \text{state}(a) \in \{\text{FAILED}, \text{TIMEOUT}, \text{MISCONFIGURED}\}$, then no actor in $\mathcal{A}$ can execute the Bailout Protocol independently. A bypass mechanism, triggered by an external watchdog or a structurally isolated safety monitor, must therefore halt the execution path without relying on any actor in $\mathcal{A}$.
\end{invariant}

The bypass is implemented as a structurally isolated process external to the agent's execution context:

\begin{enumerate}
\item A \textbf{watchdog process} $W$ runs as an independent system service, not as an agent-internal actor. $W$ monitors observable resource signals (cost accumulator, wall-clock timer, retry counter) at fixed intervals $t_\text{watchdog}$.
\item If $W$ observes $\texttt{cost\_breach} \lor \texttt{timeout\_fired} \lor (N_\text{attempts} \geq N_\text{max\_retries})$, it writes directly to the protocol state variable $\psi$, transitioning the agent's context to BAIL\_PENDING or ESCALATING without requiring any agent-internal actor to participate.
\item $W$ is itself non-blocking: if $W$ fails, a secondary watchdog $W'$ (configured at system initialization) observes $W$'s heartbeat and escalates independently.
\end{enumerate}

The bypass bypasses all machine actors because it is structurally impossible for any actor in the failed path to observe or correct the failure. This is not a design preference—it is a logical necessity. The bypass mechanism is therefore the only path by which the protocol can guarantee progress toward RESOLVED when internal recovery has been exhausted.

\bigskip

\subsection{Timeout Handling}

Timeouts are treated as first-class signals in the Bailout Protocol, not as error conditions. A timeout indicates that the agent's execution has exceeded its allocated wall-clock budget; it does not indicate failure of the agent itself. Timeout handling follows a strict sequence:

\begin{algorithm}[htbp]
\caption{Timeout Handling Sequence}
\begin{algorithmic}[1]
\Function{handle\_timeout}{\text{context}: \text{ExecutionContext}}
  \State $\text{ctx} \leftarrow \text{context}$
  \State $\Delta T \leftarrow T_\text{elapsed} - T_\text{deadline}$
  \State $\text{ctx.last\_timeout\_reason} \leftarrow \text{classify\_timeout}(\Delta T, \text{ctx})$
  \State $\text{ctx.execution\_trace}[\text{end}] \leftarrow \text{TIMEOUT\_EVENT}(\Delta T)$
  \State $\text{flag\_timeout\_fired}()$
  \State $\text{invoke\_delta\_engine}()$
\EndFunction

\Function{classify\_timeout}{\Delta T, \text{ctx}}
  \If{$\Delta T \leq 30\,\text{s}$} \Return \texttt{SOFT\_TIMEOUT}
  \ElsIf{$\Delta T \leq 300\,\text{s}$} \Return \texttt{MEDIUM\_TIMEOUT}
  \Else{} \Return \texttt{HARD\_TIMEOUT}
  \EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}

Timeout classification determines the urgency weight assigned to the notification payload delivered during escalation. All timeout classifications halt the agent immediately; the classification exists solely to communicate urgency to the human operator.

\begin{note}
After RESOLVED, the timeout flag is cleared. A new execution context begins in IDLE, with a fresh cost accumulator, fresh wall-clock timer, and zero retry counter. There is no carryover of timeout state between execution contexts.
\end{note}

\bigskip

\subsection{Resolved State and Reset}

The RESOLVED state is terminal within an execution context. Upon entering RESOLVED:

\begin{itemize}
\item The agent's execution trace is frozen and archived to durable storage.
\item All resource ceilings are lifted.
\item Tool permissions are restored to their configured defaults.
\item The protocol state $\psi$ returns to IDLE only after a full system initialization cycle—not automatically upon human acknowledgment.
\item A resolution report is generated and delivered to the operator, summarizing: total cost accrued, elapsed time, number of escalation attempts, and root-cause classification (if determinable).
\end{itemize}

\bigskip

\subsection{Invariants}

The following invariants hold for all reachable states:

\begin{enumerate}
\item $\psi \in \Psi$ at all times (no invalid states).
\item No transition can skip RESOLVED and return to IDLE directly; RESOLVED is the only exit gateway.
\item While $\psi \in \{\text{BAIL\_PENDING}, \text{ESCALATING}, \text{HUMAN\_ACKNOWLEDGED}\}$, no new agent task may be dispatched in the affected execution context.
\item The bypass mechanism does not modify $\psi$ directly—it signals the transition engine, which then executes the transition deterministically.
\item $\texttt{human\_seen}$ may only transition from \texttt{false} to \texttt{true}; it is never reset to \texttt{false} during an active protocol cycle.
\end{enumerate}
