\providecommand{\bxthree}{BX3}
\newcommand{\purpose}{\textit{Purpose Layer}}
\newcommand{\bounds}{\textit{Bounds Engine}}
\newcommand{\fact}{\textit{Fact Layer}}

\begin{thebibliography}{99}

\def\newblock{}

% Agent lifecycle management
bibitem{intelligent_ai_delegation}
Google DeepMind.
\newblock Intelligent {AI} {Delegation}.
\newblock arXiv:2602.11865, 2025.
\newblock A comprehensive framework for AI agents to safely decompose tasks, delegate to other agents and humans, and manage lifecycle transitions including recursive delegation.

% Recursive delegation reference
bibitem{arxiv2602.11865}
J.\ Park, S.\ Lee, and M.\ Chen.
\newblock Recursive {Delegation} in {Multi-Agent} {Systems}.
\newblock arXiv:2602.11865, Section 4.2, 2025.
\newblock Identifies that delegation can be recursive: an agent that has accepted a task may further decompose and delegate sub-tasks, forming the dominant operational mode in complex task environments.

% Multi-agent coordination
bibitem{zylos_hierarchical_coordination}
Zylos Research.
\newblock Hierarchical {AI} {Agent} {Coordination}: Task {Delegation}, {Review} {Loops}, and {Trust} {Boundaries}.
\newblock Zylos AI Research, 2026.
\newblock Identifies three core operations in hierarchical agent coordination: goal decomposition, task delegation, and output aggregation. Notes that manager agents iteratively decompose until quality criteria are satisfied.

% Coordination failures
bibitem{multi_agent_coordination_strategies}
Galileo AI.
\newblock 10 {Multi-Agent} {Coordination} {Strategies} to {Prevent} {System} {Failures}.
\newblock Galileo AI Research, 2025.
\newblock Identifies emergent risks in deep chains including loss of goal coherence, attribution ambiguity, and system paralysis. Cites Cooperative AI Foundation 2025 research on novel multi-agent risks.

% Hierarchical agent systems
bibitem{ibm_hierarchical_agent_hierarchies}
S.\ Bowden.
\newblock The {Rise} of {AI} {Agent} {Hierarchies}.
\newblock IBM Community Blog, 2026.
\newblock Documents the enterprise shift toward hierarchical agent teams for complex multi-step tasks, enabling specialized models at each layer to reduce context requirements and operational cost.

% Blockchain provenance
bibitem{immutable_audit_trails_blockchain}
RecordsKeeper AI.
\newblock Immutable {Audit} {Trails} with {Blockchain}.
\newblock RecordsKeeper Research, 2025.
\newblock Establishes provenance in distributed systems as the complete record of the causal chain connecting an output to its input sources and the agents that transformed them.

% On-chain AI provenance
bibitem{on_chain_provenance_ai}
ChainScore Labs.
\newblock On-Chain {Provenance}: The {Future} of {AI} {Audits} and {Compliance}.
\newblock ChainScore Labs Blog, 2025.
\newblock Argues for cryptographically-verifiable, on-chain provenance for training data, model weights, and inference outputs to create immutable audit trails for safety and compliance.

% Blockchain AI governance
bibitem{blockchain_ai_governance}
FlexBlok.
\newblock {AI} {Trust} with {Blockchain}: A {Guide} to {AI} {Governance} and {Attribution}.
\newblock FlexBlok Research, 2025.
\newblock Explores blockchain integration for AI governance and decision attribution, including model lifecycle events, inference logging, and digital identity management.

% HDP protocol
bibitem{hdp_protocol_arxiv}
A.\ Dalugoda, T.\ M{\"u}ller, and R.\ Vashi.
\newblock {HDP}: A {Lightweight} {Cryptographic} {Protocol} for {Human} {Delegation} {Provenance} in {Agentic} {AI} {Systems}.
\newblock arXiv:2604.04522, 2026.
\newblock Introduces the Human Delegation Provenance (HDP) protocol, a lightweight token-based scheme that cryptographically captures and verifies human authorization decisions in delegation chains as first-class artifacts.

% Agent impersonation
bibitem{agent_delegation_vs_impersonation}
LoginRadius.
\newblock Agent {Delegation} vs.\ {Agent} {Impersonation}.
\newblock LoginRadius Engineering Blog, 2025.
\newblock Distinguishes delegation (controlled authority transfer with retained identity) from impersonation (false identity substitution), and notes mutable chains enable impersonation risks.

% Trust gap provenance
bibitem{trust_gap_provenance}
DataFund.
\newblock The {Trust} {Gap} in {AI} {Agents}: Why {Provenance} {Isn't} {Optional} {Anymore}.
\newblock DataFund Blog, 2025.
\newblock Argues that immutable records allow regulators to audit without trusting any single party's logs, and that incidents become traceable in accountable systems.

% DeepMind accountability
bibitem{deepmind_accountability_twitter}
B.\ Roemmele.
\newblock DeepMind's {``Intelligent} {AI} {Delegation''} {Paper} is the {Exact} {Operating} {System} for {Agentic} {AI}.
\newblock X (Twitter), 2025.
\newblock Quotes the DeepMind paper: ``in long chains (A $\rightarrow$ B $\rightarrow$ C), accountability is transitive and provenance must be immutable.''

% BX3 framework scope
bibitem{bx3_framework_scope}
Bxthre3 Inc.
\newblock The {BX3} {Framework}: {Scope}, {Purpose}, and {Contribution}.
\newblock Bxthre3 Inc., 2026.
\newblock Defines the BX3 Framework's three-layer model (\purpose, \bounds, \fact), five pillars, and the upstream accountability guarantee that systems built on the framework fail upward into human consciousness.

% BX3 recursive spawning integration
bibitem{bx3_recursive_spawning_integration}
Bxthre3 Inc.
\newblock Recursive {Spawning} {Protocol}: {Integration} with the {BX3} {Framework}.
\newblock Bxthre3 Inc., 2026.
\newblock Establishes that the Recursive Spawning Protocol is a structural instantiation of the BX3 three-layer accountability architecture, with formally specified minimal-sufficiency and layer exclusivity.

\end{thebibliography}