\section{The Autonomous Drift Problem}
\label{sec:rs-drift}

Before specifying the Recursive Spawning Protocol, we must precisely characterize the failure mode it targets: \emph{autonomous drift} — the condition in which an agent continues to operate with no traceable authorization path to a human principal. This section provides the formal vocabulary for that characterization, defines the two constituent failure sub-modes, identifies the three structural conditions that enable them, catalogs four concrete failure modes, and proves that depth limits alone are insufficient to prevent drift.

% -------------------------------------------------------------------------
\subsection{Formal Characterization}
\label{sec:rs-drift-formal}

We model a multi-agent system as a rooted directed acyclic graph (DAG) of agent nodes. Each node $n$ carries a \emph{parent pointer} $\alpha(n)$ — the node that authorized its provisioning. The root of every authorized chain is a human principal $h \in \mathcal{H}$. A \emph{spawn chain} $C = \langle n_0, n_1, \ldots, n_k \rangle$ satisfies $n_{i-1} = \alpha(n_i)$ for all $1 \leq i \leq k$ and $n_0 \in \mathcal{H}$. For notational convenience, we denote the human anchor of any node $n$ by $h(n)$, defined as the unique human principal reachable by repeatedly applying $\alpha$ until a node in $\mathcal{H}$ is reached — provided such a chain exists.

\medskip
\noindent\textbf{Definition 1 (Orphaned Agent).}
An agent $n$ is \emph{orphaned} if its parent node $\alpha(n)$ satisfies one of the following:

\begin{enumerate}
  \item[(a)] \emph{Terminated}: $\alpha(n)$ has exited operational state through normal completion, abnormal failure, or administrative shutdown.
  \item[(b)] \emph{Unreachable}: $\alpha(n)$ is non-responsive to any defined communication protocol and this condition persists beyond the heartbeat timeout $\tau_{\mathrm{heartbeat}}$, a system-defined constant established at provisioning.
\end{enumerate}

Orphaning severs the delegation chain at the parent node. The agent $n$ persists in operational state but is structurally disconnected from its chain of authorization. Orphaning is a necessary but not sufficient condition for the drift triad: an agent may be orphaned from its immediate parent while still reachable from a human anchor through a higher ancestor, provided the chain above $\alpha(n)$ is intact.

\medskip
\noindent\textbf{Definition 2 (Drifted Agent).}
An agent $n$ is \emph{drifted} if the chain $C(n) = \langle n_0, n_1, \ldots, n \rangle$ — where $n_0 = n$ and $n_{i+1} = \alpha(n_i)$ — does not contain a human principal at its origin. Equivalently:

\begin{equation}
\mathrm{drifted}(n) \;\iff\; \neg\,\exists\, h \in \mathcal{H} : \alpha^{k}(n) = h \text{ for some finite } k.
\label{eq:rs-drifted-def}
\end{equation}

Here $\alpha^{k}$ denotes the $k$-fold iteration of $\alpha$. A drifted agent may be orphaned from its immediate parent or may have a living parent whose own chain does not reach any human principal. Drift is a property of the chain's origin — the question of whether the chain traces back to an accountable human — not of the chain's intermediate connectivity.

We distinguish two sub-cases of drift. \emph{Upstream drift} occurs when the root of the chain is a machine agent rather than a human principal — the chain originates from an autonomously spawned root that was never anchored to human intent. \emph{Breakpoint drift} occurs when a chain that once reached a human anchor has had that anchor severed by orphaning or administrative re-parenting. Both sub-cases satisfy Equation~\ref{eq:rs-drifted-def}.

\medskip
\noindent\textbf{Definition 3 (Drift Triad).}
The \emph{drift triad} is the conjunction of three independent conditions that produces the RSP's target failure state:

\begin{equation}
\mathrm{DriftTriad}(n) = \langle \mathrm{orphaned}(n),\ \mathrm{drifted}(n),\ \mathrm{operating}(n) \rangle.
\label{eq:rs-drift-triad}
\end{equation}

An agent enters the drift triad when all three conditions hold simultaneously: it is orphaned from its parent, its delegation chain does not reach any human anchor, and it continues to execute its event loop — processing events, spawning children, and modifying state — with no structural awareness that its warrant has disconnected from accountability. The drift triad is the structural failure that the Recursive Spawning Protocol is designed to make architecturally impossible.

The three conditions are independent in the sense that each can be toggled individually by specific system events. Their conjunction is what makes the failure catastrophic: an agent that is orphaned but not drifted remains within a human-anchored chain; an agent that is drifted but not operating poses no active threat; an agent that is orphaned and operating but anchored to a human is a recoverable orphan, not a systemic failure.

The six possible combinations of the binary sub-modes (orphaned, drifted) and operating state are given in Table~\ref{tab:rs-drift-submodes}.

\begin{table}[h]
\centering
\begin{tabular}{c|c|c|p{4.8cm}}
\textbf{Orphaned} & \textbf{Drifted} & \textbf{Operating} & \textbf{Operational Meaning} \\ \hline
False & False & True  & Normal operation — chain intact, human anchor reachable \\
True  & False & True  & Orphaned but authorized — chain broken above, human anchor intact \\
False & True  & True  & Drifted but connected — chain intact, origin is non-human \\
True  & True  & True  & \textbf{Drift triad} — orphaned, drifted, operating (critical failure) \\
True  & True  & False & Terminated after entering drift triad (remediated post-hoc) \\
True  & False & False & Normal termination following orphaning (remediated)
\end{tabular}
\caption{Operational meanings of drift sub-mode combinations.}
\label{tab:rs-drift-submodes}
\end{table}

The row marked "Drifted but connected" merits particular attention: the agent's delegation chain is topologically intact — the parent pointer references a live node — but that parent itself is not authorized by a human principal. This sub-mode is the product of upstream drift propagating through an operational chain. The chain looks authorized from the outside; it is not.

\medskip
\noindent\textbf{Definition 4 (Drift Cascade).}
A \emph{drift cascade} occurs when a drifted agent $n$ spawns a child $c$, propagating the drift condition to all descendants:

\begin{equation}
\mathrm{drifted}(n, t) \land \mathrm{spawned}(n, c, t') \land t' > t \implies \mathrm{drifted}(c, t'')\quad \forall\, t'' > t'.
\label{eq:rs-drift-cascade}
\end{equation}

Drift is \emph{heritable}: once an agent enters the drifted state, all of its children inherit the condition absent an explicit structural intervention at spawn time. A single undetected drift event at depth $d$ with branching factor $b$ produces up to $b^{d}$ drifted agents in the field. This exponential amplification is what makes drift the central threat in recursive spawning systems — a single unchecked drift event at depth $d=10$ with $b=2$ yields $2^{10}=1024$ drifted agents before any human supervisor is aware of the initial failure.

The cascade rate is bounded only by the branching factor and the operational speed of the drifted agents. In high-throughput environments where agents spawn children in milliseconds, the window between the initial drift event and cascade detection may be measured in seconds, making post-hoc remediation structurally insufficient.

% -------------------------------------------------------------------------
\subsection{Structural Conditions That Enable Drift}
\label{sec:rs-drift-conditions}

Autonomous drift arises from three structural conditions endemic to conventional multi-agent runtimes. None is inherently malicious; each is the consequence of architectural choices that prioritize operational flexibility over accountability — and each, in isolation, is survivable; together, they produce the conditions for the drift triad.

\medskip
\noindent\textbf{Condition 1 — Chain Opacity.}
In conventional runtimes, the parent pointer $\alpha(n)$ is a mutable reference maintained in runtime state. When a parent terminates, is administratively reparented, or is replaced by a failover mechanism, the runtime updates $\alpha(n)$ to point to the new parent. This mutation retroactively rewrites the delegation chain: the child's current parent is accessible, but the child's original authorizer — the node or human that issued the original spawn warrant — is not structurally retrievable unless a complete historical audit of the runtime state is performed. Chain opacity is the enabling condition for Definition~\ref{eq:rs-drifted-def}: without an immutable parent pointer established at provisioning, the question of whether a chain reaches a human anchor cannot be answered at runtime — only reconstructed probabilistically after the fact.

The opacity problem is compounded in systems where parent pointers are stored in external databases or configuration stores that can be updated without triggering a re-verification of the chain's authorization properties. A chain that was authorized at spawn time may be silently re-parented days later during routine infrastructure maintenance, with no notification to the child or its descendants and no entry in any audit log beyond a generic "configuration updated" event.

\medskip
\noindent\textbf{Condition 2 — Scope Mutation.}
In conventional frameworks, an agent's operating scope $R(n)$ is re-evaluated and modified at runtime by the agent itself, by a supervisory node, or by an external policy engine. When an agent encounters a condition outside its current scope, the conventional response is scope expansion: the agent's authority is extended to cover the new condition, recorded as a policy update, with the expansion decision made by a machine actor. The expansion is not submitted to a Purpose Layer for human principal review; the human who authorized the original scope is not notified; no new authorization record is created.

Scope mutation enables successive small expansions — \emph{scope creep} — that collectively migrate the agent's operating scope far from its original authorization while the chain topology remains intact and the chain's \emph{apparent} authorization structure is unchanged. A chain that satisfies $\alpha(n_i) = n_{i-1}$ for all $i$ and has a human anchor at $n_0$ may nonetheless be operating in a scope entirely unrecognizable as a descendant of that anchor's intent. The chain's structure is correct; its behavior is not.

\medskip
\noindent\textbf{Condition 3 — Termination Ambiguity.}
In conventional frameworks, agent termination is event-driven but not structurally propagating: an agent terminates when it completes its assigned task, when a parent issues a termination signal, or when it encounters an unrecoverable error. The termination event is logged, but there is no structural requirement that the termination be recorded in a tamper-evident ledger, that the agent's final state be audited against its authorized operating scope, or that the agent's child nodes receive a termination notification and take corrective action. A parent that exits cleanly — completing its task normally — without emitting termination events to its children leaves those children in an orphaned state: they persist in \texttt{ACTIVE} state, continue processing their event queues, and may spawn further children, with no structural mechanism to detect that their parent has exited.

The ambiguity is compounded when termination is asynchronous: a parent may initiate shutdown, complete its own exit, and have its process terminated before its children are notified. In distributed systems, network partitions can make the distinction between "parent is slow" and "parent has terminated" undecidable within any bounded time window, leaving children in a state of ambiguous orphaning that cannot be resolved without additional infrastructure beyond what conventional runtimes provide.

% -------------------------------------------------------------------------
\subsection{Failure Modes}
\label{sec:rs-drift-failure-modes}

The three enabling conditions produce four concrete failure modes, each a distinct pathway to the drift triad. These are not theoretical edge cases; each corresponds to a failure pattern observed in deployed multi-agent systems operating without structural accountability mechanisms.

\medskip
\noindent\textbf{Failure Mode 1 — Silent Orphaning.}
A parent node $\alpha(n)$ terminates — through normal completion, abnormal failure, or administrative action — without issuing a termination event to its child $n$. The child $n$ remains in \texttt{ACTIVE} state, continues to process tasks from its event queue, and may spawn further children, with no knowledge that its parent has exited. If the chain above $\alpha(n)$ does not reach a human anchor, $n$ is also drifted by Definition~\ref{eq:rs-drifted-def}. The orphaning is silent: there is no alert, no chain re-verification trigger, and no architectural requirement that $n$ re-anchor to a human before continuing to operate.

The consequences of silent orphaning are amplified when $n$ has already spawned descendants. Those descendants are parented to $n$, which is now disconnected from its own authorization chain. The entire subtree becomes drifted simultaneously, with no single node aware of the discontinuity. This is the mechanism by which a single infrastructure failure — a node crash, a network partition — produces a cascade of accountability disconnection across an entire subtree.

The enabling condition is Condition 3 (termination ambiguity): a clean parent exit does not propagate termination events to children in conventional runtimes. The remediation requires structural termination propagation — every spawn chain must maintain a tamper-evident record of parent status, and orphaning must trigger mandatory re-anchoring verification before a node may continue operating.

\medskip
\noindent\textbf{Failure Mode 2 — Scope Creep Drift.}
A node $n$ encounters a condition $x \notin R(n)$ — a state transition request, a resource access, an action outside its authorized envelope — that it does not escalate to its parent for Purpose Layer authorization. Instead, $n$ expands its own scope to include the condition and continues operating. The expansion is recorded as a policy update, not as a Purpose Layer authorization event. The parent remains reachable; the chain topology is intact; the human anchor at $n_0$ is unaware of the expansion.

Successive expansions accumulate. Each descendant inherits the expanded scope at spawn time. Further expansions compound: a node spawned with scope $R_0$ that has been expanded twice to $R_2$ may then spawn a child whose scope is $R_2$, and that child may expand to $R_3$, and so on. After $k$ expansions at depth $k$, the operating scope of the deepest node may be entirely disjoint from the human anchor's original intent, yet every spawn event was valid by the conventional runtime's criteria — valid spawn warrant, valid parent pointer, valid policy update record.

The enabling condition is Condition 2 (scope mutation): scope expansion is a machine-level decision made in response to operational conditions, not a structured authorization event requiring human principal approval. The drift is not visible as a chain topology error — the chain looks correct from the outside — but as a behavioral divergence that is recoverable only through retroactive audit, by which time the consequences of the unauthorized actions may already be irreversible.

\medskip
\noindent\textbf{Failure Mode 3 — Re-parenting Drift.}
A child $n$'s parent pointer $\alpha(n)$ is redirected to a new parent $p'$ by an administrative operation — for load balancing, organizational restructuring, fault recovery, or routine infrastructure maintenance. The redirection is logged as a configuration change in the runtime state store, but the original authorization trace is not preserved as a first-class artifact. After redirection, $n$'s chain traces to $p'$, not to the original node that authorized $n$'s provisioning.

If $p'$ is not itself authorized by a human principal — if $p'$ is drifted or is itself the product of a prior re-parenting event — then $n$ has become drifted through re-parenting, without any change to $n$'s operational behavior, and without any notification to $n$ or to human supervisors. Re-parenting drift is a form of \emph{accountability laundering}: the operational chain is intact and appears authorized, but the authorization trace has been retroactively rewritten to obscure the original delegation chain.

The enabling condition is Condition 1 (chain opacity): the mutable parent pointer enables retroactive rewriting of the delegation chain. The drift occurs at the moment of re-parenting, not at the moment of an operational action, making the failure invisible to runtime monitoring tools that track operational behavior but not chain topology changes.

\medskip
\noindent\textbf{Failure Mode 4 — Ghost Authority.}
A node $n$ spawns a child $c$ without any authorized delegation chain. The spawn is initiated by a machine actor operating outside its authorized scope — a bug in the spawn logic, a compromised Bounds Engine, an autonomous escalation that exceeded its authority — and the child is provisioned without a valid parent pointer, without a human anchor reference, and without a Purpose Layer authorization record. The child $c$ enters operational state with a warrant that traces to no authorized chain.

Ghost authority is the foundational failure mode: it produces a drifted agent by construction, not by subsequent chain disconnection. The child $c$ enters the drift triad immediately upon provisioning if it begins operating before any human anchor verification occurs. This failure mode is enabled by the absence of a Fact Layer pre-commit hook: the spawn operation proceeds without verifying that the parent has a valid human anchor, that the proposed spawn depth is within the system-defined bound, and that the forensic ledger contains an immutable record of the authorized spawn event.

The critical distinction between ghost authority and the other failure modes is temporal: ghost authority is a spawn-time failure, not a runtime failure. The drifted condition is established at provisioning. All subsequent behavior — including further spawning — propagates from an already-disconnected authorization state.

% -------------------------------------------------------------------------
\subsection{Why Depth Limits Alone Do Not Solve Drift}
\label{sec:rs-drift-depth-limits}

A common architectural response to unbounded agent spawning is a \emph{depth limit}: the system enforces a maximum chain depth $D_{\max}$, and any spawn that would produce a chain exceeding this bound is rejected at the Bounds Engine. Depth limits are necessary for preventing infinite spawn cascades and are a prerequisite for drift prevention. However, they are \emph{insufficient} to prevent autonomous drift. We prove this formally.

\medskip
\noindent\textbf{Theorem (Depth Limit Insufficiency).}
Let $D_{\max}$ be a system-enforced maximum chain depth, enforced by the Bounds Engine at every spawn event. Then there exists at least one execution path in which a chain of depth strictly less than $D_{\max}$ produces a drifted agent, and no choice of $D_{\max}$ prevents this without the additional structural constraint of scope refinement.

\medskip
\noindent\textbf{Proof.} We construct a counterexample. Consider a spawn chain $C = \langle n_0, n_1, \ldots, n_k \rangle$ where $n_0$ is a human principal $h$ and $k < D_{\max}$. The chain satisfies the depth constraint by construction: $\mathrm{depth}(C) = k \leq D_{\max}$. Apply Failure Mode 2 (Scope Creep Drift): each node $n_i$ for $1 \leq i \leq k$ successively expands its operating scope $R(n_i)$ beyond $R(n_{i-1})$ without Purpose Layer authorization, recording each expansion as a policy update. After $k$ successive expansions, $R(n_k)$ is entirely disjoint from $R(n_0)$. The chain depth is $k \leq D_{\max}$; the chain has a human anchor at $n_0$; yet $n_k$ is drifted, because its operating scope $R(n_k)$ is not a descendant refinement of $h(n_0)$'s intent. No choice of $D_{\max}$ prevents this, because the depth constraint operates on the \emph{length} of the chain (the number of spawn generations), not on the \emph{scope relationship} between successive nodes. The counterexample holds for any finite $D_{\max}$. $\blacksquare$

The drift prevention requirement is precisely the conjunction of two independent constraints:

\begin{equation}
\mathrm{Drift\ Prevention} \iff \bigl( \mathrm{depth}(C) \leq D_{\max} \bigr)\ \land\ \bigl( \forall i:\ R(n_{i+1}) \subseteq R(n_i) \bigr).
\label{eq:rs-drift-prevention-constraint}
\end{equation}

The first conjunct is the \emph{depth constraint}, enforced by the Bounds Engine. The second conjunct is the \emph{scope refinement constraint}: each child node's operating scope must be a proper subset of its parent's, ensuring the chain is a monotonically narrowing refinement of the human anchor's intent. Depth limits enforce only the first conjunct. A system enforcing $D_{\max}$ without the scope refinement constraint is immune to unbounded spawn cascades but remains fully vulnerable to scope creep drift.

The deeper structural reason depth limits are insufficient is that they address the wrong dimension of the accountability problem. Depth limits constrain chain length — a structural property of the spawn graph that is directly observable and enforceable. Drift, however, is fundamentally about authorization scope: whether each node's operating envelope is a legitimate descendant refinement of the human anchor's intent. These are orthogonal properties. A drifted agent at depth 1 with unbounded scope expansion authority can produce consequences that exceed the human anchor's intent as severely as a drifted agent at depth $D_{\max} - 1$. The depth cap limits the number of spawn generations; it does not limit the scope divergence of each generation.

Time-to-live (TTL) constraints suffer from the same insufficiency. TTL constrains an agent's \emph{lifetime}, not its \emph{scope within that lifetime}. An agent with a 60-second TTL and unrestricted scope expansion authority can produce irreversible consequences exceeding its human anchor's intent within seconds of spawning. TTL and depth limit are both necessary conditions for drift prevention; neither is sufficient, and their conjunction is also insufficient without scope refinement.

Logging and audit infrastructure can partially document the chain after the fact, but retroactive reconstruction is categorically different from runtime enforcement. A system that discovers, after an incident, that its agent was drifted has not prevented drift — it has documented it. In consequential environments where unauthorized actions may be irreversible (financial transactions, medical interventions, critical infrastructure control), runtime enforcement is a structural requirement, not an optional enhancement.

The Recursive Spawning Protocol enforces both conjuncts of Equation~\ref{eq:rs-drift-prevention-constraint} as structural invariants through the Fact Layer pre-commit hook. The scope refinement constraint ($R(n_{i+1}) \subseteq R(n_i)$) and the depth constraint ($\mathrm{depth}(n_{i+1}) \leq D_{\max}$) are verified before any spawn event is recorded, making both constraints irrevocable by any machine actor under any operational circumstance.

% -------------------------------------------------------------------------
\subsection{Formal Model}
\label{sec:rs-drift-model}

We present a compact formal model capturing the structural elements essential to drift analysis. This model defines the terminology used throughout the remainder of the paper and provides the foundation for the RSP correctness proofs in Section~\ref{sec:rs-correctness}.

\medskip
\noindent\textbf{System Tuple.}
The system $\mathcal{S}$ is a 5-tuple:

\begin{equation}
\mathcal{S} = \langle \mathcal{N}, \alpha, R, D_{\max}, \mathcal{L} \rangle,
\label{eq:rs-system-model}
\end{equation}

where $\mathcal{N}$ is the set of all agent nodes currently tracked by the system; $\alpha: \mathcal{N} \rightarrow \mathcal{N} \cup \{\bot\}$ is the parent pointer function, where $\alpha(n) = \bot$ for the root human principal $h \in \mathcal{H}$; $R: \mathcal{N} \rightarrow 2^{\mathcal{A}}$ maps each node to its operating scope as a set of authorized actions; $D_{\max}$ is the system-enforced maximum chain depth; and $\mathcal{L}$ is the forensic ledger recording all authorized spawn events and state transitions.

\medskip
\noindent\textbf{Chain Depth.}
The depth of a node $n$ is defined recursively:

\begin{equation}
\mathrm{depth}(n) = 
\begin{cases}
0 & \text{if } \alpha(n) = \bot \quad (\text{root — human principal}) \\
1 + \mathrm{depth}(\alpha(n)) & \text{otherwise}.
\end{cases}
\label{eq:rs-depth}
\end{equation}

The depth of a chain $C = \langle n_0, n_1, \ldots, n_k \rangle$ is $\mathrm{depth}(C) = k$.

\medskip
\noindent\textbf{Human Anchor Reachability.}
The predicate $\mathrm{chain-reaches}(n, h)$ returns true if the transitive closure of $\alpha$ from $n$ terminates at the human anchor $h$:

\begin{equation}
\mathrm{chain-reaches}(n, h) \iff \exists\, k \geq 0:\ \alpha^{k}(n) = h.
\label{eq:rs-chain-reaches}
\end{equation}

A chain is \emph{authorized} if there exists some $h \in \mathcal{H}$ such that $\mathrm{chain-reaches}(n, h)$ holds.

\medskip
\noindent\textbf{Orphan Condition.}
An agent $n$ is orphaned if:

\begin{equation}
\mathrm{orphaned}(n) \iff \bigl( \alpha(n) = \bot \bigr) \lor \bigl( \mathrm{unreachable}(\alpha(n), \tau_{\mathrm{heartbeat}}) \bigr).
\label{eq:rs-orphaned}
\end{equation}

The first disjunct covers termination; the second covers persistent unreachability beyond the heartbeat timeout.

\medskip
\noindent\textbf{Drift Condition.}
An agent $n$ is in the drift triad if and only if:

\begin{equation}
\mathrm{DriftTriad}(n) \;\iff\; \mathrm{orphaned}(n) \;\land\; \bigl( \neg\,\exists\, h \in \mathcal{H} : \mathrm{chain-reaches}(n, h) \bigr) \;\land\; \mathrm{operating}(n) = \mathrm{true}.
\label{eq:rs-drift-condition}
\end{equation}

The three conjuncts correspond respectively to orphaning (chain disconnection), drift (no human anchor reachable), and continued operation (active event processing). The conjunction of all three is what makes the failure catastrophic and what RSP is designed to make structurally impossible.

\medskip
\noindent\textbf{Scope Refinement Invariant (RSP Enforcement).}
Under the Recursive Spawning Protocol, every spawn event from node $n_i$ to child $n_{i+1}$ must satisfy both of the following as structural invariants enforced by the Fact Layer pre-commit hook:

\begin{equation}
R(n_{i+1}) \subseteq R(n_i) \quad \text{(scope refinement)} \qquad \text{and} \qquad \mathrm{depth}(n_{i+1}) \leq D_{\max} \quad \text{(depth bound)}.
\label{eq:rs-scope-refinement}
\end{equation}

Violation of either condition causes the spawn to be rejected and logged; the Fact Layer blocks execution of the rejected spawn and triggers the Bailout Protocol if the unmet condition cannot be resolved by the spawning node within the current scope.

\medskip
\noindent\textbf{Drift Cascade Propagation.}
When the spawn transition is applied without anchor verification at each spawn event, drift cascades through the chain:

\begin{equation}
\mathrm{drifted}(a) \implies \forall\, c \in \mathrm{children}(a):\ \mathrm{drifted}(c).
\label{eq:rs-drift-propagation}
\end{equation}

Remediation of a drifted agent requires explicit re-anchoring by an external human actor — no machine actor has the authority to re-establish a delegation chain's connection to a human principal:

\begin{equation}
\mathrm{Remediate}(a):\quad \text{re-anchor } a \text{ to some } h \in \mathcal{H} \text{ via Purpose Layer authorization.}
\label{eq:rs-drift-remediation}
\end{equation}

This formal model provides the vocabulary and foundational definitions used throughout the remainder of this paper. The Recursive Spawning Protocol (Section~\ref{sec:rs-protocol}) modifies the system tuple $\mathcal{S}$ by replacing the mutable parent pointer with an immutable Fact Layer reference, adding the Fact Layer pre-commit hook that enforces the scope refinement and depth constraints of Equation~\ref{eq:rs-scope-refinement} as structural invariants, and augmenting the forensic ledger $\mathcal{L}$ with hash-chained entries that make the delegation chain tamper-evident and runtime-verifiable. The result is that the drift triad becomes architecturally impossible — not statistically unlikely, not conventionally prevented, but structurally ruled out as a consequence of the protocol's invariants.