\section{The Bailout Protocol}\label{sec:bailout}

The Bailout Protocol is the mandatory exception-escalation mechanism of the \bxthree{} Framework. It is not an error-handling routine selected at runtime, nor an optional escalation pathway available to a system that chooses to invoke it. It is an architectural compulsion: any unresolved exception state encountered by any node in a \bxthree{} system \emph{must} propagate upward through the accountability chain until it reaches a named human principal, and no machine actor along that chain may absorb, resolve, or suppress the exception. This section provides the complete formal specification: the deterministic state machine governing exception propagation, the precise conditions governing each transition, the bail-out trigger condition, the escalation algorithm, the bypass mechanism enforcing the no-machine-actor property, and the timeout handling that guarantees liveness under all failure conditions.

The protocol operates against the backdrop of the \bxthree{} three-layer model. The \purpose{} layer carries intent, judgment, and final authority; it defines the operating range $R(n)$ for each node $n$ and receives all escalated exceptions. The \bounds{} layer carries bounded reasoning and constrained execution within $R(n)$; it evaluates whether proposed actions fall within the node's provisioned authority and initiates bailout when they do not. The \fact{} layer carries deterministic enforcement and forensic auditability; it verifies proposed actions against the authorization ledger, enforces the state machine's transition relation, and maintains the append-only Forensic Ledger as an immutable record of every protocol event. Each node carries an immutable parent pointer $\alpha(n)$ established at provisioning and a reference to its human accountability anchor $h(n)$ — a named human principal — that is similarly immutable for the node's operational lifetime. The Bailout Protocol exploits this structure to guarantee that exception states propagate to $h(n)$ without passing through any intermediate machine actor as a terminus.

% --------------------------------------------------------------
\subsection{State Machine Definition}\label{sec:state-machine}
% --------------------------------------------------------------

The Bailout Protocol is formally modeled as a deterministic finite state machine embedded in the \fact{} layer of every \bxthree{} node:

\[
\mathcal{B} \;=\; (Q,\; q_0,\; \Sigma,\; \rightarrow,\; Q_F)
\]

\begin{itemize}\itemsep=0pt
\item[$Q$] $\{\;\text{IDLE},\; \text{BOUNDED},\; \text{BAIL\_PENDING},\; \text{ESCALATING},\; \text{HUMAN\_ACKNOWLEDGED},\; \text{RESOLVED}\;\}$ is the finite set of protocol states.
\item[$q_0$] $\text{IDLE}$ is the designated initial state, entered at node startup and after every successful directive resolution.
\item[$\Sigma$] The input alphabet comprising trigger events $e_{\text{trigger}}$, authorization confirmations $e_{\text{auth}}$, timeout signals $e_{\text{timeout}}$, human acknowledgments $e_{\text{ack}}$, binding resolution directives $e_{\text{resolve}}$, and rejection directives $e_{\text{reject}}$.
\item[$\rightarrow$] $\; \subseteq Q \times \Sigma \times Q$ is the deterministic transition relation (Table~\ref{tab:bailout-transition}).
\item[$Q_F$] $\{\text{RESOLVED}\}$ is the set of terminal accepting states.
\end{itemize}

A node's state $q \in Q$ is stored in the node's \fact{} layer worksheet and is observable by its parent node and by the Bailout Monitor. State updates are atomic: no intermediate or partially-executing states are observable externally. The machine is \emph{deterministic}: for every $(q, \sigma) \in Q \times \Sigma$, at most one $q' \in Q$ satisfies $(q, \sigma, q') \in \rightarrow$. Determinism is enforced by the Bailout Monitor's priority function $\pi$, which totally orders simultaneous trigger conditions and selects the highest-priority enabled transition before committing any state change. If no transition is defined for a $(q, \sigma)$ pair, the machine stalls and the Bailout Monitor records a protocol violation in the Forensic Ledger.

\begin{table}[htbp]
\caption{Bailout Protocol state transition relation $\rightarrow \subseteq Q \times \Sigma \times Q$. Rows are current state; columns are input events. Entries denote the next state. Blank cells indicate that no transition is defined — the machine stalls and the Bailout Monitor records a protocol violation.}
\label{tab:bailout-transition}
\begin{tabular}{@{}lcccccc@{}}
\toprule
 & \multicolumn{6}{c}{Input event} \\
\cmidrule(l){2-7}
Current state & $e_{\text{trigger}}$ & $e_{\text{auth}}$ & $e_{\text{timeout}}$ & $e_{\text{ack}}$ & $e_{\text{resolve}}$ & $e_{\text{reject}}$ \\
\midrule
IDLE              & BOUNDED            & — & — & — & — & — \\
BOUNDED           & — & HUMAN\_ACKNOWLEDGED & BAIL\_PENDING & — & — & — \\
BAIL\_PENDING     & — & — & ESCALATING & — & — & — \\
ESCALATING        & — & HUMAN\_ACKNOWLEDGED & ESCALATING & IDLE & RESOLVED & RESOLVED \\
HUMAN\_ACKNOWLEDGED & — & — & — & IDLE & RESOLVED & IDLE \\
RESOLVED          & — & — & — & — & — & — \\
\bottomrule
\end{tabular}
\end{table}

\bigskip
\noindent\textbf{State definitions.}

\begin{itemize}\itemsep=0pt
\item[\textbf{IDLE.}] The node is operating within its provisioned operating range $R(n)$. No exception condition is active. The \fact{} layer dispatches normal action requests through the standard execution path. The node may transition to BOUNDED upon receipt of a trigger event satisfying the bail-out trigger condition (TC), defined in \S~\ref{sec:bailout-trigger}.

\item[\textbf{BOUNDED.}] An exception $e$ has been detected by the \bounds{} layer and is being held at the \bounds{}--\fact{} layer boundary. A time-bounded resolution attempt is in progress. All resolution paths within current authority remain open. If the Bounds Engine produces a resolution $r$ that the \fact{} layer approves before $T_{\text{bounds}}$ expires, the node returns to IDLE. If all resolution paths are exhausted or $T_{\text{bounds}}$ expires, the state advances to BAIL\_PENDING.

\item[\textbf{BAIL\_PENDING.}] The node has determined — either through an explicit \texttt{bailout\_signal} or through $T_{\text{bounds}}$ expiration without a \fact-layer-approved resolution — that the exception exceeds its current authority. The Bailout Protocol is formally activated. This is the last architectural point at which a machine actor could theoretically suppress escalation. It cannot: entry to BAIL\_PENDING triggers an immediate atomic transition to ESCALATING via $e_{\text{timeout}}$ (T4), with no dwell time and no machine-actor intervention permitted.

\item[\textbf{ESCALATING.}] The exception is propagating upward toward $h(n)$. Each intermediate node receives the exception, appends its diagnostic context to the propagation chain, and forwards the exception via \texttt{SendToParent} without attempting local resolution. The state remains ESCALATING until either (a) the human anchor acknowledges receipt ($e_{\text{ack}}$: T5), (b) the human anchor issues a binding directive ($e_{\text{resolve}}$ or $e_{\text{reject}}$: T7), or (c) all retry pathways timeout after $N_{\max}$ retries at $T_{\text{cap}}$.

\item[\textbf{HUMAN\_ACKNOWLEDGED.}] The human anchor has received the exception, observed the diagnostic context, and acknowledged it. The node awaits a directive. No machine actor may issue, simulate, or substitute a directive in this state. The node remains until the human principal issues \texttt{ACK} (return to IDLE) or a binding directive \texttt{RESOLVE} / \texttt{REJECT} (enter RESOLVED).

\item[\textbf{RESOLVED.}] The human principal has issued a directive and the protocol has recorded it in the authorization ledger. The directive is one of: \texttt{ACK} (continue with current parameters), \texttt{RESOLVE} (execute a specific binding resolution), or \texttt{REJECT} (disallow the proposed action and enter a quiescent error state). This state is terminal for the current protocol run.
\end{itemize}

\bigskip
\noindent\textbf{State invariants.} The following invariants hold for all reachable states of $\mathcal{B}$:

\begin{align}
\text{INV}_1 &: \forall n \in \text{nodes},\; \text{state}(n) \in Q \setminus \{\text{ESCALATING}\} \iff \neg\text{active\_bailout}(n). \tag{INV-1}
\end{align}

That is, a node occupies a non-escalating state \emph{iff} it has no active bailout propagation in flight. A node may carry a bailout history (a ledger entry recording a past propagation) while in IDLE, but no active propagation may be underway.

\begin{align}
\text{INV}_2 &: \text{state}(n) = \text{BAIL\_PENDING} \implies \text{active\_bailout}(n) \land \text{transition}(\text{BAIL\_PENDING}, e_{\text{timeout}}) = \text{ESCALATING}. \tag{INV-2}
\end{align}

That is, any node in BAIL\_PENDING has an active bailout and the only permissible next transition is to ESCALATING, enforced by the Bailout Monitor without machine-actor discretion.

\begin{align}
\text{INV}_3 &: \text{state}(n) = \text{HUMAN\_ACKNOWLEDGED} \implies \text{prop\_depth}(n) = \text{distance}(n, h(n)). \tag{INV-3}
\end{align}

That is, when the human anchor is acknowledged, the propagation depth equals the distance from the node to its human anchor, confirming that the exception has reached the correct terminus and no intermediate machine actor has absorbed it.

% --------------------------------------------------------------
\subsection{Transition Conditions}\label{sec:transition-conditions}
% --------------------------------------------------------------

The six transitions of $\mathcal{B}$ are formally defined as follows. Each transition $T_i$ is a tuple $(q, \sigma, q', \theta_i)$ where $q$ is the source state, $\sigma$ is the input event, $q'$ is the destination state, and $\theta_i$ is the enabling condition that must hold in the authorization ledger before the transition may fire.

\begin{enumerate}\itemsep=0pt
\item[$T_1$ (IDLE $\rightarrow$ BOUNDED).] \emph{Trigger event.} Condition $\theta_1$: $\text{TRIGGER}(n, x)$ holds for some observed input $x$, as defined in \S~\ref{sec:bailout-trigger}. The \fact{} layer records the transition in the authorization ledger with a timestamp $t_{\text{trigger}}$ synchronized to a trusted time source. No machine actor may gate $T_1$; the trigger condition is evaluated by the \bounds{} layer and directly drives the state machine without an intervening decision step.

\item[$T_2$ (BOUNDED $\rightarrow$ HUMAN\_ACKNOWLEDGED).] \emph{Authorization confirmation.} Condition $\theta_2$: The \fact{} layer has approved a resolution $r$ produced by the Bounds Engine within $T_{\text{bounds}}$, and $r$ satisfies all authorization ledger constraints for the current node. The approval is recorded as a \texttt{resolution\_approved} ledger entry. The node returns to normal operation.

\item[$T_3$ (BOUNDED $\rightarrow$ BAIL\_PENDING).] \emph{Bounds timeout.} Condition $\theta_3$: $T_{\text{bounds}}$ has expired without a \fact-layer-approved resolution. The \fact{} layer pre-commit hook records the expiration and the node enters BAIL\_PENDING with active bailout flag set. A machine actor cannot suppress $T_3$ by extending $T_{\text{bounds}}$ unilaterally; any extension requires pre-authorization via a ledger entry completed before expiration.

\item[$T_4$ (BAIL\_PENDING $\rightarrow$ ESCALATING).] \emph{Atomic escalation.} Condition $\theta_4$: The node is in BAIL\_PENDING. Transition $T_4$ fires immediately and atomically — there is no dwell time in BAIL\_PENDING during which a machine actor could intervene. The \fact{} layer records the transition as an \texttt{escalation\_initiated} event with the full exception context and a cryptographically sealed reference to the trigger event from $T_1$. No machine actor may write to the transition log between $T_3$ and $T_4$.

\item[$T_5$ (ESCALATING $\rightarrow$ HUMAN\_ACKNOWLEDGED).] \emph{Human acknowledgment.} Condition $\theta_5$: The human anchor $h(n)$ has received the exception via \texttt{DeliverToHuman} and issued an acknowledgment \texttt{ACK}. The \fact{} layer verifies the acknowledgment signature and records it in the authorization ledger. The node transitions to HUMAN\_ACKNOWLEDGED and awaits a binding directive.

\item[$T_6$ (ESCALATING $\rightarrow$ IDLE).] \emph{Unexpected ACK during propagation.} Condition $\theta_6$: $h(n)$ receives the exception and responds with \texttt{ACK} before the escalation chain completes, meaning the exception reached $h(n)$ via an alternate pathway. This transition is accepted only if the \fact{} layer can verify that the ACK originated from $h(n)$ and that the exception context matches the pending propagation. The node returns to IDLE and the propagation chain is closed.

\item[$T_7$ (ESCALATING $\rightarrow$ RESOLVED).] \emph{Binding directive from human.} Condition $\theta_7$: $h(n)$ issues a directive $d \in \{\texttt{RESOLVE}, \texttt{REJECT}\}$ with full specification of the intended outcome. The \fact{} layer records $d$ verbatim in the authorization ledger and transitions the node to RESOLVED. No machine actor may modify, override, or substitute $d$.

\item[$T_8$ (HUMAN\_ACKNOWLEDGED $\rightarrow$ RESOLVED).] \emph{Directive from human after acknowledgment.} Condition $\theta_8$: The human principal issues a binding directive while in HUMAN\_ACKNOWLEDGED state. Identical in effect to $T_7$ but fires from a different source state, reflecting the two distinct paths by which a human may issue a terminal directive.

\item[$T_9$ (HUMAN\_ACKNOWLEDGED $\rightarrow$ IDLE).] \emph{ACK without binding directive.} Condition $\theta_9$: $h(n)$ acknowledges the exception and issues a \texttt{ACK} without a binding directive, signaling that the node should continue with its current parameters. The \fact{} layer records the ACK and the node returns to IDLE.
\end{enumerate}

\bigskip
\noindent\textbf{Transition enforcement.} All transitions are enforced by the \fact{} layer pre-commit hook. Before any state change is committed to the authorization ledger, the Bailout Monitor verifies that (a) the current state matches the transition's source state, (b) the input event matches the transition's trigger, and (c) the enabling condition $\theta_i$ is satisfied by ledger records. If any check fails, the transition is rejected and the machine remains in its current state. A rejected transition attempt is itself recorded as a protocol violation, ensuring that repeated failed transition attempts are detectable in the Forensic Ledger.

% --------------------------------------------------------------
\subsection{Bail-Out Trigger Condition}\label{sec:bailout-trigger}
% --------------------------------------------------------------

The bail-out trigger is the condition that causes a node to enter BOUNDED state (T1). Rather than requiring system designers to enumerate specific error codes in advance — a practice that provides coverage only for anticipated failure modes and leaves unanticipated ones invisible — the Bailout Protocol defines the trigger in terms of \emph{observational boundary violation}: a node escalates precisely when its observable input space contains a condition outside its provisioned operating range $R(n)$ and the node cannot produce a \fact-layer-approved resolution within $T_{\text{bounds}}$.

\bigskip
\noindent\textbf{Primary trigger condition.} The trigger $\text{TRIGGER}(n, x)$ is defined as a conjunction of three sub-conditions, all of which must hold simultaneously:

\begin{equation}
\text{TRIGGER}(n, x) \;\equiv\; \underbrace{x \notin R(n)}_{(i)} \;\land\; \underbrace{\neg\text{resolved}(n, x)}_{(ii)} \;\land\; \underbrace{\text{confidence}(n, x) < \tau}_{(iii)} \tag{TC}
\end{equation}

where:

\begin{itemize}\itemsep=0pt
\item[(i)] $x$ is an element of the node's observable input space that falls outside the provisioned operating range $R(n)$. This is the \emph{bound signal}: the node observes something it was not provisioned to handle autonomously.

\item[(ii)] $x$ has not been previously resolved by the node within the current operational session. This prevents re-trigger on known-exception inputs that have already received human adjudication.

\item[(iii)] $\text{confidence}(n, x) < \tau$ is the node's self-assessed confidence that it can produce a correct resolution for $x$ without human input, where $\tau$ is the provisioned confidence threshold. The confidence function $\text{confidence} : \text{node} \times \text{input} \rightarrow [0, 1]$ is defined at node provisioning and is \emph{not} adjustable at runtime by any machine actor. This prevents a machine actor from inflating its own confidence to suppress a warranted escalation.
\end{itemize}

\bigskip
\noindent\textbf{Trigger priority function.} When multiple trigger conditions fire simultaneously for node $n$, they are totally ordered by the priority function $\pi(n, x)$, which ensures deterministic transition selection:

\begin{equation}
\pi(n, x) \;=\; w_1 \cdot \text{severity}(x) \;+\; w_2 \cdot \bigl(1 - \text{confidence}(n, x)\bigr) \;+\; w_3 \cdot \text{distance}\bigl(n, h(n)\bigr) \tag{PF}
\end{equation}

where $\text{severity}(x) \in [0, 1]$ is the assigned severity of exception $x$ (higher values indicate greater potential impact), $\text{distance}(n, h(n))$ is the number of hops from $n$ to the human anchor in the parent chain, and $w_1, w_2, w_3$ are provisioned constants satisfying $w_1 > w_2 > w_3 \geq 0$. The highest-priority trigger fires first. Ordering is enforced by the Bailout Monitor before any state transition is committed to the Forensic Ledger.

\bigskip
\noindent\textbf{Secondary trigger: explicit \texttt{bailout\_signal}.} In addition to TC, the Bounds Engine may emit an explicit \texttt{bailout\_signal} at any time, regardless of confidence scores. This handles the case where the Bounds Engine detects a condition — for example, a logical contradiction between two inputs within $R(n)$ — that it can recognize as unresolvable even though both inputs are individually within the operating range. The \texttt{bailout\_signal} bypasses TC entirely and immediately initiates T3:

\begin{equation}
\texttt{bailout\_signal}(x) \;\implies\; \text{T3 fires for node } n \text{ with exception } x. \tag{TC-secondary}
\end{equation}

\bigskip
\noindent\textbf{The no-discretion property.} The \bounds{} layer does not exercise discretion over the trigger condition. Specifically:

\begin{enumerate}
    \item The evaluation of $\text{TRIGGER}(n, x)$ is a read-only query against the operating range definition $R(n)$, not a judgment call made by a machine actor.
    \item The result of the query directly drives the state machine transition without an intervening decision step. There is no machine-actor gate between trigger evaluation and protocol activation.
    \item The confidence function is provisioned at node initialization and is immutable at runtime. No machine actor may adjust $\tau$ to suppress escalation.
\end{enumerate}

These constraints ensure that the trigger condition is architecturally enforced rather than operationally optional.

% --------------------------------------------------------------
\subsection{Escalation Algorithm}\label{sec:escalation-algorithm}
% --------------------------------------------------------------

When the state machine enters ESCALATING, the Bailout Protocol executes the escalation algorithm $\text{ESCALATE}(n, e, \mathcal{C})$ to propagate the exception from node $n$ to the human anchor $h(n)$ along the immutable parent chain $C(n) = (n = v_0, v_1, \ldots, v_k = h(n))$, where $v_{i+1} = \alpha(v_i)$ for each $i$.

\bigskip
\noindent\textbf{Algorithm.}
\begin{algorithmic}
\Function{ESCALATE}{$n, e, \mathcal{C}$}
    \State $p \gets \alpha(n)$ \Comment{parent node}
    \State $e_{\text{ctx}} \gets \text{BuildContext}(n, e, \mathcal{C})$
    \State $\text{send\_count} \gets 0$
    \State $T_{\text{current}} \gets T_{\text{prop}}$ \Comment{initial propagation timeout}
    \While{$\text{send\_count} < N_{\max}$}
        \State $\text{path} \gets \text{SelectPathway}(p, \mathcal{C})$
        \State $\text{ack\_received} \gets \text{SendToParent}(p, e_{\text{ctx}}, \text{path}, T_{\text{current}})$
        \If{$\text{ack\_received} = \texttt{TRUE}$}
            \State \Return $\texttt{ESCALATION\_DELIVERED}$
        \EndIf
        \State $\text{send\_count} \gets \text{send\_count} + 1$
        \State $T_{\text{current}} \gets \min(2 \cdot T_{\text{current}},\; T_{\text{cap}})$ \Comment{exponential backoff}
    \EndWhile
    \State \Comment{All retries exhausted}
    \State $\text{FireTimeout}(\texttt{parent\_unreachable})$
    \State \Return $\texttt{PATHWAY\_EXHAUSTED}$
\EndFunction
\end{algorithmic}

\bigskip
\noindent\textbf{Context building.} $\text{BuildContext}(n, e, \mathcal{C})$ constructs the diagnostic payload propagated alongside the exception. It includes: (i) the full exception state $e$ including the triggering input $x$ and the observed confidence value, (ii) the state machine history $H(n)$ recording all prior transitions and their timestamps, (iii) the operating range $R(n)$ as provisioned at initialization, (iv) the authorization ledger excerpt for the current operational session, (v) the propagation chain identifier $\mathcal{C}$, and (vi) a hash reference to the trigger event recorded at T1. The context is cryptographically sealed by the \fact{} layer before transmission; intermediate nodes append their own diagnostic layer but cannot modify the original exception record.

\bigskip
\noindent\textbf{Pathway selection.} $\text{SelectPathway}(p, \mathcal{C})$ queries the Pathway Health Registry (PHR) and returns the functional pathway to parent $p$ with the lowest estimated latency among currently available pathways. The PHR maintains real-time availability and latency data for all provisioned communication channels between nodes. Pathways marked \texttt{DEGRADED} are excluded from selection until a health check succeeds.

\bigskip
\noindent\textbf{Termination conditions.} The algorithm terminates in one of three states:

\begin{itemize}\itemsep=0pt
\item[\texttt{ESCALATION\_DELIVERED}:] The parent node acknowledged receipt within the timeout window. The exception has been successfully delivered to $p$ and the propagation chain continues toward $h(n)$. The algorithm returns control to the state machine, which transitions to the appropriate state based on the acknowledgment content.

\item[\texttt{PATHWAY\_EXHAUSTED}:] All $N_{\max}$ retry attempts at $T_{\text{cap}}$ have failed without acknowledgment. A \texttt{parent\_unreachable} timeout event is fired and the state machine enters RESOLVED with the \texttt{PATHWAY\_EXHAUSTED} tag. The exception is logged with full diagnostic context; human remediation is required to clear the state. No machine actor may suppress or absorb this terminal resolution.

\item[\texttt{RECURSION\_LIMIT}:] The propagation chain has traversed $D_{\max}$ hops without reaching $h(n)$. This condition indicates a structural failure in the parent chain and is treated as a protocol violation. The Bailout Monitor records the violation in the Forensic Ledger and fires a \texttt{chain\_broken} alert to the system administrator.
\end{itemize}

\begin{table}[htbp]
\caption{Bailout Protocol escalation algorithm.}
\label{tab:escalation-algorithm}
\begin{tabular}{@{}ll@{}}
\toprule
Symbol & Definition \\
\midrule
$n$ & Originating node \\
$e$ & Exception state \\
$\mathcal{C}$ & Propagation chain identifier \\
$\alpha(n)$ & Immutable parent pointer of $n$ \\
$C(n)$ & Parent chain from $n$ to $h(n)$ \\
$N_{\max}$ & Maximum retry count (default: 5) \\
$T_{\text{prop}}$ & Per-hop propagation timeout (initial: 5s) \\
$T_{\text{cap}}$ & Maximum propagation timeout (cap: 60s) \\
$D_{\max}$ & Maximum propagation depth (hardware-defined) \\
\bottomrule
\end{tabular}
\end{table}

% --------------------------------------------------------------
\subsection{Bypass Mechanism}\label{sec:bypass-mechanism}
% --------------------------------------------------------------

The bypass mechanism is the structural property that distinguishes the Bailout Protocol from conventional escalation hierarchies. It ensures that no machine actor — regardless of its position in the parent chain, its capability level, or its access to the \fact{} layer — can absorb, resolve, or suppress an exception that has entered the ESCALATING state.

\bigskip
\noindent\textbf{Why bypasses all machine actors.} The parent-chain bypass is the mechanism by which the Bailout Protocol skips every intermediate machine actor in the propagation chain $C(n)$, delivering the exception directly to $h(n)$. This bypass is not an optimization — it is a structural necessity that follows from the upstream accountability guarantee. Any machine actor $m$ that could absorb an exception in ESCALATING state would become a terminal node for that exception, and the accountability chain would terminate at $m$ rather than at a human. If $m$ then fails, or if $m$'s resolution is incorrect or insufficient, no human principal is present in the chain to detect or correct the failure. The system compounds failure instead of surfacing it.

The bypass eliminates three mutually independent failure modes — Intercept (machine actor absorbs the exception), Redirect (machine actor reroutes to the wrong human), and Delay (machine actor buffers the exception beyond $T_{\max}$) — by architectural constraint: there is no machine actor in the communication path, so none of these failures are architecturally possible. This is analogous to a hardware interrupt line hardwired to a designated handler: just as a hardware interrupt cannot be intercepted or redirected by user-space code, the bailout escalation cannot be intercepted or redirected by any agent-level component.

\bigskip
\noindent\textbf{Enforcement.} The bypass is enforced by two cooperating mechanisms operating at different layers:

\begin{enumerate}
\item[\textbf{Fact Layer gate.}] The \fact{} layer pre-commit hook rejects any ledger entry for a resolution where the underlying trigger satisfies TC and $T_{\text{bounds}}$ has expired. This prevents the Bounds Engine from issuing a retroactive resolution after timeout to suppress warranted escalation. The \fact{} layer gate operates identically on all machine actors: there is no privileged machine actor with the ability to bypass this check.

\item[\textbf{Immutable parent-pointer and human-root architecture.}] The parent pointer $\alpha(n)$ and the human-root identifier $h(n)$ are stored in \fact-layer-managed memory and are read-only at the machine-actor level. A machine actor cannot redirect the parent chain, cannot insert itself as an intermediate resolver, and cannot modify the $humanRootId$ field that terminates the chain. Any attempted modification requires \purpose{} layer authorization with a corresponding ledger entry — which cannot be completed without human approval if the modification would affect the bailout terminus.
\end{enumerate}

The bypass property is formally stated:

\begin{align}
\forall n \in \text{nodes},\; \forall e \in \text{exceptions}:\; \text{state}(n) = \text{ESCALATING} \implies \text{direct\_path}(n, h(n), e). \tag{BP}
\end{align}

That is, for any node and any exception in ESCALATING state, there exists a direct communication path from $n$ to $h(n)$ for that exception that involves no intermediate machine actor as a terminus.

% --------------------------------------------------------------
\subsection{Timeout Handling}\label{sec:timeout-handling}
% --------------------------------------------------------------

Timeouts in the Bailout Protocol serve a liveness function: they prevent the protocol from permanently stalling when a parent node is unreachable, a communication pathway is unresponsive, or the human anchor is temporarily unavailable. Each timeout has a defined scope, a defined scope-exit action, and an escalation path that prevents deadlock.

\bigskip
\noindent\textbf{Timeout definitions.}

\begin{enumerate}
\item[\textbf{$T_{\text{bounds}}$ — Bounds Engine resolution timeout.}] The maximum time a node may remain in BOUNDED state attempting autonomous resolution. Default: 60 seconds, provisioned per node class according to the expected latency of the node's resolution procedure. If $T_{\text{bounds}}$ expires without a \fact-layer-approved resolution, the node transitions to BAIL\_PENDING $\rightarrow$ ESCALATING via T3 and T4. The \fact{} layer cannot extend $T_{\text{bounds}}$ unilaterally; any extension requires human authorization via a separate ledger entry recorded before $T_{\text{bounds}}$ expires.

\item[\textbf{$T_{\text{prop}}$ — Per-hop propagation timeout.}] The maximum time node $n$ waits for an acknowledgment from its parent after sending an exception via \texttt{SendToParent}. Default initial value: 5 seconds. If $T_{\text{prop}}$ expires without ACK, the algorithm retries with exponential backoff: $T_{\text{prop}} \leftarrow \min(2 \cdot T_{\text{prop}},\; T_{\text{cap}})$. This backoff continues until either an ACK is received or $N_{\max}$ retries have been exhausted at $T_{\text{cap}}$.

\item[\textbf{$T_{\text{cap}}$ — Maximum propagation timeout value.}] The ceiling value for the exponential backoff of $T_{\text{prop}}$. Default: 60 seconds. After $N_{\max}$ consecutive retries at $T_{\text{cap}}$ without ACK, the algorithm fires a \texttt{parent\_unreachable} timeout event, which propagates to $h(n)$ via the next available functional pathway tracked by the Pathway Health Registry.

\item[\textbf{$T_{\text{human}}$ — Human acknowledgment timeout.}] The maximum time the human anchor may remain in HUMAN\_ACKNOWLEDGED state without issuing a directive. Default: 300 seconds (5 minutes), configurable per deployment based on the expected human response time for the node's operational domain. If $T_{\text{human}}$ expires, the protocol issues a reminder notification to $h(n)$ and transitions the originating node to RESOLVED with an \texttt{unresolved} tag. The node becomes quiescent: no autonomous retry occurs, no further actions are attempted, and the \texttt{unresolved} tag in the Forensic Ledger signals that human re-engagement is required to clear the state. The protocol does not re-escalate automatically; automatic re-escalation would create a potential denial-of-service condition against the human anchor.
\end{enumerate}

\bigskip
\noindent\textbf{Pathway Health Registry.} The Pathway Health Registry (PHR) is a runtime data structure maintained by the Bailout Monitor that tracks the availability, latency, and error rate of each provisioned communication pathway to the human anchor. Before each \texttt{SelectPathway} call, the registry is queried and the pathway with the lowest estimated latency among currently functional pathways is selected. Pathways that fail $K$ consecutive health checks (default $K = 3$) are marked \texttt{DEGRADED} and excluded from pathway selection until a health check succeeds. A background health-ping thread updates the PHR at intervals of $T_{\text{health}}$ (default: 30 seconds), operating with the same scheduling priority as the Bailout Monitor to ensure that pathway status remains current even under heavy agent workload.

\bigskip
\noindent\textbf{Liveness guarantee.} Together, $T_{\text{bounds}}$, $T_{\text{prop}}$, $T_{\text{cap}}$, and $T_{\text{human}}$ define a finite upper bound on the wall-clock time from trigger firing (T1) to human acknowledgment under any combination of parent unavailability and pathway degradation:

\begin{equation}
T_{\max} \;=\; T_{\text{bounds}} \;+\; N_{\max} \cdot T_{\text{cap}} \;+\; T_{\text{human}} \tag{T-max}
\end{equation}

A deployment that cannot guarantee human acknowledgment within the required $T_{\max}$ under all failure scenarios must provision additional or higher-bandwidth pathways, or must reduce $T_{\text{prop}}$ and $T_{\text{cap}}$ to bring the worst-case bound within the required limit. This analysis is a required part of deployment certification for any \bxthree{} system operating in a regulated or safety-critical environment.

\bigskip
\noindent\textbf{Timeout interaction with the bypass mechanism.} The bypass mechanism ensures that timeouts cannot be weaponized by a machine actor to delay or suppress escalation. Specifically: (a) $T_{\text{bounds}}$ is enforced by the \fact{} layer pre-commit hook, not by the \bounds{} layer; a machine actor cannot artificially extend $T_{\text{bounds}}$ to prevent the transition to ESCALATING. (b) $T_{\text{prop}}$ backoff is calculated by the Bailout Monitor, not by any machine actor; the backoff curve is defined in the protocol specification and cannot be altered at runtime. (c) If all pathways timeout at $T_{\text{cap}}$, the protocol resolves with \texttt{PATHWAY\_EXHAUSTED} — this is a terminal RESOLVED state, not a suppression of escalation. The \texttt{PATHWAY\_EXHAUSTED} tag in the Forensic Ledger ensures that the failure of the pathway is recorded as an incident requiring human remediation, not as a silent resolution by a machine actor.