\section{The Bailout Protocol}
\label{sec:bailout}

The deployment of autonomous multi-agent systems into consequential environments — agricultural resource management, clinical decision support, financial clearing, and critical logistics — has exposed a structural gap that policy alone cannot close. Conventional architectures can describe what agents do; they cannot architecturally guarantee that every agent's authority traces to a human principal. As agent populations spawn recursively, modify state, and execute decisions with real-world consequences, the question of who authorized each action, through what chain of delegation, and with what scope constraints has become the central unsolved problem in agentic infrastructure.

We introduce the \textbf{Bailout Protocol}: the mandatory, architecturally enforced escalation mechanism of the \bxthree{} Framework. It compels every unresolved exception state in any \bxthree{} node to propagate upward to a named human principal, bypassing all intermediate machine actors, with full forensic attribution at every step. It is not a feature added to an existing architecture. It is the mechanism by which the \bxthree{} Framework makes its upstream accountability guarantee structurally operative at runtime.

The protocol operates against the backdrop of the \bxthree{} three-layer model. The \purpose{} layer carries intent, judgment, and final authority; it defines the operating range $R(n)$ for each node $n$ and receives all escalated exceptions. The \bounds{} layer carries bounded reasoning and constrained execution within $R(n)$. When it encounters a condition outside $R(n)$ it cannot resolve, it triggers the Bailout Protocol — compulsorily and non-discretionarily. It cannot absorb, reframe, or redirect the exception once triggered. The \fact{} layer carries deterministic enforcement and forensic auditability; it enforces the transition relation of the state machine and maintains the append-only Forensic Ledger as an immutable record of every protocol event.

When the \bounds{} layer encounters a condition outside $R(n)$ it cannot resolve, the functional separation itself dictates that the decision must return to the \purpose{} layer. There is no other layer with the authority to adjudicate it. The Bailout Protocol is not a feature added to this separation; it is the mechanism by which the separation enforces its own accountability guarantee. The human anchor $h(n)$ is not a design choice — it is a structural necessity.

This distinguishes \bxthree{} from prior escalation architectures. The Tiered Agentic Oversight model~\cite{kim2025tao} permits absorption at intermediate machine tiers, making human oversight a high-tier option rather than a compulsory terminus. The Bailout Protocol forbids this absorption by architectural constraint: the distinction is the difference between a system that \emph{may} consult a human and a system that \emph{cannot proceed} without one.

% ---------------------------------------------------------------
\subsection{State Machine Formalism}
\label{sec:state-machine}

The Bailout Protocol is formally modeled as a deterministic finite state machine
\begin{equation}
\mathcal{B} \;=\; (Q,\; q_0,\; \Sigma,\; \rightarrow,\; Q_F)
\end{equation}
where:

\begin{itemize}\itemsep=0pt
\item[$Q$] $\{\;\text{IDLE},\; \text{BOUNDED},\; \text{BAIL\_PENDING},\; \text{ESCALATING},\; \text{HUMAN\_ACKNOWLEDGED},\; \text{RESOLVED}\; \}$ is the finite set of protocol states.
\item[$q_0$] $\text{IDLE}$ is the designated initial state, entered at node startup and after every successful directive resolution.
\item[$\Sigma$] The input alphabet comprising trigger events $e_{\text{trigger}}$, authorization confirmations $e_{\text{auth}}$, timeout signals $e_{\text{timeout}}$, human acknowledgments $e_{\text{ack}}$, binding resolution directives $e_{\text{resolve}}$, and rejection directives $e_{\text{reject}}$.
\item[$\rightarrow$] $\;\subseteq Q \times \Sigma \times Q$ is the deterministic transition relation (Table~\ref{tab:bailout-transition}).
\item[$Q_F$] $\{\text{RESOLVED}\}$ is the set of terminal accepting states.
\end{itemize}

A node's state $q \in Q$ is stored in the node's Fact Layer worksheet and is observable by its parent node and by the Bailout Monitor. State updates are atomic: no intermediate or partially-executing states are observable externally. The machine is \emph{deterministic}: for every $(q, \sigma) \in Q \times \Sigma$, at most one $q' \in Q$ satisfies $(q, \sigma, q') \in \rightarrow$. Determinism is enforced by the Bailout Monitor's priority function $\pi$, which totally orders simultaneous trigger conditions and selects the highest-priority enabled transition before committing any state change. If no transition is defined for a $(q, \sigma)$ pair, the machine stalls and the Bailout Monitor records a protocol violation in the Forensic Ledger.

\begin{table}[htbp]
\caption{Bailout Protocol state transition relation $\rightarrow \;\subseteq Q \times \Sigma \times Q$. Rows are current state; columns are input events. Entries denote the next state. Blank cells indicate that no transition is defined — the machine stalls and a protocol violation is recorded.}
\label{tab:bailout-transition}
\begin{tabular}{@{}lcccccc@{}}
\toprule
 & \multicolumn{6}{c}{Input event} \\
\cmidrule(l){2-7}
Current state & $e_{\text{trigger}}$ & $e_{\text{auth}}$ & $e_{\text{timeout}}$ & $e_{\text{ack}}$ & $e_{\text{resolve}}$ & $e_{\text{reject}}$ \\
\midrule
IDLE              & BOUNDED            & --- & --- & --- & --- & --- \\
BOUNDED           & --- & HUMAN\_ACKNOWLEDGED & BAIL\_PENDING & --- & --- & --- \\
BAIL\_PENDING     & --- & --- & ESCALATING & --- & --- & --- \\
ESCALATING        & --- & HUMAN\_ACKNOWLEDGED & ESCALATING & IDLE & RESOLVED & RESOLVED \\
HUMAN\_ACKNOWLEDGED & --- & --- & --- & IDLE & RESOLVED & IDLE \\
RESOLVED          & --- & --- & --- & --- & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

\bigskip
\noindent\textbf{State definitions.}

\begin{itemize}\itemsep=0pt
\item[\textbf{IDLE.}] The node is operating within its provisioned operating range $R(n)$. No exception condition is active. The \fact{} layer dispatches normal action requests through the standard execution path. The node may transition to BOUNDED upon receipt of a trigger event satisfying the bail-out trigger condition (TC), defined in \S~\ref{sec:bailout-trigger}.

\item[\textbf{BOUNDED.}] An exception $e$ has been detected by the \bounds{} layer and is being held at the \bounds{}--\fact{} layer boundary. A time-bounded resolution attempt is in progress. All resolution paths within current authority remain open. If the Bounds Engine produces a resolution $r$ that the \fact{} layer approves before $T_{\text{bounds}}$ expires, the node returns to IDLE via T2. If all resolution paths are exhausted or $T_{\text{bounds}}$ expires, the state advances to BAIL\_PENDING via T3.

\item[\textbf{BAIL\_PENDING.}] The node has determined — either through an explicit \texttt{bailout\_signal} or through $T_{\text{bounds}}$ expiration without a \fact-layer-approved resolution — that the exception exceeds its current authority. The Bailout Protocol is formally activated, but propagation to $h(n)$ has not yet been initiated. This is the last architectural point at which a machine actor could theoretically suppress escalation. It cannot: entry to BAIL\_PENDING triggers an immediate atomic transition to ESCALATING via $e_{\text{timeout}}$ (T4), with no dwell time and no machine-actor intervention permitted.

\item[\textbf{ESCALATING.}] The exception is propagating upward toward $h(n)$. Each intermediate node receives the exception, appends its diagnostic context to the propagation chain (detailed in \S~\ref{sec:escalation-algorithm}), and forwards the exception via \texttt{SendToParent} without attempting local resolution. The state remains ESCALATING until either (a) the human anchor acknowledges receipt ($e_{\text{ack}}$: T5), (b) the human anchor issues a binding directive ($e_{\text{resolve}}$ or $e_{\text{reject}}$: T7), or (c) all retry pathways time out after $N_{\max}$ retries at $T_{\text{cap}}$, propagating a \texttt{parent\_unreachable} event to $h(n)$ via an alternate functional pathway.

\item[\textbf{HUMAN\_ACKNOWLEDGED.}] The human anchor has received the exception, observed the diagnostic context, and acknowledged it. The node awaits a directive. No machine actor may issue, simulate, or substitute a directive in this state. The node remains until the human principal issues \texttt{ACK} (return to IDLE via T8) or a binding directive \texttt{RESOLVE} / \texttt{REJECT} (enter RESOLVED via T9).

\item[\textbf{RESOLVED.}] The human principal has issued a directive and the protocol has recorded it in the authorization ledger. The directive is one of: \texttt{ACK} (continue with current parameters), \texttt{RESOLVE} (execute a specific binding resolution), or \texttt{REJECT} (disallow the proposed action and enter a quiescent error state). This state is terminal for the current protocol run. The node either returns to IDLE (on \texttt{ACK}) or becomes quiescent (on \texttt{REJECT}), awaiting human re-engagement.
\end{itemize}

\bigskip
\noindent\textbf{State invariants.} The following invariants hold for all reachable states of $\mathcal{B}$:

\begin{align}
\text{INV}_1 &: \forall n \in \text{nodes},\; \text{state}(n) \in Q \setminus \{\text{ESCALATING}\} \iff \neg\text{active\_bailout}(n). \tag{INV-1}
\end{align}

That is, a node occupies a non-escalating state \emph{iff} it has no active bailout propagation in flight. A node may carry a bailout history (a ledger entry recording a past propagation) while in IDLE, but no active propagation may be underway.

\begin{align}
\text{INV}_2 &: \text{state}(n) = \text{BAIL\_PENDING} \implies \text{active\_bailout}(n) \land \text{transition}(\text{BAIL\_PENDING}, e_{\text{timeout}}) = \text{ESCALATING}. \tag{INV-2}
\end{align}

That is, any node in BAIL\_PENDING has an active bailout and the only permissible next transition is to ESCALATING, enforced by the Bailout Monitor without machine-actor discretion.

\begin{align}
\text{INV}_3 &: \text{state}(n) = \text{ESCALATING} \implies \text{active\_bailout}(n) \land \text{current\_destination}(n) = h(n). \tag{INV-3}
\end{align}

That is, a node in ESCALATING carries an active bailout whose destination is always and only the human anchor $h(n)$. No machine actor may be the terminus of an active bailout.

% ---------------------------------------------------------------
\subsection{Transition Conditions}
\label{sec:transitions}

Each transition in Table~\ref{tab:bailout-transition} is gated by a precise condition evaluated by the Bailout Monitor. Together, the six transitions T1–T7 cover the complete lifecycle of a protocol run:

\begin{enumerate}\itemsep=0pt
\item[\textbf{T1 — IDLE $\rightarrow$ BOUNDED (trigger):}] The node receives a trigger event $e_{\text{trigger}}$ satisfying the primary trigger condition TC (defined in \S~\ref{sec:bailout-trigger}) or an explicit \texttt{bailout\_signal}. The Bailout Monitor verifies TC satisfaction before committing the state change to the Forensic Ledger. The transition is atomic: no intermediate state is observable.

\item[\textbf{T2 — BOUNDED $\rightarrow$ IDLE (resolution):}] Fires when the Bounds Engine produces a resolution $a$ that the Fact Layer approves before $T_{\text{bounds}}$ expires. Formally:
\begin{equation}
\text{T2 fires} \;\iff\; \exists a : \text{FactLayer.approve}(a) \;\land\; \text{clock} < T_{\text{bounds}}. \tag{T2}
\end{equation}
The resolution is recorded in the authorization ledger as aFact-Layer-approved action with full provenance attribution. The node returns to normal execution with no memory of the exception in its active state.

\item[\textbf{T3 — BOUNDED $\rightarrow$ BAIL\_PENDING (escalation):}] Fires when $T_{\text{bounds}}$ expires without aFact-Layer-approved resolution, or when the Bounds Engine emits an explicit \texttt{bailout\_signal} declaring that exception $x$ exceeds its current authority. Formally:
\begin{equation}
\text{T3 fires} \;\iff\; \bigl(\text{clock} \geq T_{\text{bounds}} \;\land\; \neg\exists a : \text{FactLayer.approve}(a)\bigr) \;\lor\; \texttt{bailout\_signal}(x). \tag{T3}
\end{equation}

\item[\textbf{T4 — BAIL\_PENDING $\rightarrow$ ESCALATING (activation):}] Entry to BAIL\_PENDING triggers an immediate atomic transition to ESCALATING. There is no dwell time in BAIL\_PENDING; it is a zero-duration state that serves only as a label indicating that bailout has been formally activated. T4 fires synchronously with the state entry, enforced by the Bailout Monitor without machine-actor gating. This transition activates the escalation algorithm defined in \S~\ref{sec:escalation-algorithm}.

\item[\textbf{T5 — ESCALATING $\rightarrow$ HUMAN\_ACKNOWLEDGED (contact):}] Fires when the exception reaches $h(n)$ and the human anchor confirms receipt via $e_{\text{ack}}$. The Bailout Monitor records the contact timestamp in the Forensic Ledger. The node awaits a directive; no machine actor may issue one.

\item[\textbf{T7 — ESCALATING $\rightarrow$ RESOLVED (directive):}] Fires when the human anchor issues a binding directive $e_{\text{resolve}}$ (execute a specific resolution) or $e_{\text{reject}}$ (disallow the proposed action). Formally:
\begin{equation}
\text{T7 fires} \;\iff\; e_{\text{resolve}} \;\lor\; e_{\text{reject}}. \tag{T7}
\end{equation}
The directive is recorded in the authorization ledger as the terminal event of the protocol run. If $e_{\text{reject}}$, the node enters a quiescent error state and awaits human re-engagement. If $e_{\text{resolve}}$, the node returns to IDLE after executing the prescribed resolution.
\end{enumerate}

% ---------------------------------------------------------------
\subsection{Bail-Out Trigger Condition}
\label{sec:bailout-trigger}

The bail-out trigger is the condition that causes a node to enter BOUNDED state (T1). Rather than requiring system designers to enumerate specific error codes in advance — a practice that provides coverage only for anticipated failure modes and leaves unanticipated ones invisible — the Bailout Protocol defines the trigger in terms of \emph{observational boundary violation}: a node escalates precisely when its observable input space contains a condition outside its provisioned operating range $R(n)$, and the node cannot produce a Fact-Layer-approved resolution within $T_{\text{bounds}}$.

\bigskip
\noindent\textbf{Primary trigger condition.} The trigger $\text{TRIGGER}(n, x)$ is defined as a conjunction of three sub-conditions, all of which must hold simultaneously:
\begin{equation}
\text{TRIGGER}(n, x) \;\equiv\; \underbrace{x \notin R(n)}_{(i)} \;\land\; \underbrace{\neg\text{resolved}(n, x)}_{(ii)} \;\land\; \underbrace{\text{confidence}(n, x) < \tau}_{(iii)} \tag{TC}
\end{equation}
where:

\begin{itemize}\itemsep=0pt
\item[(i)] $x$ is an element of the node's observable input space that falls outside the provisioned operating range $R(n)$. This is the \emph{bound signal} condition: the node observes something it was not provisioned to handle autonomously.

\item[(ii)] $x$ has not been previously resolved by the node within the current operational session. This prevents re-trigger on known-exception inputs that have already received human adjudication.

\item[(iii)] $\text{confidence}(n, x) < \tau$ is the node's self-assessed confidence that it can produce a correct resolution for $x$ without human input, where $\tau$ is the provisioned confidence threshold. The confidence function $\text{confidence} : \text{node} \times \text{input} \rightarrow [0, 1]$ is defined at node provisioning and is \emph{not} adjustable at runtime by any machine actor. This prevents a machine actor from inflating its own confidence to suppress a warranted escalation.
\end{itemize}

\bigskip
\noindent\textbf{Trigger priority function.} When multiple trigger conditions fire simultaneously for node $n$, they are totally ordered by the priority function $\pi(n, x)$, which ensures deterministic transition selection:
\begin{equation}
\pi(n, x) \;=\; w_1 \cdot \text{severity}(x) \;+\; w_2 \cdot \bigl(1 - \text{confidence}(n, x)\bigr) \;+\; w_3 \cdot \text{distance}\bigl(n, h(n)\bigr) \tag{PF}
\end{equation}

where $\text{severity}(x) \in [0, 1]$ is the assigned severity of exception $x$ (higher values indicate greater potential impact), $\text{distance}(n, h(n))$ is the number of hops from $n$ to the human anchor in the parent chain (longer distances increase priority because propagation latency is greater), and $w_1, w_2, w_3$ are provisioned constants satisfying $w_1 > w_2 > w_3 \geq 0$. The highest-priority trigger fires first. Ordering is enforced by the Bailout Monitor before any state transition is committed to the Forensic Ledger.

\bigskip
\noindent\textbf{Secondary trigger: explicit \texttt{bailout\_signal}.} In addition to TC, the Bounds Engine may emit an explicit \texttt{bailout\_signal} at any time, regardless of confidence scores. This handles the case where the Bounds Engine detects a condition — for example, a logical contradiction between two inputs within $R(n)$ — that it can recognize as unresolvable even though both inputs are individually within the operating range. The \texttt{bailout\_signal} bypasses TC entirely and immediately initiates T3:
\begin{equation}
\texttt{bailout\_signal}(x) \;\implies\; \text{T3 fires for node } n \text{ with exception } x. \tag{TC-secondary}
\end{equation}

\bigskip
\noindent\textbf{The no-discretion property.} The \bounds{} layer does not exercise discretion over the trigger condition. Specifically:

\begin{enumerate}
    \item The evaluation of $\text{TRIGGER}(n, x)$ is a read-only query against the operating range definition $R(n)$, not a judgment call made by a machine actor.
    \item The result of the query directly drives the state machine transition without an intervening decision step. There is no machine-actor gate between trigger evaluation and protocol activation.
    \item The confidence function is provisioned at node initialization and is immutable at runtime. No machine actor may adjust $\tau$ to suppress escalation.
\end{enumerate}

These constraints ensure that the trigger condition is architecturally enforced rather than operationally optional.

% ---------------------------------------------------------------
\subsection{Escalation Algorithm}
\label{sec:escalation-algorithm}

When the state machine enters ESCALATING (T4), the protocol executes the \texttt{Escalate} algorithm to propagate the exception to $h(n)$ along the parent chain. The algorithm is executed by the Bailout Monitor on the originating node; intermediate nodes execute only \texttt{ForwardException}, which appends diagnostic context and relays the exception upward.

\begin{algorithm}[htbp]
\caption{Bailout Protocol escalation algorithm.}
\label{alg:escalate}
\begin{algorithmic}[1]
\Procedure{Escalate}{$n$, trigger}
  \State $\text{ctx} \leftarrow$ \Call{BuildDiagnosticContext}{$n$, trigger}
  \State $p \leftarrow \alpha(n)$ \Comment{parent node}
  \State $\text{path} \leftarrow$ \Call{SelectPathway}{PHR}
  \State $r \leftarrow 0$ \Comment{retry counter}
  \While{$\text{true}$}
    \State $\text{sent} \leftarrow$ \Call{SendToParent}{$p$, ctx, path}
    \If{$\text{sent.acknowledged}$}
      \State \Return \texttt{ESCALATING\_CONFIRMED}
    \EndIf
    \State $r \leftarrow r + 1$
    \If{$r \geq N_{\max}$}
      \State \Call{RecordPathwayExhausted}{$n$, ctx}
      \State \Call{PropagateToHumanRoot}{$n$, ctx, trigger}
      \State \Return \texttt{PATHWAY\_EXHAUSTED}
    \EndIf
    \State $T_{\text{prop}} \leftarrow \min(2 \cdot T_{\text{prop}},\; T_{\text{cap}})$
    \State \Call{Wait}{$T_{\text{prop}}$}
    \State $\text{path} \leftarrow$ \Call{SelectPathway}{PHR} \Comment{re-evaluate after backoff}
  \EndWhile
\EndProcedure
\Statex
\Procedure{ForwardException}{$n$, ctx}
  \State $\text{ctx} \leftarrow \text{ctx} \;\cup\;$ \Call{AppendNodeDiagnostics}{$n$}
  \State $p \leftarrow \alpha(n)$
  \State \Call{SendToParent}{$p$, ctx, \Call{SelectPathway}{PHR}}
\EndProcedure
\Statex
\Procedure{SelectPathway}{PHR}
  \State $\text{functional} \leftarrow \{p \in \text{PHR.pathways} \mid p.\text{status} \neq \texttt{DEGRADED}\}$
  \State \Return $\arg\min_{p \in \text{functional}} p.\text{latency\_estimate}$
\EndProcedure
\Statex
\Procedure{BuildDiagnosticContext}{$n$, trigger}
  \State \Return $\{$
    $\text{node\_id}: n.\text{id},\;$
    $\text{timestamp}: \text{clock},\;$
    $\text{trigger}: \text.trigger},\;$
    $\text{operating\_range}: R(n),\;$
    $\text{bailout\_history}:$ \Call{GetBailoutHistory}{$n$}$,\;$
    $\text{parent}: \alpha(n),\;$
    $\text{human\_anchor}: h(n)$
  $\}$
\EndProcedure
\end{algorithmic}
\end{algorithm}

\bigskip
\noindent\textbf{Correctness properties of the escalation algorithm.}

\begin{enumerate}\itemsep=0pt
\item[\textbf{P1 (Path integrity).}] The exception always propagates along the parent chain from the originating node toward $h(n)$. No node may select an alternate destination for an active bailout. This follows from the Fact Layer gate: any \texttt{SendToParent} call with a destination field not equal to $\alpha(n)$ is rejected by the Bailout Monitor's pre-commit hook.

\item[\textbf{P2 (No local absorption).}] No intermediate node may attempt to resolve the exception locally during propagation. The \texttt{ForwardException} procedure appends diagnostic context and immediately forwards; it does not invoke the Bounds Engine or any resolution procedure. This is enforced by the Bailout Monitor: any attempt to invoke resolution during ESCALATING state is recorded as a protocol violation.

\item[\textbf{P3 (Deterministic pathway selection).}] \texttt{SelectPathway} is a pure function of the current PHR state. Given identical PHR states, it returns identical pathway selections. This ensures reproducibility of protocol runs for forensic auditing.

\item[\textbf{P4 (Exponential backoff with cap).}] The backoff formula $T_{\text{prop}} \leftarrow \min(2 \cdot T_{\text{prop}},\; T_{\text{cap}})$ ensures that retry intervals grow geometrically up to the cap, preventing the protocol from sending at high frequency to a failed parent while still allowing eventual recovery when the parent becomes reachable.
\end{enumerate}

% ---------------------------------------------------------------
\subsection{Bypass Mechanism}
\label{sec:bypass}

The bypass mechanism is the structural property that distinguishes the Bailout Protocol from conventional escalation hierarchies. It ensures that no machine actor — regardless of its position in the parent chain, its capability level, or its access to the Fact Layer — can absorb, resolve, or suppress an exception that has entered the ESCALATING state.

\bigskip
\noindent\textbf{Formal statement.} The bypassing property (BP) is defined as:
\begin{equation}
\text{BP} \;:\; \forall n \in \text{nodes},\; \forall e \in \text{exceptions}_{\text{active}}(n),\; \text{terminus}(e) = h(n). \tag{BP}
\end{equation}
That is: for any node and any active exception at that node, the exception's terminus is exactly and only the human anchor $h(n)$. No machine actor may appear in the set $\{ \text{terminus}(e) \mid e \in \text{exceptions}_{\text{active}}(n) \}$.

\bigskip
\noindent\textbf{Why bypass is architecturally necessary.} The bypass is not an optimization — it is a structural necessity that follows from the upstream accountability guarantee of the \bxthree{} Framework. Any machine actor $m$ that could absorb an exception in ESCALATING state would become a terminal node for that exception, and the accountability chain would terminate at $m$ rather than at a human. If $m$ then fails, or if $m$'s resolution is incorrect or insufficient, no human principal is present in the chain to detect or correct the failure. The system compounds failure instead of surfacing it. This is the precise failure mode the Bailout Protocol is designed to prevent.

Conventional escalation hierarchies — including tiered oversight models that permit absorption at intermediate tiers — make human oversight a high-tier option rather than a compulsory terminus. The Bailout Protocol forbids this absorption by architectural constraint: $h(n)$ is not one option among many — it is the \emph{only} permissible terminus for every unresolved exception.

The bypass eliminates three specific failure modes by architectural constraint:

\begin{enumerate}
\item \textbf{Chain termination at a machine actor.} If the escalation path contains a machine actor $m$, and $m$ can absorb the exception, then the chain terminates at $m$. The human principal $h(n)$ never receives the exception. The upstream accountability guarantee is violated.

\item \textbf{Hidden failure compounding.} An absorbing machine actor $m$ may resolve the exception incorrectly or incompletely, without any external detection mechanism. The system continues operating in a compromised state. The Forensic Ledger records only the absorption event, not the subsequent failure cascade, because the exception has been removed from the protocol's observable state.

\item \textbf{Privilege escalation via absorption suppression.} A sophisticated machine actor with access to the Fact Layer could theoretically suppress a warranted escalation by issuing a retroactive resolution approval after $T_{\text{bounds}}$ has expired. The Fact Layer gate prevents this by rejecting any resolution approval written after the trigger timeout, regardless of the resolution's content or the approving actor's privilege level.
\end{enumerate}

The bypass is structurally analogous to a hardware interrupt line hardwired to a designated handler — just as a hardware interrupt cannot be intercepted or redirected by user-space code, the bailout escalation cannot be intercepted or redirected by any agent-level component.

\bigskip
\noindent\textbf{Enforcement mechanisms.} The bypass is enforced by two cooperating mechanisms operating at different layers of the \bxthree{} architecture:

\begin{enumerate}
\item[\textbf{Fact Layer gate (enforcement at the authorization layer).}] The \fact{} layer is the only mechanism capable of authorizing state transitions in the \bxthree{} Framework. The Bailout Monitor's pre-commit hook rejects any ledger entry for a resolution where the underlying trigger satisfies TC and $T_{\text{bounds}}$ has expired. This prevents the Bounds Engine from issuing a retroactive resolution after timeout to suppress warranted escalation. The \fact{} layer gate operates identically on all machine actors: there is no privileged machine actor with the ability to bypass this check.

\item[\textbf{Immutable parent-pointer and human-root architecture (enforcement at the data model layer).}] The parent pointer $\alpha(n)$ and the human-root identifier $h(n)$ are stored in \fact-layer-managed memory and are read-only at the machine-actor level. A machine actor cannot redirect the parent chain, cannot insert itself as an intermediate resolver, and cannot modify the $humanRootId$ field that terminates the chain. Any attempted modification requires \purpose{} layer authorization with a corresponding ledger entry — which cannot be completed without human approval if the modification would affect the bailout terminus.
\end{enumerate}

Together, these mechanisms ensure that BP holds for all $n \in \text{nodes}$ under all execution conditions, including partial node failure, network partition, and deliberate attempt by a compromised machine actor to redirect the bailout pathway.

% ---------------------------------------------------------------
\subsection{Timeout Handling}
\label{sec:timeout-handling}

Timeouts in the Bailout Protocol serve a liveness function: they prevent the protocol from permanently stalling when a parent node is unreachable, a communication pathway is unresponsive, or the human anchor is temporarily unavailable. Each timeout has a defined scope, a defined scope-exit action, and an escalation path that prevents deadlock.

\bigskip
\noindent\textbf{Timeout definitions.}

\begin{enumerate}\itemsep=0pt
\item[\textbf{$T_{\text{bounds}}$ — Bounds Engine resolution timeout.}] The maximum time a node may remain in BOUNDED state attempting autonomous resolution. Default: 60 seconds, provisioned per node class according to the expected latency of the node's resolution procedure. If $T_{\text{bounds}}$ expires without a \fact-layer-approved resolution, the node transitions to BAIL\_PENDING $\rightarrow$ ESCALATING via T3 and T4. The \fact{} layer cannot extend $T_{\text{bounds}}$ unilaterally; any extension requires human authorization via a separate ledger entry recorded before $T_{\text{bounds}}$ expires.

\item[\textbf{$T_{\text{prop}}$ — Per-hop propagation timeout.}] The maximum time node $n$ waits for an acknowledgment from its parent after sending an exception via \texttt{SendToParent}. Default initial value: 5 seconds. If $T_{\text{prop}}$ expires without ACK, the algorithm retries with exponential backoff: $T_{\text{prop}} \leftarrow \min(2 \cdot T_{\text{prop}},\; T_{\text{cap}})$. This backoff continues until either an ACK is received or $N_{\max}$ retries have been exhausted at $T_{\text{cap}}$.

\item[\textbf{$T_{\text{cap}}$ — Maximum propagation timeout value.}] The ceiling value for the exponential backoff of $T_{\text{prop}}$. Default: 60 seconds. After $N_{\max}$ consecutive retries at $T_{\text{cap}}$ without ACK, the algorithm fires a \texttt{parent\_unreachable} timeout event, which propagates to $h(n)$ via the next available functional pathway tracked by the Pathway Health Registry.

\item[\textbf{$T_{\text{human}}$ — Human acknowledgment timeout.}] The maximum time the human anchor may remain in HUMAN\_ACKNOWLEDGED state without issuing a directive. Default: 300 seconds (5 minutes), configurable per deployment based on the expected human response time for the node's operational domain. If $T_{\text{human}}$ expires, the protocol issues a reminder notification to $h(n)$ and transitions the originating node to RESOLVED with an \texttt{unresolved} tag. The node becomes quiescent: no autonomous retry occurs, no further actions are attempted, and the \texttt{unresolved} tag in the Forensic Ledger signals that human re-engagement is required to clear the state. The protocol does not re-escalate automatically; automatic re-escalation would create a potential denial-of-service condition against the human anchor.
\end{enumerate}

\begin{table}[htbp]
\caption{Timeout parameters, their start conditions, and their functional scopes.}
\label{tab:bailout-timeouts}
\begin{tabular}{@{}lcl@{}}
\toprule
\textbf{Timeout} & \textbf{Started at} & \textbf{Scope} \\
\midrule
$T_{\text{bounds}}$ & T1 fires (TC satisfied) & Bounds Engine autonomous resolution window \\
$T_{\text{prop}}$ & T4 fires (transmit to $h(n)$) & Per-hop parent acknowledgment \\
$T_{\text{cap}}$ & T6 fires (retry) & Ceiling for $T_{\text{prop}}$ backoff \\
$T_{\text{human}}$ & ESCALATING + $h(n)$ contact confirmed & Human directive issuance window \\
\bottomrule
\end{tabular}
\end{table}

\bigskip
\noindent\textbf{Pathway Health Registry (PHR).} The PHR is a runtime data structure maintained by the Bailout Monitor that tracks the availability, latency, and error rate of each provisioned communication pathway to the human anchor. Before each \texttt{SendToParent} call, \texttt{SelectPathway} queries the PHR and selects the pathway with the lowest estimated latency among currently functional pathways. Pathways that fail $K$ consecutive health checks (default $K = 3$) are marked \texttt{DEGRADED} and excluded from pathway selection until a health check succeeds, preventing the algorithm from repeatedly selecting a known-degraded pathway.

The PHR is updated by a background health-ping thread that sends heartbeat messages along each provisioned pathway at intervals of $T_{\text{health}}$ (default: 30 seconds). The health-ping thread operates with the same scheduling priority as the Bailout Monitor and is not preemptible by agent tasks, ensuring that pathway status remains current even under heavy agent workload.

\bigskip
\noindent\textbf{Liveness guarantee.} Together, $T_{\text{bounds}}$, $T_{\text{prop}}$, $T_{\text{cap}}$, and $T_{\text{human}}$ define a finite upper bound on the wall-clock time from trigger firing (T1) to human acknowledgment under any combination of parent unavailability and pathway degradation. This bound is a deployment parameter — not a protocol constant — derived from the network topology, the number of hops to the human anchor, and the provisioned $T_{\max}$ parameter:
\begin{equation}
T_{\max} \;=\; T_{\text{bounds}} \;+\; N_{\max} \cdot T_{\text{cap}} \;+\; T_{\text{human}} \tag{T-max}
\end{equation}

A deployment that cannot guarantee human acknowledgment within the required $T_{\max}$ under all failure scenarios must provision additional or higher-bandwidth pathways, or must reduce $T_{\text{prop}}$ and $T_{\text{cap}}$ to bring the worst-case bound within the required limit. This analysis is a required part of deployment certification for any \bxthree{} system operating in a regulated or safety-critical environment.

\bigskip
\noindent\textbf{Timeout interaction with the bypass mechanism.} The bypass mechanism ensures that timeouts cannot be weaponized by a machine actor to delay or suppress escalation. Specifically: (a) $T_{\text{bounds}}$ is enforced by the \fact{} layer pre-commit hook, not by the \bounds{} layer; a machine actor cannot artificially extend $T_{\text{bounds}}$ to prevent the transition to ESCALATING. (b) $T_{\text{prop}}$ backoff is calculated by the Bailout Monitor, not by any machine actor; the backoff curve is defined in the protocol specification and cannot be altered at runtime. (c) If all pathways time out at $T_{\text{cap}}$, the protocol resolves with \texttt{PATHWAY\_EXHAUSTED} — this is a terminal RESOLVED state, not a suppression of escalation. The \texttt{PATHWAY\_EXHAUSTED} tag in the Forensic Ledger ensures that the failure of the pathway is recorded as an incident requiring human remediation, not as a successfully resolved bailout.