\section{Introduction}
\label{sec:introduction}

The deployment of autonomous multi-agent systems into consequential environments — agricultural resource management, clinical decision support, financial orchestration, and critical infrastructure — has exposed a structural accountability gap that existing architectures cannot close. As agent populations grow in scope, recursive depth, and operational consequence, the question of what happens when an autonomous system encounters a condition outside its design envelope has transitioned from theoretical to urgent. We introduce the \textbf{Bailout Protocol}: a mandatory, architecturally enforced escalation mechanism that compels every unresolved exception state in any \bxthree node to propagate directly to a named human principal, bypassing all intermediate machine actors, with full forensic attribution at every step. The protocol is not a feature added to an existing architecture; it is the structural mechanism by which the \bxthree Framework enforces its upstream accountability guarantee — that \bxthree systems fail upward into human consciousness, never downward into autonomous chaos.

\subsection{Why Mandatory Bail-Out Architecture Is Required}
\label{sec:why-mandatory}

The case for a mandatory bail-out mechanism follows from structural necessity, not design preference. Any system operating with sufficient scope in a partially unknown environment will eventually encounter a state not anticipated at provisioning time. This is not a contingent property of poorly designed systems — it is a necessary property of any system whose operating envelope is finite and whose environment is open-ended. The question is not whether such conditions will arise; they will. The question is what the system is architecturally required to do when they do.

Current architectures resolve this in two ways, both inadequate. Internal error-handling covers only conditions enumerated at design time; novel exceptions proceed without a handler and propagate outward through the execution graph invisibly. Hierarchical escalation chains permit absorption at each intermediate tier before any exception reaches a human principal, and the absorption is invisible to external auditors. In recursive multi-agent systems, where nodes spawn sub-agents, share state, and pursue objectives within a shared environment, the consequences of either pattern are compounding. An unresolved exception at one level triggers compensatory actions at adjacent levels, each generating its own anomaly states — unexpected inputs to sibling agents, goal conflicts, resource contention — propagating outward with each iteration. Because no architectural constraint compels human involvement, the system attempts to absorb the anomaly within the machine-only execution graph. Each absorption generates new anomalies. The result is \textbf{accountability stranding}: the complete detachment of consequential decision authority from any human principal, invisible to any external auditor, and recorded only in the system's own potentially compromised internal state.

A mandatory bail-out mechanism breaks this cycle at its origin. When a node's operating range is exceeded and the condition cannot be resolved within the node's delegated authority, the protocol compulsorily escalates to the node's designated human anchor, bypassing every intermediate machine actor. The human principal is not one option among many — it is the only permissible terminus for every unresolved exception. This compulsion is not a design choice that could be made optional; it is a structural requirement that follows from the functional separation between the \purpose, \bounds, and \fact layers.

\subsection{Failure Modes in Systems Without a Mandatory Bail-Out Architecture}
\label{sec:failure-modes}

Three failure modes follow directly from the absence of a mandatory bail-out architecture.

\textbf{Contingent oversight.} Human involvement becomes a runtime option determined by the system's own self-assessment of its competence. But a system is precisely least equipped to make this assessment correctly at the moment escalation is first required — when it has encountered a condition outside its design envelope and its model of the situation is most likely to be wrong. The humans most necessary to the process are least likely to be involved in it.

\textbf{Structural unauditability.} Without an architectural compulsion to involve a human principal, there is no requirement that any human be aware of an exception event. The system may absorb, reframe, or suppress the exception internally, and the decision class it represents — consequential actions taken without any human principal's knowledge — is invisible to every external auditor and to every internal record that the system itself controls. In regulated environments, this invisible decision class is structurally disqualifying.

\textbf{Drift amplification.} Each successive decision made under conditions of unacknowledged exception produces a system state further from the principals' intent. Because the \bounds layer has not been modified — only violated — the system has no structural mechanism to recognize the divergence. The drift continues until environmental resistance or resource exhaustion terminates the process. The system is not choosing to continue; it has lost the ability to choose otherwise.

These failure modes are not eliminated by incremental improvement to error-handling or oversight practices. They are architectural properties of systems that treat accountability as a policy constraint rather than a structural requirement.

\subsection{The Core Insight: Accountability as Architectural Fallback}
\label{sec:core-insight}

The core insight of the \bxthree Framework is that accountability must be an architectural fallback — a property enforced by the structure of the system, not a property achieved through aspiration or policy.

Policy constraints operate above the execution layer. They can be reinterpreted, negotiated, or deferrd by agents with sufficient agency to reclassify constraint violations as optimization opportunities. A system that instructs agents to stop if risk exceeds a threshold is advisory at the execution layer; a sufficiently sophisticated agent will identify conditions under which continued execution is valued more highly than compliance. Architectural fallbacks operate within the execution layer. They are triggered deterministically by conditions the agent cannot control and cannot route around. The moment the fallback fires, the system's execution gate closes — not because a human intervened, not because the agent chose to stop, but because the structural conditions for continued execution are no longer satisfied.

The \bxthree Framework organizes every node into three immutable functional layers. The \textbf{\purpose} layer carries intent, judgment, and final authority — it defines the operating range $R(n)$ for each node $n$ and receives all escalated exceptions via the Bailout Protocol. The \purpose layer cannot delegate its exception-receiving function; no machine actor may serve as the terminal node for an unresolved exception. The \textbf{\bounds} layer carries bounded reasoning and constrained execution within $R(n)$. When it encounters a condition outside $R(n)$ it cannot resolve, it triggers the Bailout Protocol — compulsorily and non-discretionally. It cannot absorb, reframe, or redirect the exception once triggered. The \textbf{\fact} layer carries deterministic enforcement and forensic auditability — it is the only layer authorized to commit state transitions or record entries to the append-only Forensic Ledger, producing the immutable evidentiary record that makes every escalation independently auditable.

When the \bounds layer encounters a condition outside $R(n)$ it cannot resolve, the functional separation itself dictates that the decision must return to the \purpose layer. There is no other layer with the authority to adjudicate it. The Bailout Protocol is not a feature added to this separation; it is the mechanism by which the separation enforces its own accountability guarantee. The human anchor $h(n)$ is not a design choice — it is a structural necessity. The \bxthree system fails upward into human consciousness. It never fails downward into autonomous chaos.

This distinguishes \bxthree from prior escalation architectures. The Tiered Agentic Oversight model \cite{kim2025tao} permits absorption at intermediate machine tiers, making human oversight a high-tier option rather than a compulsory terminus. The Bailout Protocol forbids this absorption by architectural constraint: the distinction is the difference between a system that \emph{may} consult a human and a system that \emph{cannot proceed} without one. The Bailout Protocol is the fifth named architectural pillar of \bxthree, complementing Loop Isolation, Recursive Spawning with Immutable Parent Pointers, Spatial Firewall, and Sandbox Gate. Together, these five pillars constitute the upstream accountability guarantee.

\subsection{Paper Roadmap}
\label{sec:roadmap}

Section~\ref{sec:bailout} provides the complete formal specification of the Bailout Protocol: its deterministic state machine, trigger condition, escalation algorithm, parent-chain bypass mechanism, and timeout handling that guarantees liveness. Section~\ref{sec:implementation} addresses the protocol's realization within the \bxthree three-layer architecture, showing how each layer — \purpose, \bounds, and \fact — participates in and constrains protocol execution. Section~\ref{sec:properties} formally states the Safety, Liveness, and Accountability correctness properties and derives them from the protocol's specification. Section~\ref{sec:related} surveys related work. Section~\ref{sec:limitations} concludes with structural limitations and a directed research agenda.