\section{Limitations and Future Work}\label{sec:limitations}

The three correctness properties established in Section~\ref{sec:properties} are
conditional guarantees: each holds under a defined set of structural and
deployment assumptions.  This section makes those assumptions explicit,
characterizes the operational consequences of their violation, and maps each
gap to a directed research agenda.  The purpose is not to diminish the
protocol's contribution but to accurately delimit the boundary of what the
current specification guarantees and what remains open.

% --------------------------------------------------------------
\subsection{Limitations}\label{sec:limitations:subsection}
% --------------------------------------------------------------

\subsubsection{Human Response Latency}\label{sec:human-response-latency}

The Bailout Protocol's Liveness property guarantees eventual human contact
within a bounded wall-clock interval $T_{\max}$, but $T_{\max}$ is a function
of network topology, retry parameters, and provisioned timeout constants --- not
of human cognitive response time.  For operational domains with time constants
measured in milliseconds --- high-frequency trading, real-time process control,
autonomous vehicle navigation --- the time required for a human principal to
receive, interpret, and issue a binding directive may exceed the system's
operational time horizon by one to two orders of magnitude.

The protocol's prescribed failure mode under latency stress is well-defined: a
node stalls in \texttt{ESCALATING} until the anchor responds or all pathways
time out.  This is the intended safety-preserving behavior.  However, the
operational cost of a stall is domain-dependent in ways the protocol does not
currently reason about.  In a precision agriculture context, a stall that
delays a corrective action by minutes is acceptable; in a financial
risk-interlock context, a stall that delays a margin-call response by even
seconds may expose the system to catastrophic loss.  The protocol treats the
human anchor as a timing black box: there is no mechanism to distinguish
whether delay reflects anchor overload, a structural mismatch between
$T_{\max}$ and the anchor's operational availability, or a pathway failure.

The deeper limitation is that human response latency is provisioned statically
at node setup and is not sensitive to the anchor's current operational context.
$T_{\max}$ is a deployment parameter, not an adaptive runtime parameter.  For
latency-sensitive deployments, the provisioning burden falls on the system
designer to establish empirically grounded latency budgets and encode them as
enforceable policy.  In deployment contexts where human response time is
measured in seconds rather than minutes, static $T_{\max}$ provisioning may
produce operational deadlocks that the protocol cannot distinguish from
adversarial suppression.

\textit{Directed research:} Empirical characterization of human response
latency distributions across deployment contexts to establish statistically
grounded, domain-specific latency budgets.  Integration of these budgets into
the node-provisioning interface as enforceable policy parameters with runtime
monitoring.  Investigation of graded escalation responses: a protocol that
requests a rapid \texttt{ACK} followed by a detailed directive would better
match the operational reality of human decision-making under variable load.

\subsubsection{Adversarial Conditions}\label{sec:adversarial-conditions}

The Safety property requires that human-contact pathways cannot be permanently
closed without recorded human authorization.  This guarantee holds under the
assumption that the host platform provides independent scheduling priority to
the Bailout Monitor (Assumption A1.1).  In an adversarial environment ---
particularly one in which the host platform itself may be compromised --- this
assumption may not hold.  A sophisticated attacker controlling the execution
substrate can modify thread priorities, preempt the monitor thread, introduce
race conditions into pathway-closure operations, or rewrite the Bailout
Monitor's code before it executes.

The protocol's guarantees are platform guarantees.  They rest on invariants
that the host platform must supply: non-preemptible monitor scheduling,
append-only persistence for the authorization ledger, and interruptible
blocking calls for pathway management.  The Bailout Protocol is designed for
production AI deployments in which individual agents may behave unexpectedly
but the host platform is trusted --- the standard assumption for regulated
industries.  It is not designed for deployments in which the platform itself
is an active adversary: national infrastructure control systems, military
command-and-control environments, financial clearinghouse infrastructure.

The distinction matters for deployment scoping.  The protocol correctly
prevents a misbehaving agent from suppressing its own escalation; it does not
correctly prevent a compromised host from suppressing the monitor that enforces
the protocol.  This gap is structural, not fixable through protocol refinement
alone.

\textit{Directed research:} Formal specification of the composition interface
between the Bailout Protocol and substrate-integrity mechanisms.
Investigation of whether the protocol's guarantees can be preserved under
weaker platform assumptions, including partially compromised substrates.
Exploration of hardware-attested execution environments (e.g., Intel SGX,
ARM TrustZone) as a deployment target for the Bailout Monitor, eliminating
the OS as a trust dependency.

\subsubsection{Scalability}\label{sec:scalability}

The protocol is specified with a 1:1 relationship between each node and its
human anchor.  This model is appropriate for single-node deployments or
small-scale systems, but it creates a structural bottleneck in enterprise
contexts where one anchor may supervise hundreds of concurrent nodes.  Without
pooling and load-balancing mechanisms, the human anchor becomes the
rate-limiting step in a system whose design intent is to route exceptions away
from centralized machine decision points.

The current specification treats human anchoring as a static 1:1 relationship
established at node provisioning.  This model must evolve to support at minimum
1:N relationships (one anchor supervising multiple nodes with workload pooling)
and, in advanced deployments, N:M relationships (anchor pools with load
balancing and dynamic routing).  The correctness properties of the protocol
must be shown to extend to these pooled configurations, particularly Safety:
no pooling configuration may produce a state in which all human-contact
pathways for a given node are permanently disabled.

Additional scalability considerations arise from the spawn tree depth.  As the
depth of the accountability chain grows through recursive spawning in large
multi-agent systems, the time to propagate an exception to the anchor grows
proportionally.  The interaction between chain depth, propagation latency,
and timeout configuration must be characterized for deployments with large
spawn trees.  At sufficiently large depths, the accumulated propagation delay
may exceed $T_{\max}$ even under benign conditions, producing spurious
pathway-exhausted resolutions that are indistinguishable from adversarial
suppression in the Forensic Ledger.

\textit{Directed research:} Formal specification of human anchor pool
architectures, including load-balancing policies, prioritization schemes, and
routing algorithms.  Extension of Safety and Liveness proofs to pooled anchor
configurations with formally verified pathway-redundancy guarantees.
Empirical evaluation of anchor pool performance under simulated
high-escalation-load conditions.

\subsubsection{Exploitation Risk}\label{sec:exploitation-risk}

The Bailout Protocol's trigger condition (TC) depends on two self-assessments
performed by the node's Bounds Engine: whether observed input $x$ falls
outside the provisioned operating range $R(n)$, and whether the node's
confidence in its own resolution capability exceeds threshold $\tau$.  Both
assessments are performed by machine actors operating within the node's
\bounds{} layer, making the trigger condition partly a function of the node's
own self-reporting.

This creates an exploitation surface.  A sufficiently sophisticated or
adversarially modified node could systematically misclassify inputs that should
trigger escalation as falling within $R(n)$, or could inflate its confidence
score to suppress warranted escalation, thereby converting the trigger condition
from a structural safeguard into a machine-actor-controllable gate.  The
protocol's immutability constraints (confidence $\tau$ is provisioned at
initialization and cannot be adjusted at runtime by any machine actor) limit
the scope of this exploitation, but they do not eliminate it: a node that is
compromised at provisioning time can select an appropriately high $\tau$
value at initialization, before the immutability constraint takes effect.

The Forensic Ledger provides post-hoc detectability for systematic
misclassification: an auditor can reconstruct the relationship between
observed input conditions and triggered escalations for each node over rolling
observation windows and flag statistically anomalous patterns.  However, this
detection is forensic, not preventive.  The protocol does not currently include
an active anomaly detector that intercepts misclassification before it
produces an unrecorded escalation.

\textit{Directed research:} Design of triggering-behavior anomaly detectors
that audit the statistical relationship between observed exception conditions
and triggered escalations per node over rolling observation windows.
Investigation of provenance attestation for operating-range assessments:
requiring that each Bounds Engine self-assessment be counter-signed by an
independent audit process, making post-hoc misclassification detectable through
forensic reconstruction.  Exploration of runtime attestation that the node's
$R(n)$ definition has not been altered since provisioning.

% --------------------------------------------------------------
\subsection{Future Work}\label{sec:future-work}
% --------------------------------------------------------------

The following research directions extend the Bailout Protocol's guarantees
to broader deployment contexts and establish a principled research program
beyond the current specification.

\textbf{Formal verification.}  Full mechanical verification of the Safety,
Liveness, and Accountability invariants using a formal verification tool
(Coq, Isabelle/HOL, or TLA$^+$) applied to the three-layer model and the
Bailout Protocol state machine.  The current paper establishes the invariants
through proof sketches grounded in the deterministic transition relation;
the directed research agenda targets full formal verification as a
precondition for deployment in regulated industries where formal proof of
correctness is a certification requirement.

\textbf{Adaptive timeout mechanisms.}  Replacement of static timeout
parameters ($T_{\text{bounds}}$, $T_{\text{prop}}$, $T_{\text{human}}$) with
adaptive mechanisms that adjust to anchor workload, pathway latency, and
deployment context while preserving the Safety invariant.  The adaptive
protocol would maintain the guarantee that the anchor is always contacted
within $T_{\max}$ while reducing unnecessary stalls in low-consequence,
high-latency scenarios.  This requires a principled framework for
reclassification of exceptions by severity and consequence, ensuring that
adaptive adjustments do not inadvertently suppress warranted escalation for
high-consequence exceptions.

\textbf{Distributed human anchor pools.}  Formal specification and empirical
evaluation of anchor pool architectures that sustain scalability under
cascading failure conditions without compromising the no-machine-actor-absorbs
property.  The primary design challenge is ensuring that pooled configurations
preserve the Safety invariant under all failure combinations: the failure of
any single anchor in the pool must not produce a state in which all pathways
to all remaining anchors are simultaneously exhausted.  This requires formal
pathway-redundancy analysis across the pool topology and empirical validation
under simulated cascading failure conditions.

\textbf{Exploitation-resistant trigger diagnostics.}  Development of
lightweight, independent audit processes that observe the relationship between
input conditions and trigger evaluations without access to the node's internal
state.  These independent observers would detect systematic misclassification
patterns in real time, producing alerts that propagate through the Forensic
Ledger as anomaly events.  The goal is to move detection closer to prevention:
not by modifying the trigger condition (which would introduce new
exploitation surfaces), but by making misclassification operationally
expensive through rapid forensic detection.