\section{Immutable Parent Pointers}
\label{sec:immutable-parent-pointers}

The immutable parent pointer is the foundational structural primitive of the Recursive Spawning Protocol. It is the single mechanism that converts the delegation chain from a forensic reconstruction attempted after the fact into a runtime-verifiable artifact, that elevates chain integrity from a policy assertion to a structural guarantee, and that renders all modes of autonomous drift architecturally impossible regardless of operational urgency, architectural complexity, or adversarial sophistication of any machine actor in the chain. This section specifies the parent pointer formally, defines the chain traversal operator $\Chain{\cdot}$, establishes the root human anchor as the explicit base case, specifies immutability enforcement through the Fact Layer write protocol, describes the Purpose Layer's mandatory role in spawn authorization, and provides the complete chain traversal algorithm with correctness arguments.

% ------------------------------------------------------------------------------%
\subsection{Formal Definition: ParentPointer}
\label{sec:parent-pointer-definition}

Let $\mathcal{N}$ be the set of all agent nodes in a \bxthree{} system implementing the Recursive Spawning Protocol, and let $P$ be the set of all Purpose Layer actors. The \emph{parent pointer} $\alpha(n)$ is defined for every node $n \in \mathcal{N}$ as a total function:

\begin{dfn}[Parent Pointer]
\label{dfn:parent-pointer}
\begin{equation}
\alpha : \mathcal{N} \longrightarrow P \cup \{\bot\}
\end{equation}
where $\alpha(n) = p$ if $p \in P$ is the Purpose Layer actor that authorized the spawn event resulting in $n$'s activation, and $\alpha(n) = \bot$ if and only if $n$ is the root human anchor established at system initialization. The pointer is established once at node provisioning via a Fact Layer write operation and satisfies the following immutability property:
\begin{equation}
\forall n,\ p:\ \bigl(\alpha(n) = p \ \land \ \text{provisioned}(n)\bigr) \ \implies \ \neg\exists\ \text{modify}(\alpha(n))
\label{eq:parent-pointer-immutability}
\end{equation}
No actor --- including the Purpose Layer after provisioning, the Bounds Engine, any administrative process, or any external principal --- may execute an operation that overwrites, redirects, or deletes $\alpha(n)$ after the provisioning write is confirmed. The Fact Layer write protocol rejects any such attempt at the protocol level, before execution reaches the ledger.
\end{dfn}

The parent pointer is not a reference variable held by the parent node toward its child. It is a Fact Layer property of the child node, established at provisioning time and stored in the node's sealed memory envelope. The directionality is architecturally critical: the child carries its parent reference as an immutable intrinsic property, not as a mutable external reference held by the parent. This inversion of reference direction means the chain is traversable from any node outward to its full lineage regardless of the operational state of any other node. A parent that has terminated, crashed, or been decommissioned does not affect the child's ability to reference its parent through $\alpha(n)$, because $\alpha(n)$ is a property of the child, not a handle held by the parent.

The parent pointer is architecturally distinct from a conventional access-controlled variable in a second critical respect: access control restricts \emph{which principals} may issue modification operations; if sufficient privilege is acquired through credential theft, software vulnerability, or administrative compromise, the access control barrier is circumvented and immutability is defeated. The Fact Layer write protocol enforces immutability by making the modification operation \emph{structurally impossible} --- there is no mechanism for processing a $\text{modify}(\alpha(n))$ request regardless of privilege level. There is no privileged override, no emergency bypass, and no administrative role with the capability to alter the parent pointer after provisioning. This is not a stronger access control policy; it is a different architectural category.

The initialization of $\alpha(n)$ occurs atomically during the spawn event. The Fact Layer records the spawn event in the forensic ledger and writes $\alpha(n) = p$ into the node's sealed memory envelope as part of the same atomic transaction. This separation of authorization (Purpose Layer) from assignment (Fact Layer) is essential: it prevents a parent from claiming to have spawned a node whose parent pointer was set to a different actor, and it prevents a child from falsifying its own ancestry after provisioning.

\begin{dfn}[Spawn Event]
\label{dfn:spawn-event}
A \emph{spawn event} is the 5-tuple
\begin{equation}
\operatorname{spawn}(c,\ p,\ \sigma_c,\ t,\ \phi)
\end{equation}
where $c$ is the newly created child node, $p \in P$ is the authorizing Purpose Layer actor, $\sigma_c$ is the authorized execution scope, $t$ is the wall-clock timestamp, and $\phi$ is the cryptographic signature produced by the Purpose Layer over the spawn request. The Fact Layer atomically: (1) creates node $c$, (2) sets $\alpha(c) \leftarrow p$, (3) sets the node's authorized scope to $\sigma_c$, (4) seals the memory envelope, and (5) appends the event to the forensic ledger $\mathcal{L}$.
\end{dfn}

\begin{prop}[Immutability of ParentPointer]
\label{prop:parent-pointer-immutable}
For any child node $c$ with parent $p$, once the spawn event $\operatorname{spawn}(c, p, \sigma_c, t, \phi)$ has been recorded in the forensic ledger $\mathcal{L}$, the value $\alpha(c)$ is fixed for the operational lifetime of $c$. No subsequent operation --- including operations issued by $p$, by any Purpose Layer actor, by the Bounds Engine, or by $c$ itself --- can alter $\alpha(c)$.
\end{prop}

This property distinguishes $\ParentPointer{p}{c}$ from conventional mutable parent references in multi-agent systems. A mutable reference can be overwritten by a subsequent administrative operation; the parent pointer cannot. The distinction is not behavioral --- it is architectural.

% ------------------------------------------------------------------------------%
\subsection{The Chain Operator}
\label{sec:chain-operator}

The chain operator $\Chain{\cdot}$ constructs the complete delegation lineage from any node to the human anchor by recursively applying the parent pointer function. This operator is the primary tool for chain verification, audit, and accountability attribution.

\begin{dfn}[Chain Operator]
\label{def:chain-operator}
For any agent node $a$ in a Recursive Spawning chain, the \emph{chain} $\Chain{a}$ is defined recursively as:
\begin{equation}
\Chain{a} =
\begin{cases}
\{\, a \,\}\ \cup\ \Chain{\alpha(a)} & \text{if }\ \alpha(a) \neq \bot \\[8pt]
\{\, a,\ h(a) \,\} & \text{if }\ \alpha(a) = \bot\ \text{(root human anchor)}
\end{cases}
\label{eq:chain-operator}
\end{equation}
where $h(a)$ is the human anchor designation of the root node. Equivalently, expanding the recursive definition:
\begin{equation}
\Chain{a} = \bigl\{\, a,\ \alpha(a),\ \alpha(\alpha(a)),\ \alpha(\alpha(\alpha(a))),\ \ldots,\ n_0 \,\hmannchor{n_0}\, \bigr\}
\label{eq:chain-expanded}
\end{equation}
where $n_0$ is the root of the chain satisfying $\alpha(n_0) = \bot$ and $\hmannchor{n_0}$ is the human anchor designation.
\end{dfn}

The chain is the set of all nodes on the delegation path from $a$ to and including the human anchor. It includes the terminal human anchor node, which is the only node in any RSP chain with $\alpha(n) = \bot$. The chain is always finite: by the RSP termination property (proved in Section~\ref{sec:rs-correctness}), every RSP chain has finite depth bounded by $D_{\max}$, so the recursion in Equation~\ref{eq:chain-operator} terminates at the root within at most $D_{\max} + 1$ applications.

The chain operator is unique and deterministic. For any node $a$, $\Chain{a}$ contains exactly one element for each ancestor node on the delegation path. There is no alternative chain, no ambiguous parentage, and no optional path. The immutability of $\alpha(n)$ guarantees that $\Chain{a}$ is identical at any runtime point to what it was at the moment of $a$'s provisioning. The chain is not reconstructed from logs --- it is directly computable by iterating $\alpha$ from any node until reaching $\bot$.

\begin{note}
\textbf{On the base case: root human anchor.}
The root of every RSP chain is a Purpose Layer entity with a human principal designation, established at system initialization when a human principal $h$ authorizes the first agent node $n_0$. The Purpose Layer records $h(n_0) = h$ and $\alpha(n_0) = \bot$ in the Fact Layer authorization ledger. This is the only base case in the chain: the root human anchor has no parent by definition. The non-collapsing human anchor property ensures $h(n) = h(n_0)$ for all nodes in all chains spawned from $n_0$. No machine actor can serve as a chain root in RSP; this is structurally enforced by the Fact Layer pre-commit hook, which rejects any spawn event whose root is not a registered human principal.
\end{note}

The relationship between the parent pointer and the chain operator can be made explicit: $\Chain{a} = \ParentPointer{\alpha(a)}{a} \cup \Chain{\alpha(a)}$, where $\ParentPointer{p}{c}$ denotes the singleton set containing the ordered pair $(p, c)$. This formulation makes explicit that the chain is built by adding one link at a time, each link being an immutable parent-child pair, until the root is reached.

% ------------------------------------------------------------------------------%
\subsection{Immutability Enforcement: Fact Layer Write Protocol}
\label{sec:fact-layer-write-protocol}

The immutability of $\alpha(n)$ is not a policy assertion enforced by privilege-gated access control. It is a structural property of the Fact Layer write protocol, which defines the set of valid state transitions for every node field and explicitly excludes modification of $\alpha(n)$ after provisioning. This section specifies the write protocol properties that collectively constitute protocol-level immutability.

The Fact Layer write protocol satisfies four properties:

\begin{enumerate}
\item[(W1)] \textbf{No modification operation defined.} The protocol specification contains no operation for altering $\alpha(n)$ after the provisioning write. This is not a permission --- it is an absence from the operation set. There is no \texttt{modify\_parent\_pointer}, no \texttt{reparent}, and no \texttt{redirect\_chain} operation in the Fact Layer interface.

\item[(W2)] \textbf{Sealed memory envelope.} The parent pointer field $\alpha(n)$ is stored in an encrypted envelope (AES-256-GCM) with an HMAC-SHA-256 computed over the ciphertext. The decryption key is held exclusively by the Fact Layer and is never exposed to the Bounds Engine, the Purpose Layer after provisioning, or the child node's own code. Any attempt to tamper with the envelope invalidates the HMAC and causes the Fact Layer to reject the read.

\item[(W3)] \textbf{Write-once, read-many with re-sealing.} The parent pointer $\alpha(n)$ is written exactly once --- at provisioning --- and is thereafter read-only. Each authorized read decrypts the envelope, returns the value, and re-encrypts and re-signs the envelope in a single atomic operation. There is no persistent plaintext state that could be subject to replay or corruption.

\item[(W4)] \textbf{Atomic provisioning.} The parent pointer $\alpha(c)$ and its Purpose Layer warrant $\phi$ are created as a single atomic Fact Layer write. There is no temporal window in which the pointer exists without authorization, or in which authorization exists without a corresponding pointer.
\end{enumerate}

These four properties together constitute \emph{protocol-level immutability}: the immutability of $\alpha(n)$ is a property of the protocol's operational definitions, not a property of an access control policy layered over a conventional read/write interface.

The sealed memory envelope mechanism reinforces immutability at the physical layer. The parent pointer field $\alpha(n)$ is encrypted at rest using a Fact Layer-only symmetric key. Even a compromised Bounds Engine --- one whose runtime has been subverted by an adversarial exploit --- cannot read $\alpha(n)$ because it never has access to the decryption key. The envelope carries an HMAC-SHA-256 signature computed over its contents; any modification to the ciphertext invalidates the signature and causes the Fact Layer to reject the read. After each authorized read, the Fact Layer re-encrypts and re-signs the envelope, preventing replay attacks based on stale plaintext.

The Fact Layer operates as a standalone, isolated process with its own protected memory space, architecturally inaccessible to the child node's address space. The combination of the protocol-level write constraint and the sealed envelope ensures that $\alpha(n)$ satisfies the immutability property in all deployment contexts, including partial system failure, process restart, and targeted exploitation of adjacent components.

\begin{prop}[Protocol-Level Immutability]
\label{prop:protocol-level-immutability}
For any provisioned node $n \in \mathcal{N}$, the parent pointer $\alpha(n)$ cannot be modified by any actor, under any operational circumstance, through any software vulnerability, credential compromise, or administrative privilege. The Fact Layer write protocol has no defined operation for this modification, and the sealed memory envelope has no exposed decryption interface.
\end{prop}

This is the precise failure mode that the Fact Layer write protocol eliminates: retroactive manipulation of chain topology for the purpose of accountability laundering. Without protocol-level immutability, an adversarial actor could re-parent a node to a favorable parent after the fact, obscuring the original authorization chain. With protocol-level immutability, this manipulation is structurally impossible.

% ------------------------------------------------------------------------------%
\subsection{Spawn Authorization: Purpose Layer Role}
\label{sec:purpose-layer-spawn-authorization}

The Purpose Layer plays two structurally necessary roles in the parent pointer lifecycle: it originates the spawn authorization policy that defines the conditions under which a new parent pointer may be created, and it issues the spawn warrant that is the prerequisite for the Fact Layer provisioning write.

\medskip
\noindent\textbf{Authorization policy origination.} At system initialization, the Purpose Layer defines the spawn authorization policy $P_{\mathrm{spawn}}$ and records it in the Fact Layer authorization ledger. $P_{\mathrm{spawn}}$ specifies the mandatory conditions for any valid spawn event:

\begin{enumerate}
\item[(S1)] \textbf{Scope refinement.} The child scope must be a strict subset of the parent scope: $R(n_{\mathrm{child}}) \subset R(n_{\mathrm{parent}})$. No lateral expansion, no superset, no equality.

\item[(S2)] \textbf{Operational necessity.} The parent must declare an operational necessity that cannot be satisfied within $R(n_{\mathrm{parent}})$ and that justifies child creation. The necessity declaration is recorded in the Fact Layer authorization ledger and is subject to human anchor review.

\item[(S3)] \textbf{Depth within bound.} The resulting chain depth must not exceed $D_{\max}$. The depth bound is Purpose-Layer-defined and Fact-Layer-enforced.
\end{enumerate}

These three conditions are the only authorized paths to parent pointer creation. There is no bypass, no emergency override, and no operational urgency path that permits parent pointer creation without satisfying all three conditions. This is what makes the RSP's drift prevention property structural rather than conventional.

\medskip
\noindent\textbf{Spawn warrant issuance.} When a Purpose Layer actor $p \in P$ authorizes a spawn event $\operatorname{spawn}(c, p, \sigma_c, t, \phi)$, it produces the cryptographic warrant $\phi$, signing the spawn request using a Purpose-Layer-only private key. The warrant $\phi$ encodes the child identity, parent identity, authorized scope, and timestamp, binding these fields cryptographically. The Fact Layer pre-commit hook verifies $\phi$ before accepting the provisioning write. If the warrant is absent, malformed, or fails signature verification, the Fact Layer rejects the provisioning write and the node is not created. This warrant is the only mechanism by which the Fact Layer accepts a new parent pointer value; without a valid warrant, $\alpha(c)$ is never written.

The warrant mechanism ensures that parent pointer creation is a two-party contract: the Purpose Layer actor $p$ attests to the legitimacy of the spawn, and the Fact Layer independently verifies this attestation before committing the pointer. Neither party can unilaterally establish a parent pointer. The Purpose Layer cannot force the Fact Layer to write a pointer without a valid spawn event; the Fact Layer cannot generate its own authorization to write a pointer without a Purpose Layer warrant.

% ------------------------------------------------------------------------------%
\subsection{Chain Traversal Algorithm}
\label{sec:chain-traversal-algorithm}

The chain traversal algorithm is the runtime procedure for constructing the delegation chain from any node to its human anchor. It is provided here in full algorithmic detail with correctness arguments.

\medskip
\noindent\textbf{Algorithm: Chain-Traverse}

\begin{algorithm}
\caption{Chain-Traverse($n$) --- Build the delegation chain from node $n$ to the human anchor}
\label{alg:chain-traverse}
\begin{algorithmic}[1]
\REQUIRE $n$ is a provisioned RSP node with $\alpha$ defined
\ENSURE $\Chain{n}$ --- the ordered set of all ancestor nodes including the human anchor
\STATE $chain \leftarrow \langle\ \rangle$ \COMMENT{Initialize empty ordered list}
\STATE $current \leftarrow n$ \COMMENT{Start at the query node}

\WHILE{$\text{true}$}
    \STATE $\text{append}(chain,\ current)$ \COMMENT{Add current node to chain}
    \IF{$\alpha(current) = \bot$}
        \STATE \textbf{break} \COMMENT{Root reached --- chain complete}
    \ENDIF
    \STATE $current \leftarrow \alpha(current)$ \COMMENT{Move to parent}
\ENDWHILE

\RETURN $chain$
\end{algorithmic}
\end{algorithm}

\medskip
\noindent\textbf{Correctness arguments.}

\textit{Termination.} The algorithm terminates because RSP chains are depth-bounded by construction (proved in Section~\ref{sec:rs-correctness}). Each iteration of the while-loop moves $current$ to the parent node $\alpha(current)$, which is strictly closer to the root. Since the chain has finite depth at most $D_{\max} + 1$, the loop executes at most $D_{\max} + 1$ iterations before $\alpha(current) = \bot$ is reached. The loop therefore terminates in finite time.

\textit{Correctness of the returned set.} By Definition~\ref{def:chain-operator}, $\Chain{a}$ is the set containing $a$, $\alpha(a)$, $\alpha(\alpha(a))$, and so on, until the root. The algorithm produces exactly this sequence: it starts at $a$, appends each node, moves to its parent, and stops when the root is reached. The output is identical to $\Chain{a}$ by construction.

\textit{Uniqueness of result.} The algorithm produces a unique result for any input node because $\alpha$ is a total function (defined for every provisioned node) and is injective on the chain direction (every node has exactly one parent, and the root has no parent). There is no branching, no alternative path, and no non-deterministic choice.

\textit{Immutability guarantee.} The algorithm reads $\alpha$ values but never writes them. The traversal is therefore read-only and does not interact with the write protocol. The algorithm can be executed at any runtime point without affecting the immutability of the chain it traverses.

\medskip
\noindent\textbf{Algorithm: Chain-Verify}

The chain traversal algorithm enables a comprehensive verification procedure, Chain-Verify, which confirms the structural validity of a chain at any runtime point:

\begin{algorithm}
\caption{Chain-Verify($n$) --- Verify the authorization chain from node $n$ to human anchor}
\label{alg:chain-verify}
\begin{algorithmic}[1]
\REQUIRE $n$ is a provisioned RSP node
\ENSURE $\text{Boolean}$ indicating whether the chain is a valid RSP chain

\STATE $chain \leftarrow \text{Chain-Traverse}(n)$
\STATE $m \leftarrow \text{len}(chain)$

\STATE \COMMENT{\textit{Step V1: Verify chain terminates at human anchor}}
\IF{$\alpha(chain[m-1]) \neq \bot$}
    \STATE \RETURN $\text{false}$ \COMMENT{Root is not $\bot$ --- malformed chain}
\ENDIF
\IF{$\neg h(chain[m-1])$}
    \STATE \RETURN $\text{false}$ \COMMENT{Root is not a human principal}
\ENDIF

\STATE \COMMENT{\textit{Step V2: Verify each spawn event has a Purpose Layer warrant}}
\FOR{$i \leftarrow 0$ \TO $m-2$}
    \STATE $child \leftarrow chain[i]$
    \STATE $parent \leftarrow chain[i+1]$
    \IF{$\neg \text{warrant\_exists}(child,\ parent)$}
        \STATE \RETURN $\text{false}$ \COMMENT{Missing Purpose Layer warrant}
    \ENDIF
\ENDFOR

\STATE \COMMENT{\textit{Step V3: Verify scope refinement at each link}}
\FOR{$i \leftarrow 0$ \TO $m-2$}
    \STATE $child \leftarrow chain[i]$
    \STATE $parent \leftarrow chain[i+1]$
    \IF{$\neg (R(child) \subset R(parent))$}
        \STATE \RETURN $\text{false}$ \COMMENT{Scope refinement violated}
    \ENDIF
\ENDFOR

\STATE \COMMENT{\textit{Step V4: Verify depth bound at each node}}
\FOR{$i \leftarrow 0$ \TO $m-1$}
    \IF{$\text{depth}(chain[i]) > D_{\max}$}
        \STATE \RETURN $\text{false}$ \COMMENT{Depth bound exceeded}
    \ENDIF
\ENDFOR

\RETURN $\text{true}$
\end{algorithmic}
\end{algorithm}

Chain-Verify returns \texttt{true} if and only if all four verification conditions hold simultaneously. The verification does not inspect current operational state --- it verifies the structural chain: each parent pointer is correctly established, each spawn event has a Purpose Layer warrant, each scope is a descendant refinement of its parent, and each depth is within the architectural bound. A chain that passes Chain-Verify is, by the definitions of the RSP, a fully authorized delegation chain terminating at a human principal.

The algorithm is deterministic and produces the same result regardless of which node in the chain initiates verification. There is no self-serving interpretation of authorization that a drifted or orphaned node could produce. Chain verification is a property of the chain topology, not a property of the node being verified.

\medskip
\noindent\textbf{Constant-time escalation.} The depth bound $D_{\max}$ provides a constant-time upper bound on chain traversal cost. The worst-case number of pointer dereferences required to reach the human anchor from any node is $D_{\max} + 1$, independent of the total number of nodes in the system. A system with 10,000 spawned nodes and $D_{\max} = 5$ requires at most six pointer dereferences to reach a human --- the same as a system with 10 nodes. Mutable parent pointers would allow the chain to be extended retroactively, making the effective depth potentially unbounded and the worst-case escalation time unbounded as well. Immutability is what makes the depth bound a structural guarantee, not merely a policy.

In the Bailout Protocol context (Section~\ref{sec:bailout-protocol}), chain traversal is used to identify the correct Human Accountability Anchor when an agent at any depth encounters an unresolvable exception. The upward propagation of the exception signal follows the same parent-pointer structure as chain traversal, but uses the asynchronous trigger pathway rather than the synchronous verification pathway. Chain traversal ensures that this walk never diverges, never shortcuts, and never fails to reach a human.

% ------------------------------------------------------------------------------%
\subsection{Summary: Why Immutability Is Structural, Not Conventional}
\label{sec:immutability-structural-summary}

The immutable parent pointer is the load-bearing constraint of the Recursive Spawning Protocol. Without it, the chain is mutable --- subject to retroactive modification by actors with sufficient privilege --- and the RSP's correctness properties (drift prevention, chain integrity, termination) collapse from structural guarantees to policy conventions.

The immutability is structural because it is enforced by the Fact Layer write protocol, which defines no valid modification operation for $\alpha(n)$ after provisioning. This is architecturally distinct from access-control-based immutability in four ways:

\begin{enumerate}
\item[(1)] \textbf{No override path.} There is no privileged actor, administrative role, or emergency pathway capable of modifying $\alpha(n)$. Access control systems have override paths; the Fact Layer write protocol does not.

\item[(2)] \textbf{Failure-state resilience.} The immutability holds under all system failure conditions, including those that would disable other enforcement mechanisms. The write protocol is a local constraint that does not depend on distributed agreement or external service availability.

\item[(3)] \textbf{Source-equivalence of rejection.} All modification requests --- from any source --- receive identical treatment: protocol-level rejection. The Fact Layer does not distinguish between principals.

\item[(4)] \textbf{Atomic warrant-pointer creation.} The parent pointer and its Purpose Layer warrant are created as a single atomic Fact Layer write. There is no temporal window in which the pointer exists without authorization, or in which authorization exists without a corresponding pointer.
\end{enumerate}

The chain operator $\Chain{a}$ is uniquely determined by the immutable parent pointers in the chain. The chain traversal algorithm computes this operator correctly and terminates. Chain-Verify provides the runtime verification procedure for chain structural validity. Together, these mechanisms make the delegation chain an immutable, verifiable, and auditable runtime artifact --- the foundation on which all other RSP correctness properties rest.
