\section{Immutable Parent Pointers}
\label{sec:immutable-parent-pointers}

The immutable parent pointer is the foundational structural primitive of the Recursive Spawning Protocol. It is the single mechanism that makes the delegation chain a runtime-verifiable artifact rather than a forensic reconstruction, that makes chain integrity a structural guarantee rather than a policy assertion, and that renders all modes of autonomous drift architecturally impossible regardless of the operational urgency, architectural complexity, or adversarial sophistication of any machine actor in the chain.

This section specifies the parent pointer formally (Section~\ref{sec:parent-pointer-formal}), defines the chain traversal operator and establishes the base case for the root human anchor (Section~\ref{sec:chain-operator-definition}), specifies the immutability enforcement mechanism through the Fact Layer write protocol (Section~\ref{sec:fact-layer-write-protocol}), describes the Purpose Layer's spawn authorization role (Section~\ref{sec:purpose-layer-authorization}), provides the complete chain traversal algorithm with correctness arguments (Section~\ref{sec:chain-traversal}), and summarizes why immutability is an architectural constraint rather than a policy choice (Section~\ref{sec:immutability-summary}).

% -------------------------------------------------------
\subsection{Formal Definition: ParentPointer}
\label{sec:parent-pointer-formal}

\begin{definition}[Parent Pointer]
\label{def:parent-pointer}
Let $\mathcal{N}$ be the set of all agent nodes in a \bxthree{} system implementing the Recursive Spawning Protocol. For any node $c \in \mathcal{N}$ that was spawned via a valid RSP spawn event, the \emph{parent pointer} $\ParentPointer{p}{c}$ is the immutable, cryptographically signed reference from child $c$ to its immediate parent $p \in \mathcal{N}$, established exactly once at node provisioning via a Fact Layer atomic write operation.

The parent pointer relation satisfies the following structural constraints:

\begin{enumerate}
    \item[(P1) \textbf{Uniqueness}] For every node $c \in \mathcal{N}$ spawned under RSP, there exists exactly one node $p \in \mathcal{N}$ such that $\ParentPointer{p}{c}$ holds at all times after provisioning. No node may have multiple simultaneous parent pointers, and no parent pointer may be removed or redirected after establishment.

    \item[(P2) \textbf{Immutability}] After the provisioning write confirming $\ParentPointer{p}{c}$ is recorded in the Fact Layer ledger, no operation exists --- from any source, with any privilege level, under any system condition --- that can modify or delete $\ParentPointer{p}{c}$. The pointer is permanent for the lifetime of $c$. This is not a policy enforced by access control; it is a structural property of the protocol: the modification operation is not defined.

    \item[(P3) \textbf{Directionality}] The parent pointer is an intrinsic property of the child node, not a reference held by the parent node. The child $c$ carries $\ParentPointer{p}{c}$ as an immutable attribute recorded in its own Fact Layer state record. The parent $p$ holds no mutable reference to $c$. This means the chain is traversable from any node outward to its complete lineage regardless of the current operational state of any other node in the chain. A parent that has terminated, crashed, or been decommissioned does not affect the child's ability to reference its parent through $\alpha(n)$.

    \item[(P4) \textbf{Termination at Human Anchor}] The root of any RSP chain is a human principal node $h$ for which $\ParentPointer{\bot}{h}$ holds --- the null parent marker indicating that $h$ is the ultimate delegation origin and the chain terminates at a named human. No machine actor in the RSP model may serve as a chain root.
\end{enumerate}
\end{definition}

The parent pointer is not a conventional pointer variable subject to access control restrictions. It is a Fact Layer state property with no defined modification operation after provisioning. The distinction is architectural: an access-controlled pointer can be modified by any principal with sufficient privileges, and immutability is enforced by the access control policy until that policy is circumvented --- through credential theft, privilege escalation, or administrative compromise. The Fact Layer write protocol makes parent pointer immutability a structural property of the protocol itself; there is no modification pathway to circumvent.

\bigskip
\noindent\textbf{Notation.} We write $\ParentPointer{p}{c}$ to denote the parent pointer relation between parent $p$ and child $c$. We write $\alpha(c) = p$ as the equivalent functional notation. The notation $\alpha(c) = \bot$ denotes that node $c$ has no parent --- it is the root human anchor, the ultimate delegation origin. We use $p = \parent(c)$ and $c \in \children(p)$ interchangeably with the pointer relation.

% -------------------------------------------------------
\subsection{The Chain Operator}
\label{sec:chain-operator-definition}

The chain operator $\Chain{\cdot}$ is the primary tool for chain traversal, verification, and accountability attribution. It constructs the complete delegation lineage from any node to the human anchor by recursively applying the parent pointer function.

\begin{definition}[Chain Operator]
\label{def:chain-operator}
For any agent node $a \in \mathcal{N}$ in a Recursive Spawning chain, the \emph{chain} $\Chain{a}$ is defined recursively as:

\begin{equation}
\Chain{a} =
\begin{cases}
\{\,a\,\} \cup \Chain{\parent(a)} & \text{if } \alpha(a) \neq \bot \\[10pt]
\{\,a\,\hmannchor{a}\,\} & \text{if } \alpha(a) = \bot \quad \text{(root human anchor)}
\end{cases}
\label{eq:chain-recursive}
\end{equation}

Equivalently, expanding the recursion to reveal the full lineage:

\begin{equation}
\Chain{a} = \bigl\{\, a,\ \alpha(a),\ \alpha^{(2)}(a),\ \alpha^{(3)}(a),\ \ldots,\ \alpha^{(k)}(a) \,\bigr\}
\label{eq:chain-expanded}
\end{equation}
where $k$ is the depth of node $a$ in the chain (the number of successive parent applications required to reach the root), $\alpha^{(j)}(a)$ denotes the $j$-fold application of $\alpha$, and $\alpha^{(k)}(a)$ is the unique root node $n_0$ satisfying $\alpha(n_0) = \bot$ and $\hmannchor{n_0}$.
\end{definition}

The chain $\Chain{a}$ is the ordered set of all nodes on the delegation path from $a$ to and including the human anchor. It includes the terminal human anchor node, which is the only node in any RSP chain with $\alpha(n) = \bot$. The chain is always finite: by the RSP termination property, every RSP chain has depth at most $D_{\max}$, so the recursion in Equation~\ref{eq:chain-recursive} terminates at the root within at most $D_{\max} + 1$ successive applications of $\alpha$.

\bigskip
\noindent\textbf{Base case: root human anchor.} The root of every RSP chain is a Purpose Layer entity designated as a human principal, established at system initialization when a human principal $h$ authorizes the first agent node $n_0$. The Purpose Layer records $h(n_0) = h$ and $\alpha(n_0) = \bot$ in the Fact Layer authorization ledger. This is the only valid chain-root in RSP: no machine actor may serve as a chain origin, and no node may have $\alpha(n) = \bot$ unless it carries the $\hmannchor{\cdot}$ designation. The non-collapsing human anchor property ensures $h(n) = h(n_0)$ for all nodes in all chains spawned from $n_0$. If this property were violated, the chain would not terminate at a human principal and the RSP's accountability guarantee would be void.

\bigskip
\noindent\textbf{Properties of the chain operator.} The chain operator satisfies three critical properties that make it suitable for runtime verification:

\begin{itemize}
    \item[(C1) \textbf{Determinism}] For any node $a$, $\Chain{a}$ contains exactly one element for each ancestor on the delegation path. There is no alternative chain, no ambiguous parentage, and no non-deterministic branching. The parent pointer function $\alpha$ is a total function with unique values at every node, so the chain is uniquely determined.

    \item[(C2) \textbf{Stability}] The immutability of $\alpha$ guarantees that $\Chain{a}$ at any runtime point $t$ is identical to $\Chain{a}$ at the moment of $a$'s provisioning. The chain is not reconstructed from mutable logs or reconstructed retroactively from event records --- it is directly computable at any time by iterating $\alpha$ from the query node to the root. This stability is what makes chain verification a runtime property rather than a forensic reconstruction.

    \item[(C3) \textbf{Compositionality}] For any two nodes $a$ and $b$ in the same RSP chain, either $\Chain{a} \subseteq \Chain{b}$ or $\Chain{b} \subseteq \Chain{a}$. The chains are nested rather than intersecting. No cross-links or merging paths exist in a valid RSP chain, which follows directly from the uniqueness constraint (P1) on parent pointers.
\end{itemize}

% -------------------------------------------------------
\subsection{Immutability Enforcement: Fact Layer Write Protocol}
\label{sec:fact-layer-write-protocol}

The immutability of the parent pointer is enforced by the Fact Layer write protocol --- the \bxthree{} mechanism for making certain state transitions structurally irreversible. The write protocol is not access control with elevated restrictions; it is a protocol-level constraint on what operations are defined for $\alpha(n)$.

\bigskip
\noindent\textbf{Write protocol definition.} The Fact Layer write protocol defines the valid write operations on each node property at each node state transition. For the parent pointer $\alpha(n)$, the protocol defines exactly one valid write: the initial provisioning write that sets $\alpha(n) = p$ at node creation. The protocol defines \emph{no valid write operations} for modifying $\alpha(n)$ after provisioning. This is a structural constraint --- the modification operation does not exist in the protocol's defined operation set. Any attempt to introduce such an operation is rejected at the protocol boundary.

Any operation --- from any source, with any justification --- that attempts to overwrite, redirect, or delete $\alpha(n)$ after provisioning is rejected at the protocol level. The Fact Layer does not process the request, does not invoke an override pathway, and does not log the attempt as a rejected operation with retry options. The request is malformed with respect to the write protocol and is discarded before reaching any evaluation stage.

\bigskip
\noindent\textbf{Four properties of Fact Layer immutability enforcement.}

\begin{enumerate}
    \item[(I1) \textbf{Failure-state resilience}] If the system is in a degraded operational state --- partial network partition, Bounds Engine timeout, Purpose Layer unavailability --- the Fact Layer write protocol for $\alpha(n)$ immutability remains in force. The modification request is rejected regardless of system state, because the write protocol is a local constraint at the Fact Layer, not a distributed agreement protocol that could be blocked by network failure or component timeout. This property distinguishes the write protocol from conventional access control systems, which typically depend on availability of the authorization service.

    \item[(I2) \textbf{Administrative override absence}] There is no administrative role, system operator, or security principal with the capability to modify $\alpha(n)$. This is a deliberate architectural omission. In conventional systems, administrative compromise leads directly to immutability violation because the administrative role has override authority. In RSP, there is no override authority for the parent pointer by design. The absence is not an oversight that could be corrected with a future role --- it is structurally necessary for the correctness of the chain integrity property.

    \item[(I3) \textbf{Source-equivalence of rejection}] The Fact Layer write protocol treats all modification requests identically regardless of source. A request from the Purpose Layer, the Bounds Engine, an authenticated administrator, a system security officer, or an external adversarial actor receives the same response: protocol-level rejection. There is no privileged source that bypasses this check. This is distinct from capability-based security where held capabilities determine permissibility; here the protocol itself has no mechanism to evaluate the request.

    \item[(I4) \textbf{Atomic warrant-pointer creation}] The parent pointer and its Purpose Layer warrant are created as a single atomic Fact Layer write. There is no temporal window in which $\alpha(n)$ is established without a matching warrant, or in which a warrant exists without a corresponding parent pointer. The provisioning atomicity ensures the chain topology is always consistent with its authorization record. This property eliminates the class of exploit where a parent pointer is first established and then retroactively authorized.
\end{enumerate}

\bigskip
\noindent\textbf{Why access control is insufficient.} The distinction between access-control-based immutability and protocol-level immutability is the distinction between a promise and a guarantee. In an access-control model, immutability is enforced by restricting which principals may issue modification operations. If an attacker acquires sufficient privileges --- through credential theft, insider compromise, or software vulnerability --- the access control barrier is circumvented and immutability is violated. The immutability was a policy, enforceable until it was not.

In the Fact Layer write protocol, the immutability of $\alpha(n)$ is a property of the protocol itself. There is no operation defined for modifying $\alpha(n)$ after provisioning, so no amount of privilege elevation, credential compromise, or software vulnerability can produce a valid modification operation. The protocol does not enforce immutability --- it makes immutability the only valid state transition. The distinction is architectural, not a matter of degree.

This is the precise failure mode that the Fact Layer write protocol eliminates: retroactive manipulation of chain topology for the purpose of accountability laundering. Without protocol-level immutability, an adversarial actor could re-parent a node to a favorable parent after the fact, obscuring the original authorization chain and making drift detection impossible. With protocol-level immutability, this manipulation is structurally impossible.

The same logic applies to Scope modification after provisioning. While scope refinement is enforced at each spawn event through the Purpose Layer warrant check, the parent pointer immutability ensures the structural backbone of the chain cannot be corrupted after the fact. Even if a node's operating scope were somehow compromised post-provisioning --- a condition the protocol does not allow but which would be detectable through Chain-Verify --- the chain topology itself remains intact and auditable.

% -------------------------------------------------------
\subsection{Spawn Authorization: Purpose Layer Role}
\label{sec:purpose-layer-authorization}

The Purpose Layer plays two structurally necessary roles in the parent pointer lifecycle: it originates the spawn authorization policy that defines the conditions under which a new parent pointer may be created, and it issues the spawn warrant that is the prerequisite for the Fact Layer provisioning write. These two roles are inseparable from the parent pointer's immutability: if the Purpose Layer could issue warrants after the fact, or if a machine actor could create a parent pointer without a Purpose Layer warrant, the chain's integrity guarantees would be void.

\bigskip
\noindent\textbf{Authorization policy origination.} At system initialization, the Purpose Layer defines the spawn authorization policy $P_{\text{spawn}}$ and records it in the Fact Layer authorization ledger. $P_{\text{spawn}}$ specifies the mandatory conditions for any valid spawn event:

\begin{enumerate}
    \item[(S1) \textbf{Scope refinement}] The child scope must be a strict subset of the parent scope: $R(n_{\text{child}}) \subset R(n_{\text{parent}})$. No lateral expansion, no superset, no equality. This ensures that each spawn event narrows the delegation scope, preserving the intent-refinement property of the chain.

    \item[(S2) \textbf{Operational necessity}] The parent must declare, in the Fact Layer authorization ledger, the specific operational condition that necessitates spawning a child and cannot be resolved within the parent's scope. This condition must be a concrete, verifiable state (such as a sensor reading outside $R(n_{\text{parent}})$ or an exception condition) rather than a vague or future-oriented justification.

    \item[(S3) \textbf{Minimum scope proportionality}] The child scope must be the minimum scope necessary to address the declared operational condition. Broader scopes are not authorized. This prevents scope inflation at spawn events, where a child is provisioned with a scope larger than necessary to address the declared condition.

    \item[(S4) \textbf{Depth bound compliance}] The proposed spawn must satisfy $\depth(n_{\text{parent}}) + 1 \leq D_{\max}$. If this condition fails, the spawn is rejected and mandatory escalation is triggered to the human anchor. This ensures the termination property is preserved at every spawn event, rather than enforced only after the fact.
\end{enumerate}

These conditions are non-negotiable. The Purpose Layer does not issue a spawn warrant if any condition is unmet. The Fact Layer pre-commit hook independently re-verifies these conditions before recording the provisioning write. The pre-commit hook operates without relying on the Purpose Layer's own output --- it re-evaluates the conditions directly against the Fact Layer ledger.

\bigskip
\noindent\textbf{Spawn warrant issuance.} When all four conditions are satisfied, the Purpose Layer issues a spawn warrant $w$ --- a cryptographically signed, Fact Layer-recorded authorization record:

\begin{equation}
w = \bigl\langle \text{parent}=n_i,\ \text{child}=n_{i+1},\ R(n_{i+1}),\ \text{justification},\ \text{timestamp},\ \sigma_{\text{Purpose Layer}} \bigr\rangle
\label{eq:spawn-warrant}
\end{equation}

where $\sigma_{\text{Purpose Layer}}$ is the cryptographic signature of the Purpose Layer over the warrant contents, providing non-repudiation and integrity verification for the authorization record.

The warrant is written to the Fact Layer authorization ledger atomically with the provisioning write that establishes $\alpha(n_{i+1}) = n_i$. The atomicity of this write is critical: the parent pointer and the spawn warrant are created as a single indivisible ledger entry. There is no temporal window in which $\alpha(n_{i+1})$ is established without a matching warrant, or in which a warrant exists without a corresponding parent pointer. The Fact Layer ledger entry encompasses both the pointer state and the authorization state simultaneously.

\bigskip
\noindent\textbf{The Purpose Layer is the sole authorization origin.} No machine actor may establish a parent pointer without a Purpose Layer warrant. This is not an advisory constraint --- it is a structural consequence of the Fact Layer write protocol: the pre-commit hook verifies the existence of a valid Purpose Layer warrant before recording any provisioning write. A request to establish $\alpha(n)$ without a matching warrant is rejected at the pre-commit hook, before the operation reaches the ledger. The Purpose Layer's exclusive role as authorization origin prevents the chain from being bootstrapped by a machine actor acting autonomously. Even if every other RSP mechanism failed simultaneously, the sole-authorization origin property would prevent a machine actor from independently constructing a delegation chain.

% -------------------------------------------------------
\subsection{Chain Traversal Algorithm}
\label{sec:chain-traversal}

The chain traversal algorithm is the runtime procedure for constructing the delegation chain from any node to its human anchor. It is provided here in full algorithmic detail with correctness arguments establishing that it terminates, returns the correct chain, and preserves chain integrity during traversal.

\bigskip
\noindent\textbf{Algorithm: Chain-Traverse}
\begin{algorithm}
\caption{Chain-Traverse($n$) --- Build the delegation chain from node $n$ to the human anchor}
\label{alg:chain-traverse}
\begin{algorithmic}[1]
\REQUIRE $n$ is a provisioned RSP node with $\alpha$ defined
\ENSURE $\Chain{n}$ --- the ordered set of all ancestor nodes including the human anchor
\STATE $chain \leftarrow \langle \ \rangle$ \COMMENT{Initialize empty ordered list}
\STATE $current \leftarrow n$ \COMMENT{Start at the query node}

\WHILE{$\text{true}$}
    \STATE $\text{append}(chain,\ current)$ \COMMENT{Add current node to chain}
    \IF{$\alpha(\text{current}) = \bot$}
        \STATE \textbf{break} \COMMENT{Root human anchor reached --- chain complete}
    \ENDIF
    \STATE $\text{current} \leftarrow \alpha(\text{current})$ \COMMENT{Move to parent}
\ENDWHILE

\RETURN $chain$
\end{algorithmic}
\end{algorithm}

\bigskip
\noindent\textbf{Correctness arguments.}

\textit{Termination.} The algorithm terminates because RSP chains are depth-bounded by construction. Each iteration of the while-loop moves $\text{current}$ to the parent node $\alpha(\text{current})$, which is strictly closer to the root along the chain. Since the chain has finite depth at most $D_{\max} + 1$, the loop executes at most $D_{\max} + 1$ iterations before $\alpha(\text{current}) = \bot$ is reached. The loop therefore terminates in finite time. This argument holds regardless of the current operational state of any node in the chain, including nodes that may have terminated or entered error states, because the traversal only reads the immutable $\alpha$ values.

\textit{Correctness of the returned set.} By Definition~\ref{def:chain-operator}, $\Chain{a}$ is the set containing $a$, $\alpha(a)$, $\alpha(\alpha(a))$, and so on, until the root. The algorithm produces exactly this sequence: it starts at $a$, appends each node, moves to its parent via $\alpha$, and stops when $\alpha(\text{current}) = \bot$ is reached. The output is identical to $\Chain{a}$ by construction.

\textit{Uniqueness of result.} The algorithm produces a unique result for any input node because $\alpha$ is a total function (defined for every provisioned node) and is injective on the chain direction (every node has exactly one parent, and the root has no parent). There is no branching, no alternative path, and no non-deterministic choice at any step in the algorithm.

\textit{Immutability guarantee.} The algorithm reads $\alpha$ values but never writes them. The traversal is read-only and does not interact with the Fact Layer write protocol. The algorithm can be executed at any runtime point without affecting the immutability of the chain it traverses, and without creating any modification opportunities for adversarial actors.

\bigskip
\noindent\textbf{Algorithm: Chain-Verify}
The chain traversal algorithm enables Chain-Verify, which confirms the structural validity of a delegation chain at any runtime point. This algorithm is the operational counterpart to the chain operator definition: it provides the runtime verification procedure that confirms a given node's chain satisfies all RSP structural constraints.

\begin{algorithm}
\caption{Chain-Verify($n$) --- Verify the authorization chain from node $n$ to human anchor}
\label{alg:chain-verify}
\begin{algorithmic}[1]
\REQUIRE $n$ is a provisioned RSP node
\ENSURE $\text{Boolean}$ indicating whether the chain is a structurally valid RSP chain

\STATE $chain \leftarrow \text{Chain-Traverse}(n)$
\STATE $m \leftarrow \text{len}(chain)$
\STATE \COMMENT{\textit{Step V1: Verify chain terminates at human anchor}}
\IF{$\alpha(chain[m-1]) \neq \bot$}
    \RETURN $\text{false}$ \COMMENT{Root is not $\bot$ --- malformed chain}
\ENDIF
\IF{$\neg\hmannchor{chain[m-1]}$}
    \RETURN $\text{false}$ \COMMENT{Root is not a designated human principal}
\ENDIF

\STATE \COMMENT{\textit{Step V2: Verify each spawn event has a Purpose Layer warrant}}
\FOR{$i \leftarrow 0$ \TO $m-2$}
    \STATE $child \leftarrow chain[i]$
    \STATE $parent \leftarrow chain[i+1]$
    \IF{$\neg\text{warrant\_exists}(child,\ parent)$}
        \RETURN $\text{false}$ \COMMENT{Missing Purpose Layer warrant for spawn}
    \ENDIF
\ENDFOR

\STATE \COMMENT{\textit{Step V3: Verify scope refinement at each link}}
\FOR{$i \leftarrow 0$ \TO $m-2$}
    \STATE $child \leftarrow chain[i]$
    \STATE $parent \leftarrow chain[i+1]$
    \IF{$\neg(R(child) \subset R(parent))$}
        \RETURN $\text{false}$ \COMMENT{Scope refinement constraint violated}
    \ENDIF
\ENDFOR

\STATE \COMMENT{\textit{Step V4: Verify depth bound compliance at each node}}
\FOR{$i \leftarrow 0$ \TO $m-1$}
    \IF{$\text{depth}(chain[i]) > D_{\max}$}
        \RETURN $\text{false}$ \COMMENT{Depth bound exceeded at node}
    \ENDIF
\ENDFOR

\RETURN $\text{true}$
\end{algorithmic}
\end{algorithm}

Chain-Verify returns \texttt{true} if and only if all four verification conditions hold simultaneously. The verification inspects structural chain properties rather than current operational state. Each condition corresponds to a fundamental RSP constraint: V1 confirms the chain terminates at a human anchor (not at a machine actor or orphaned node), V2 confirms every spawn was authorized by the Purpose Layer (not self-bootstrapped by a machine actor), V3 confirms scope refinement was respected at every link (not lateral expansion), and V4 confirms the depth bound was respected throughout (not approaching infinity). A chain that passes Chain-Verify is, by the RSP definitions, a fully authorized delegation chain terminating at a human principal.

The algorithm is deterministic and produces the same result regardless of which node in the chain initiates verification. There is no self-serving interpretation of authorization that a drifted or orphaned node could produce, because Chain-Verify inspects the complete chain topology and warrants in the Fact Layer ledger, not the node's own claims about its authorization. Chain verification is a property of the chain topology, not a property of the node being verified.

\bigskip
\noindent\textbf{Computational cost.} Chain-Traverse runs in $O(d)$ time where $d$ is the chain depth (at most $D_{\max} + 1$). Chain-Verify runs in $O(m)$ time for the traversal plus $O(m)$ for each of the three verification loops, yielding $O(m)$ total where $m$ is chain length (also bounded by $D_{\max} + 1$). The algorithms are linear in chain length and do not depend on the total number of nodes in the system.

% -------------------------------------------------------
\subsection{Summary: Immutability as Architectural Constraint}
\label{sec:immutability-summary}

The immutable parent pointer is the load-bearing constraint of the Recursive Spawning Protocol. Without it, the chain is mutable --- subject to retroactive modification by actors with sufficient privilege --- and the RSP's correctness properties (drift prevention, chain integrity, termination) collapse to policy conventions rather than structural guarantees.

The immutability is architectural --- not a stronger access control policy --- because it is enforced by the Fact Layer write protocol, which defines no valid modification operation for $\alpha(n)$ after provisioning. This is distinct from access-control-based immutability in four ways: there is no override path for any actor; the immutability holds under all system failure conditions including those that disable other enforcement mechanisms; all sources receive equivalent enforcement treatment regardless of privilege level; and the parent pointer and its Purpose Layer warrant are created atomically with no temporal window for inconsistency.

The chain operator $\Chain{a}$ is uniquely determined by the immutable parent pointers in the chain. The chain traversal algorithm computes this operator correctly and terminates. Chain-Verify provides the runtime verification procedure for chain structural validity. Together, these mechanisms make the delegation chain an immutable, verifiable, and auditable runtime artifact --- the foundation on which all other RSP correctness properties rest.
