\section{Background}
\label{sec:background}

The problem of accountability in multi-agent AI systems has become urgent as agents gain the ability to spawn sub-agents, delegate tasks recursively, and operate at depths that render human oversight structurally difficult. The gap between operational capability and accountability infrastructure has widened into a chasm. This section situates the Recursive Spawning Protocol within five intersecting research threads: agent lifecycle management, accountability and provenance in distributed systems, the structure of delegation chains, hierarchical task decomposition, and the BX3 Framework as a substrate for immutable parent pointers.

% ------------------------------------------------------------------------- %
\subsection{Agent Lifecycle Management in Multi-Agent Systems}
\label{sec:lifecycle-management}

Modern multi-agent frameworks treat agent creation and termination as first-class operational events. Google DeepMind's Tiered Agentic Oversight (TAO) defines explicit lifecycle boundaries: an agent is provisioned with a defined scope, executes within that scope, and is decommissioned when its task completes or its authority is revoked \cite{tao2025}. In TAO, delegation is a privileged operation --- a parent agent must hold explicit warrant authority to create a child. This is a meaningful improvement over unconstrained spawning, but TAO's escalation path still routes through intermediate machine actors: when an agent fails, it escalates to a superior AI tier before potentially reaching a human. The chain of machine intermediaries creates latency, introduces single points of failure, and does not eliminate the possibility that an intermediate agent modifies its own delegation state retroactively.

More recent work has examined lifecycle management through the lens of organizational hierarchy. Enterprise agentic AI architecture guidance identifies agent communication and coordination as a core operational layer requiring structured protocols and explicit coordination patterns \cite{enterprise2025}. Without explicit lifecycle boundaries, agents may spawn children that outlive their parent's operational context, creating orphaned execution contexts that continue indefinitely with no accountable principal. The agent management platform literature confirms that governance infrastructure --- not just technical capability --- determines whether agent lifecycles are manageable at scale \cite{kore2025}. Only 15\% of IT leaders report having adequate governance models for AI agents, and organizations without unified governance are over three times less likely to report high value from AI investments \cite{kore2025}.

The lifecycle management problem is acute in recursive spawning scenarios. When an agent can spawn a child that itself spawns grandchildren, the total number of active agents can grow exponentially within a single operational session. Without explicit lifecycle boundaries and deterministic termination conditions, the agent graph becomes difficult to audit, impossible to verify completely, and susceptible to drift --- a child agent continuing to execute after its parent has been decommissioned, with no operational connection to any living principal.

The Human Delegation Provenance (HDP) protocol establishes that lifecycle integrity requires not only provisioning and termination primitives but also cryptographically signed records of every delegation hop across the agent lifecycle \cite{hdp2026,helixar2026}. HDP introduces an append-only provenance chain in which each agent's delegation action is recorded as a signed hop, enabling offline verification of the full chain using only the issuer's public key and session identifier. The protocol addresses the ``delegation gap'' in multi-hop agentic workflows where human intent and authorized scope are not verifiably preserved as actions cascade through autonomous agents \cite{arxiv2604.04522}. This cryptographic framing of lifecycle management is necessary but not sufficient: a signed delegation chain is only as reliable as the parent pointer it traces. If the parent pointer is mutable, the provenance chain can be retrospectively corrupted after the delegation occurs.

% ------------------------------------------------------------------------- %
\subsection{Accountability and Provenance in Distributed Systems}
\label{sec:accountability-provenance}

The provenance problem in distributed systems predates AI by decades. Database systems, distributed ledgers, and cloud infrastructure have all confronted the challenge of maintaining a complete, tamper-evident record of which actor performed which action at what time. Provenance tracking in distributed systems is typically implemented as an audit log: append-only, timestamped records that establish a historical chain of custody for data or actions. However, audit logs are only as reliable as the system that writes them. If the writing process itself can be manipulated --- by privilege escalation, insider compromise, or software vulnerability --- the audit log becomes a record of the attack rather than a record of the original action.

The specific failure mode most relevant to multi-agent AI systems is what Hood \cite{hood2025} terms \emph{authority laundering}: when delegation chains become sufficiently long and fast, the origin of authority becomes obscured. An agent at depth $n$ takes an action that appears to originate from depth $n-1$, which appears to originate from depth $n-2$, and so on, eventually reaching what is claimed to be an authorized human principal. But if any link in this chain is mutable --- if a parent pointer can be redirected after the fact --- then the chain of authority is retrospectively corruptible. The action taken at depth $n$ could be re-parented to a different ancestor after the fact, creating a plausible but false narrative of authorized delegation.

The provenance paradox identified in multi-agent LLM routing is directly relevant: current protocols do not govern delegation under unverifiable quality claims \cite{openexecution2026}. Without an explicit model of what each delegation hop is authorized to do, and without a tamper-evident record of the chain at the time the action was taken, retrospective audit is structurally unreliable.

The FCA Mills Review, launched January 27, 2026, explicitly flags the need for stronger traceability and assurance in agentic AI systems operating in retail financial environments \cite{ssrnpolicyprovenance2026}. The EU AI Act requires that high-risk AI systems maintain technical documentation including audit logs that enable third-party conformity assessment \cite{euaiact2024}. NIST's AI Risk Management Framework specifies that AI systems should maintain ``records of data lineage, model development, and decision-making processes'' to support accountability \cite{nistairmf2023}. ISO/IEC 42001:2023, the international standard for AI management systems, requires documented procedures for the operation of AI systems including provisions for monitoring and controlling changes to system behavior \cite{iso420012023}. These regulatory requirements create compliance pressure for systems that can demonstrate accountability at depth --- but the technical means to achieve this in recursive spawning environments remain underdeveloped.

Recent architectural work has sought to close this gap. SentinelAgent introduces a Delegation Authority Service (DAS) with eight components including an Identity Registry, a Policy Engine evaluating NIST 800-53 constraints, an Intent Verifier performing three-layer intent checking, a Token Signer issuing HMAC-signed delegation tokens, and a Chain Store maintaining complete chain history \cite{sentinelagent2026}. The Delegation Chain Calculus defined in SentinelAgent specifies seven properties, six deterministic and one probabilistic: authority never escalates through delegation (P1), intent preservation is verified at every step (P2), NIST 800-53 compliance is enforced throughout the chain (P3), every action is forensically traceable to the human who authorized it (P4), cascade damage is formally bounded (P5), only permitted API calls execute (P6), and only conformant outputs are returned (P7) \cite{sentinelagent2026}. These properties provide a formal specification for what accountability in multi-agent systems must satisfy.

The critical insight for the Recursive Spawning Protocol is that all seven properties of the Delegation Chain Calculus depend on the chain being immutable. An immutable parent pointer means that at the moment any action is taken, the accountable ancestor is structurally fixed and cannot be altered retroactively. This is the property that transforms the delegation chain from a policy description into a forensically reliable record.

% ------------------------------------------------------------------------- %
\subsection{Fixed vs. Mutable Delegation Chains}
\label{sec:fixed-vs-mutable}

The distinction between fixed and mutable delegation chains is the central structural decision point for accountability in multi-agent systems. A mutable delegation chain permits the parent pointer of an active agent to be redirected after provisioning --- either by the parent, by an administrator, or by some automated rebalancing process. A fixed delegation chain prohibits this redirection entirely: the parent pointer established at provisioning is the only parent pointer the agent ever has.

Mutable chains offer operational flexibility: workloads can be re-parented for load balancing, organizational restructuring, or fault recovery. In traditional computing environments where administrators have privileged access and where operations are reversible, this flexibility is often desirable. In AI agent systems, the flexibility comes at a catastrophic accountability cost. When auditing an action taken by an agent at depth $n$, the audit must reconstruct which parent was actually responsible at the time of the action. If the parent pointer has been modified since the action occurred, the reconstruction is fundamentally unreliable. The audit trail may show a current parent, but the current parent is not necessarily the parent that authorized the action. This is not a policy failure --- it is a structural failure of the chain itself. No amount of access control, auditing, or administrative oversight closes this gap if the parent pointer is mutable.

Fixed delegation chains impose a one-time write constraint: the parent pointer is established at provisioning and may not be modified thereafter. This is not a stronger access control policy --- it is a different architectural category. An access control policy restricts which principals may issue modification operations; if sufficient privilege is acquired, the modification succeeds. A fixed chain has no modification operation defined at the protocol level --- the constraint is not enforced, it is architecturally impossible. This distinction is critical for the Recursive Spawning Protocol's correctness guarantees.

The case for fixed chains in AI systems is reinforced by the non-determinism problem. Because AI inference is probabilistic, the same input may produce different outputs on different invocations. A mutable delegation chain combined with probabilistic behavior means that even if the delegation record is accurate at time $t$, it may not reflect the actual execution context at time $t+1$. An agent that was a reliable delegate at provisioning may have drifted --- through model updates, context corruption, or adversarial manipulation --- into an unreliable actor by the time it takes its next action. If the chain is mutable, this drift is invisible. If the chain is fixed, the drift is at least attributable: the original parent remains on record as responsible regardless of the child's current state.

The operational cost of fixed chains is reduced flexibility. Re-parenting is a useful operation in traditional computing --- it allows workload migration, load balancing, and organizational restructuring. In AI agent systems, these operations are secondary to the requirement that every action be attributable to a human principal at the root of the chain. The Recursive Spawning Protocol trades re-parenting flexibility for structural accountability, enforcing immutability at the protocol level rather than relying on policy to constrain behavior.

The Agentic Control Plane specification formalizes the invariants of delegation chains: scopes only narrow, budgets only decrease, cycles are forbidden, and every hop is audited \cite{agenticcontrolplane2026}. It recommends plan capping of 4--8 hops to prevent runaway delegation topology. These invariants are necessary but implemented as policy constraints rather than structural properties. The Recursive Spawning Protocol complements these policy constraints with a structural mechanism --- the immutable parent pointer --- that enforces the same invariants at the protocol level, making them immune to privilege escalation or software vulnerability.

% ------------------------------------------------------------------------- %
\subsection{Hierarchical Task Decomposition in AI}
\label{sec:htn}

Hierarchical task decomposition provides the organizational model for recursive delegation in AI systems. The dominant formal framework is Hierarchical Task Network (HTN) planning, in which a complex task is recursively decomposed into primitive, executable subtasks through the application of domain-specific methods \cite{xu2025htn,emergetmindhtn2025}. HTN planning semantics strictly distinguish between the domain control knowledge encoded in hierarchical methods and the search process, ensuring that solution plans closely follow human-specified task reduction protocols \cite{xu2025htn}. Each HTN method encodes admissible sub-plans for reducing domain-specific compound tasks, with method selection governed by precondition satisfaction in the current planning state \cite{emergetmindhtn2025}.

The integration of LLMs with HTN planning has produced a rapidly evolving body of work. ReCAP (Recursive Context-Aware Planning) enables LLMs to perform long-horizon reasoning through task decomposition, execution, and backtracking, combining plan-ahead decomposition with structured parent-plan re-injection and memory-efficient execution that scales linearly with task depth \cite{recap2025}. Xu et al. introduce a verifier-method learner that inserts an explicit verifier task after each LLM decomposition to ensure the decomposition achieves the desired effect, and generalizes observed primitive decompositions into reusable methods by lifting constants to variables \cite{xu2025htn}. The General Framework for HTN Modeling with LLMs establishes that real-world problems exhibit inherent hierarchical structure that mirrors human problem-solving strategies, which are typically more hierarchical than sequential \cite{icaps25htn2025}.

The connection between HTN and recursive spawning is structural: a compound task decomposed by a parent agent into subtasks, each subtask potentially spawning a child agent that applies further decomposition. The hierarchical task network thus maps directly onto the delegation graph of a multi-agent system, with HTN methods serving as the delegation protocols and primitive tasks serving as the termination condition at which no further spawning is permitted.

However, existing HTN formalisms treat the parent-child relationship as a planning construct --- a transient relationship established for the duration of a plan and dissolved when the plan completes. In AI agent systems, the parent-child relationship is not transient; it is a persistent accountability relationship that outlasts any individual plan or task. The BX3 Framework's worksheet architecture makes this distinction explicit: each child loop operates within a containerized worksheet that preserves local state and local goal structures while remaining subordinate to the parent's accountability \cite{bx3whitepaper2026}. The parent pointer is not a planning convenience; it is a permanent accountability anchor.

The GSA's AI Agent Specification Template formalizes this distinction by defining HTN decomposition as the core methodology for agentic task planning, specifying that agents recursively break down abstract goals into a dependency graph of primitive, executable tasks \cite{gsadevcrew2025}. The template is explicit that the decomposition graph is a durable artifact --- not a transient plan that dissolves after execution --- because each node in the graph corresponds to an accountable agent with a persistent chain of authority.

% ------------------------------------------------------------------------- %
\subsection{The BX3 Framework as Substrate for Immutable Parent Chains}
\label{sec:bx3-substrate}

The BX3 Framework provides the architectural substrate on which the Recursive Spawning Protocol is built. The framework defines three immutable functional layers --- Purpose Layer, Bounds Engine, and Fact Layer --- with enforced properties that hold regardless of which actor occupies each layer \cite{bx3whitepaper2026}. The actor-agnostic, function-centered definition means that any actor satisfying a layer's functional requirements may occupy that layer, including humans, AI systems, and mechanical processes.

The Purpose Layer provides intent, judgment, and accountability. It is the layer at which goals are set, trade-offs are weighed, and final accountability is assigned. The Purpose Layer does not execute actions; it defines what ought to happen and is responsible for what happens. Any actor occupying the Purpose Layer must be capable of reasoning about goals, consequences, and accountability \cite{bx3whitepaper2026}.

The Bounds Engine provides interpretation, bounded reasoning, and constrained execution. It is the layer that translates the Purpose Layer's intent into executable plans within hard constraints. The Bounds Engine may not execute any action that has not been validated by the Fact Layer, and it may not modify its own boundary conditions \cite{bx3whitepaper2026}.

The Fact Layer provides deterministic enforcement, hard physical constraint, and forensic auditability. It is the only layer that can produce non-reversible physical effects, and its defining characteristic is 100\% deterministic behavior: given the same inputs, it must produce identical outputs \cite{bx3whitepaper2026}. The Fact Layer maintains a forensic ledger of all state changes, providing a complete and tamper-evident record of all physical actions. The Fact Layer enforces constraints that are not merely business rules or software permissions but physical limits --- circuit breakers, mechanical interlocks, chemical process controls, and access restrictions that exist independent of any computational system.

The immutability of the parent pointer is enforced by the Fact Layer write protocol: there is no defined modification operation for $\alpha(n)$ after provisioning, making immutability the only valid state transition \cite{bx3whitepaper2026}. This is architecturally distinct from access-control-based immutability. The distinction is between a promise and a guarantee: access control enforces immutability by restricting which principals may issue modification operations; if sufficient privilege is acquired, the barrier is circumvented. The Fact Layer makes the modification operation structurally impossible --- there is no privileged override, no emergency bypass, and no administrative role with the capability to alter $\alpha(n)$ after provisioning.

The BX3 Framework's recursive spawning and worksheet architecture enables systems to maintain goal coherence across arbitrarily long chains of delegated reasoning and action without requiring continuous cloud connectivity or centralized control \cite{bx3whitepaper2026}. A parent node births a child loop that carries a hard-coded pointer to the parent's Purpose Layer, and each child loop operates within a containerized worksheet that preserves local state while remaining subordinate to the parent's accountability. This is the operationalization of immutable parent chains at the framework level: the parent pointer is established at provisioning, stored in the child's sealed memory envelope, and is never mutable after the atomic Fact Layer write confirms the spawn event.

The Recursive Spawning Protocol extends the BX3 Framework by specifying the exact mechanism by which parent pointers are established, traversed, verified, and maintained across the delegation chain. The protocol draws on the framework's three-layer architecture to separate authorization (Purpose Layer) from assignment (Fact Layer), ensuring that a parent cannot claim to have spawned a node whose pointer was set to a different actor, and that a child cannot falsify its own ancestry after provisioning. The result is a delegation chain that is not only accountable by policy but structurally accountable by design.

% ------------------------------------------------------------------------- %
\subsection{Summary}
\label{sec:background-summary}

The five threads reviewed here converge on a structural gap in existing multi-agent AI architectures. Agent lifecycle management provides provisioning and termination primitives, but does not enforce immutable parentage. Accountability and provenance systems provide audit infrastructure, but cannot close the gap created by mutable delegation chains. Hierarchical task decomposition provides the organizational model for recursive delegation, but treats the parent-child relationship as mutable configuration rather than as an immutable structural property. Fixed vs. mutable delegation chains represent the critical design choice: mutable chains offer operational flexibility at the cost of retrospective corruptibility, while fixed chains sacrifice re-parenting capability in exchange for cryptographic accountability. The BX3 Framework provides the architectural substrate --- specifically the Fact Layer's deterministic enforcement and the immutable three-layer separation --- on which immutable parent chains become structurally possible rather than merely policy-desirable.

The Recursive Spawning Protocol builds on this convergence by specifying the exact mechanism: an immutable parent pointer established at provisioning by an atomic Fact Layer write, traversable in constant time via the chain operator, and verifiable at runtime via Chain-Verify. The protocol eliminates the accountability gap that mutable delegation chains create, ensuring that every spawned node maintains a fixed, immutable, cryptographically signed connection to its authorizing Purpose Layer actor, terminating at a human.

% ------------------------------------------------------------------------- %
\begin{thebibliography}{99}
\bibliographystyle{plain}

\itemsep0em

\item Agentic Control Plane. \emph{What is an Agent Delegation Chain?}, 2026. \url{https://agenticcontrolplane.com/what-is-an-agent-delegation-chain}.

\item Agentic Control Plane. \emph{Agent-to-Agent Governance --- Delegation Chains, Trust Boundaries, and Audit}, 2026. \url{https://agenticcontrolplane.com/agent-to-agent}.

\item arxiv. \emph{HDP: A Lightweight Cryptographic Protocol for Human Delegation Provenance in Agentic AI Systems}, arXiv:2604.04522, April 2026. \url{https://arxiv.org/abs/2604.04522}.

\item Auditable Agents. \emph{Auditable Agents}, arXiv:2604.05485, April 2026. \url{https://arxiv.org/abs/2604.05485}.

\item BX3 Framework White Paper v2. \emph{The BX3 Framework: A Universal Architecture of Functional Roles for Purpose, Bounded Reasoning, and Deterministic Fact}, April 2026.

\item Colorado Senate Bill 24-205. \emph{Colorado AI Act}, 2024.

\item Emergent Mind. \emph{Hierarchical Task Network Planning}, 2025. \url{https://www.emergentmind.com/topics/hierarchical-task-network-htn-planning-framework}.

\item EU AI Act. \emph{Regulation (EU) 2024/1689}, European Parliament and Council, 2024.

\item Helixar. \emph{HDP: The Open Protocol That Gives AI Agents a Verifiable Chain of Authority}, 2026. \url{https://helixar.ai/press/hdp-human-delegation-provenance-protocol/}.

\item IETF. \emph{Human Delegation Provenance Protocol (HDP): Cryptographic Chain-of-Custody for Agentic AI Systems}, draft-helixar-hdp-agentic-delegation-00, March 2026. \url{https://dt.ietf.org/doc/draft-helixar-hdp-agentic-delegation/}.

\item ISO/IEC 42001:2023. \emph{Information Technology --- Artificial Intelligence --- Management System}, 2023.

\item Kore.ai. \emph{AI Agent Management Platform Research}, 2025.

\item NIST. \emph{Artificial Intelligence Risk Management Framework (AI RMF 1.0)}, NIST AI 100-1, 2023.

\item OpenExecution. \emph{OpenExecution Provenance Specification}, 2026. \url{https://github.com/Open-Execution/openexecution-provenance-spec}.

\item Xu et al. \emph{Online Learning of HTN Methods for Integrated LLM-HTN Planning}, November 2025.

\item arXiv:2505.11814. \emph{ChatHTN: Integrating Large Language Models with Hierarchical Task Network Planning}, May 2025. \url{https://arxiv.org/pdf/2505.11814}.

\item Zylos Research. \emph{AI Agent Accountability: Audit Trails, Attribution, and Non-Repudiation in Multi-Agent Systems}, March 2026.

\end{thebibliography}